---
title: "Annotation of factors from scRNA analysis"
author: "Ziang Zhang"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Process the result {#Data_Processing}

Use this function for plotting:

```{r}
source("https://raw.githubusercontent.com/stephenslab/single-cell-jamboree/refs/heads/main/code/group_factors.R")
library(ggplot2)
library(fastTopics)
```

Read in the meta-information and the factorization result:

```{r}
data_path <- "../data/"
code_path <- "../code/"
# data_path <- "data/"
# code_path <- "code/"
source(paste0(code_path, "ROC.R"))
seurat_meta <- readRDS(paste0(data_path, "seurat_meta.rds"))
flashier_snmf_summary <- readRDS(paste0(data_path, "flashier_snmf_summary.rds"))
```

Align the cells to remove the extra cells in the flashier result:

```{r}
cells_flashier <- rownames(flashier_snmf_summary$L_pm)
cells_seurat <- seurat_meta$cellID
which(!cells_flashier %in% cells_seurat)
L_pm <- flashier_snmf_summary$L_pm[cells_flashier %in% cells_seurat, ]
all(rownames(L_pm) == cells_seurat)
```

Let's consider the normalization approach Peter suggested:

```{r}
filter_cells_by_total_membership <- function (L, max_val = 10, numiter = 10) {
  n <- nrow(L)
  rows <- 1:n
  for (iter in 1:numiter) {
    x <- rowSums(L)
    cat(sprintf("%d. Filtered out %d cells.\n",iter,sum(x > max_val)))
    i <- which(x <= max_val)
    L <- L[i,]
    rows <- rows[i]
    d <- apply(L,2,max)
    L <- scale_cols(L,1/d)
  }
  return(rows)
}
scale_cols <- function (A, b)
  t(t(A) * b)
```


```{r}
D <- diag(1 / apply(L_pm, 2, function(x) max(x)))
L <- L_pm %*% D
cells <- filter_cells_by_total_membership(L,numiter = 12)
seurat_meta <- seurat_meta[cells,]
L_pm <- L_pm[cells,]
```


```{r}
d <- apply(L_pm,2,max)
L_pm <- scale_cols(L_pm,1/d)
D <- d * D
```


Only keep factors with pve larger than $10^{-4}$.

```{r}
L_pm_old <- L_pm
L_pm <- L_pm[, flashier_snmf_summary$pve > 1e-4]
colnames(L_pm) <- paste0("K", which(flashier_snmf_summary$pve > 1e-4))
```



## Annotate the factors {#annotate_factors}

To annotate each factor with the meta-information, we will try two approaches:
- \textbf{Variance explained}: for each meta-variable (e.g., cell type, organ, condition), we will run ANOVA to see how much variation (measured in $R^2$) in the factor is explained by each meta-variable.
- \textbf{Maximal AUC}: for each meta-variable (e.g., cell type, organ, condition), we will compute the AUC of each factor for predicting each group in the meta-variable, and keep the maximal AUC.


### Cell-type Level1: {#level_1_annotation}

As discussed with David, all T-cells were differentiated from thymocytes, so we should remove the thymocytes from the analysis.
Also, to avoid the complication, let's also remove all unhealthy cells.

```{r}
cells_thymocyte <- which(seurat_meta$annotation_level1 == "thymocyte")
cells_unhealthy_thymocyte <- which(seurat_meta$annotation_level1 == "thymocyte" | seurat_meta$condition_broad != "healthy")
L_pm_old_no_thymocytes <- L_pm_old[-cells_thymocyte, ]
L_pm_old_no_unhealthy_thymocytes <- L_pm_old[-cells_unhealthy_thymocyte, ]
L_pm_no_unhealthy_thymocytes <- L_pm[-cells_unhealthy_thymocyte, ]
```


#### Variance explained: {#Variance_Explained_level_1}

```{r eval = FALSE}
annot_level1 <- ANOVA_factors(L_pm_no_unhealthy_thymocytes, seurat_meta$annotation_level1[-cells_unhealthy_thymocyte])
annot_level1$factor_original <- which(flashier_snmf_summary$pve > 1e-4)[annot_level1$factor]
saveRDS(annot_level1, file = paste0(data_path, "annot_level1.rds"))
```

```{r}
annot_level1 <- readRDS(paste0(data_path, "annot_level1.rds"))
plot(annot_level1$rank, annot_level1$stats, type = "o", xlab = "Rank", ylab = "R2", main = "Celltype Level1", ylim = c(0,1.1))
text(annot_level1$rank, annot_level1$stats, labels = annot_level1$factor_original, pos = 3, cex = 0.5)
abline(h = c(0.1,0.2,0.4), col = "red", lty = 2)
```


#### Maximal AUC: {#Maximal_AUC_level_1}

```{r eval = FALSE}
level_1_AUC <- compute_auc_matrix(loading_mat = L_pm_old_no_unhealthy_thymocytes, group_info = seurat_meta$annotation_level1[-cells_unhealthy_thymocyte])
rownames(level_1_AUC) <- sort(unique(seurat_meta$annotation_level1[-cells_unhealthy_thymocyte]))
saveRDS(level_1_AUC, file = paste0(data_path, "level_1_AUC.rds"))
```


As discussed with David, Treg should really be thought as a subset of CD4 cells. 
Therefore, let's define a new category "all_CD4" which includes both "CD4" and "Treg", and recompute the AUC:

```{r eval = FALSE}
annotation_level1_all_CD4 <- as.character(seurat_meta$annotation_level1[-cells_unhealthy_thymocyte])
annotation_level1_all_CD4[annotation_level1_all_CD4 == "Treg"] <- "CD4"
annotation_level1_all_CD4[annotation_level1_all_CD4 != "CD4"] <- "other"
level_1_AUC_all_CD4 <- compute_auc_matrix(loading_mat = L_pm_old_no_unhealthy_thymocytes,
                                          group_info = annotation_level1_all_CD4)
level_1_AUC <- readRDS(paste0(data_path, "level_1_AUC.rds"))
level_1_AUC <- rbind(level_1_AUC, level_1_AUC_all_CD4[1, , drop = FALSE])
rownames(level_1_AUC)[nrow(level_1_AUC)] <- "all_CD4"
saveRDS(level_1_AUC, file = paste0(data_path, "level_1_AUC.rds"))
```


Furthermore, some of the gdT cells (gdT_cl2, 4, 11, 12, 38, 25) should be viewed as CD8aa.

```{r eval = FALSE}
annotation_level1_all_CD8aa <- as.character(seurat_meta$annotation_level1[-cells_unhealthy_thymocyte])
annotation_level2_all_CD8aa <- as.character(seurat_meta$annotation_level2[-cells_unhealthy_thymocyte])
# if level 2 is gdT_cl2, 4, 11, 12, 38, 25, then change level 1 to CD8aa
annotation_level1_all_CD8aa[annotation_level2_all_CD8aa %in% paste0("gdT_cl", c(2,4,11,12,25,38))] <- "CD8aa"
annotation_level1_all_CD8aa[annotation_level1_all_CD8aa != "CD8aa"] <- "other"
level_1_AUC_all_CD8aa <- compute_auc_matrix(loading_mat = L_pm_old_no_unhealthy_thymocytes,
                                            group_info = annotation_level1_all_CD8aa)
level_1_AUC <- readRDS(paste0(data_path, "level_1_AUC.rds"))
level_1_AUC <- rbind(level_1_AUC, level_1_AUC_all_CD8aa[1, , drop = FALSE])
rownames(level_1_AUC)[nrow(level_1_AUC)] <- "all_CD8aa"
saveRDS(level_1_AUC, file = paste0(data_path, "level_1_AUC.rds"))
```

```{r}
annotation_level1_all_CD4 <- as.character(seurat_meta$annotation_level1)[-cells_unhealthy_thymocyte]
annotation_level1_all_CD4[annotation_level1_all_CD4 == "Treg"] <- "CD4"
annotation_level1_all_CD4[annotation_level1_all_CD4 != "CD4"] <- "other"

# boxplot of K58 on all_CD4
boxplot(L_pm_old_no_unhealthy_thymocytes[, 58] ~ annotation_level1_all_CD4, main
 = "K58", ylab = "membership", outline = FALSE)

# boxplot of K29 on all_CD4
boxplot(L_pm_old_no_unhealthy_thymocytes[, 29] ~ annotation_level1_all_CD4, main =
 "K29", ylab = "membership", outline = FALSE)
```

```{r}
annotation_level1_all_CD8aa <- as.character(seurat_meta$annotation_level1[-cells_unhealthy_thymocyte])
annotation_level2_all_CD8aa <- as.character(seurat_meta$annotation_level2[-cells_unhealthy_thymocyte])
# if level 2 is gdT_cl2, 4, 11, 12, 38, 25, then change level 1 to CD8aa
annotation_level1_all_CD8aa[annotation_level2_all_CD8aa %in% paste0("gdT_cl", c(2,4,11,12,25,38))] <- "CD8aa"
annotation_level1_all_CD8aa[annotation_level1_all_CD8aa != "CD8aa"] <- "other"

# boxplot of K3 on all_CD8aa
boxplot(L_pm_old_no_unhealthy_thymocytes[, 3] ~ annotation_level1_all_CD8aa, main = "K3", ylab = "membership", outline = FALSE)

# boxplot of K171 on all_CD8aa
boxplot(L_pm_old_no_unhealthy_thymocytes[, 171] ~ annotation_level1_all_CD8aa, main = "K171", ylab = "membership", outline = FALSE)
```


```{r}
level_1_AUC <- readRDS(paste0(data_path, "level_1_AUC.rds"))
level_1_AUC_max <- apply(level_1_AUC, 2, max)
level_1_AUC_max_name <- apply(level_1_AUC, 2, function(x) rownames(level_1_AUC)[which.max(x)])
o <- order(level_1_AUC_max, decreasing = TRUE)
table_level_1_AUC <- data.frame(
  Factor = colnames(level_1_AUC)[o],
  Max_AUC = level_1_AUC_max[o],
  Annotation = level_1_AUC_max_name[o]
)
# keep rows that are at least 0.8
knitr::kable(table_level_1_AUC[level_1_AUC_max[o] >= 0.8, ])
```


```{r}
table_level_1_AUC_pve <- table_level_1_AUC[table_level_1_AUC$Factor %in% colnames(L_pm),]
# color by table_level_1_AUC_pve$Annotation
colors <- RColorBrewer::brewer.pal(length(unique(table_level_1_AUC_pve$Annotation)), "Set1")
names(colors) <- unique(table_level_1_AUC_pve$Annotation)

table_level_1_AUC_pve <- table_level_1_AUC[table_level_1_AUC$Factor %in% colnames(L_pm),]
plot(1:nrow(table_level_1_AUC_pve), table_level_1_AUC_pve$Max_AUC, type = "o", xlab = "Rank by max AUC", ylab = "Max AUC", main = "Celltype Level1", col = colors[table_level_1_AUC_pve$Annotation], pch = 16)
text(1:nrow(table_level_1_AUC_pve), table_level_1_AUC_pve$Max_AUC, labels = gsub("K", "", table_level_1_AUC_pve$Factor), pos = 3, cex = 0.45, col = colors[table_level_1_AUC_pve$Annotation])
legend("topright", legend = names(colors), col = colors, pch = 16, cex = 1)
```

Highlight factors with maximal AUC at least $0.8$:

```{r}
highlights <- data.frame(
  AUC = level_1_AUC_max[level_1_AUC_max >= 0.80],
  Annotation = level_1_AUC_max_name[level_1_AUC_max >= 0.80]
)

# Arranging by AUC value
highlights <- highlights[order(highlights$AUC, decreasing = TRUE), ]
knitr::kable(highlights)
```

Structure plot of the highlighted factors:

```{r, results="hide", message=FALSE}
set.seed(12345)
cell_type_factors <- as.numeric(gsub("K", rownames(highlights), replacement = ""))
fit2 <- L_pm_old_no_unhealthy_thymocytes[, cell_type_factors, drop = FALSE]
colnames(fit2) <- paste0("k", cell_type_factors)

cell_type <- seurat_meta$annotation_level1[-cells_unhealthy_thymocyte]
cells <- which(cell_type == "CD4" | cell_type == "CD8")
cells <- sample(cells,1e5)
cells <- sort(c(cells,which(cell_type != "CD4" & cell_type != "CD8")))

structure_plot(fit2[cells,],gap = 10,n = 3000,
                    grouping = cell_type[cells]) +
  labs(y = "membership",color = "",fill = "") +
  ylim(0,2) +
  theme(legend.position = "bottom")
```



### Organ: {#organ_annotation}

Again, as discussed with David, we will focus on healthy cells:

```{r}
cells_unhealthy <- which(seurat_meta$condition_broad != "healthy")
L_pm_old_no_unhealthy <- L_pm_old[-cells_unhealthy, ]
L_pm_no_unhealthy <- L_pm[-cells_unhealthy, ]
```


#### Variance explained:

```{r eval = FALSE}
organ <- ANOVA_factors(L_pm_no_unhealthy, seurat_meta$organ_simplified[-cells_unhealthy])
organ$factor_original <- which(flashier_snmf_summary$pve > 1e-4)[organ$factor]
saveRDS(organ, file = paste0(data_path, "organ.rds"))
```

```{r}
organ <- readRDS(paste0(data_path, "organ.rds"))
plot(organ$rank, organ$stats, type = "o", xlab = "Rank", ylab = "R2", main = "organ", ylim = c(0,1.1))
text(organ$rank, organ$stats, labels = organ$factor_original, pos = 3, cex = 0.5)
abline(h = c(0.1,0.2,0.4), col = "red", lty = 2)
```

#### Maximal AUC: {#Maximal_AUC_organ}

```{r eval = FALSE}
organ_AUC <- compute_auc_matrix(loading_mat = L_pm_old_no_unhealthy, group_info = seurat_meta$organ_simplified[-cells_unhealthy])
rownames(organ_AUC) <- sort(unique(seurat_meta$organ_simplified[-cells_unhealthy]))
saveRDS(organ_AUC, file = paste0(data_path, "organ_AUC.rds"))
```

```{r}
organ_AUC <- readRDS(paste0(data_path, "organ_AUC.rds"))
organ_AUC_max <- apply(organ_AUC, 2, max)
organ_AUC_max_name <- apply(organ_AUC, 2, function(x) rownames(organ_AUC)[which.max(x)])
o <- order(organ_AUC_max, decreasing = TRUE)
table_organ_AUC <- data.frame(
  Max_AUC = organ_AUC_max[o],
  Annotation = organ_AUC_max_name[o]
)
```

```{r}
table_organ_AUC_pve <- table_organ_AUC[rownames(table_level_1_AUC) %in% colnames(L_pm), ]
annotations <- unique(table_organ_AUC_pve$Annotation)
colors <- grDevices::colorRampPalette(RColorBrewer::brewer.pal(9, "Set3"))(length(annotations))
names(colors) <- annotations
plot(1:nrow(table_organ_AUC_pve), table_organ_AUC_pve$Max_AUC, type = "o", xlab = "Rank by max AUC", ylab = "Max AUC", main = "Organ", col = colors[table_organ_AUC_pve$Annotation], pch = 16)
text(1:nrow(table_organ_AUC_pve), table_organ_AUC_pve$Max_AUC, labels = gsub("K", "", rownames(table_organ_AUC_pve)), pos = 3, cex = 0.5, col = colors[table_organ_AUC_pve$Annotation])
legend("topright", legend = names(colors), col = colors, pch = 16, cex = 1)
```

Keep factors with maximal AUC at least $0.85$, and remove factors that correspond to synovial fluid or CNS:

```{r}
highlights <- table_organ_AUC[organ_AUC_max[o] >= 0.85, ]
highlights <- highlights[!grepl("synovial", highlights$Annotation), ]
highlights <- highlights[!grepl("CNS", highlights$Annotation), ]
# produce a table
knitr::kable(highlights)
```

Structure plot:

```{r, results="hide", message=FALSE}
set.seed(12345)
organ_factors <- as.numeric(gsub("K", rownames(highlights), replacement = ""))
fit2 <- L_pm_old_no_unhealthy[, organ_factors, drop = FALSE]
colnames(fit2) <- paste0("K", organ_factors)
organ <- seurat_meta$organ_simplified[-cells_unhealthy]
cells <- sort(c(sample(which(organ == "spleen"),1e4),
                sample(which(organ == "LN"),1e4),
                sample(which(organ == "thymus"),3e3),
                which(organ != "spleen" & organ != "LN" & organ != "thymus")))
structure_plot(fit2[cells,],gap = 5,n = 3000,
                    grouping = organ[cells]) +
  labs(y = "membership",color = "",fill = "") +
  ylim(0,2) +
  theme(legend.position = "bottom")
```


#### Stratified Analysis {#Stratified_Analysis_organ}

Look at the boxplot for K37 (mammary gland specific) and K6 (skin specific):

```{r}
data_boxplot <- fit2[,c("K6", "K37")]
is_mammary <- seurat_meta$organ_simplified[-cells_unhealthy] == "mammary gland"
is_skin <- seurat_meta$organ_simplified[-cells_unhealthy] == "skin"
boxplot(data_boxplot[,1] ~ is_skin, main = "K6", ylab = "membership", outline = FALSE, names = c("other","skin"))
boxplot(data_boxplot[,2] ~ is_mammary, main = "K37", ylab = "membership", outline = FALSE, names = c("other","mammary gland"))
```

A more detailed boxplot for each:

```{r}
cells_organ_healthy <- seurat_meta$organ_simplified[-cells_unhealthy]
boxplot(data_boxplot[,1] ~ cells_organ_healthy, 
        main = "K6", ylab = "membership", outline = FALSE, las = 2,
        xlab = "", cex.axis = 0.7
        )
boxplot(data_boxplot[,2] ~ cells_organ_healthy, 
        main = "K37", ylab = "membership", outline = FALSE, las = 2,
        xlab = "", cex.axis = 0.7
        )
```

What if we further condition on a specific level 2 annotation?

```{r}
subset_cells <- which(seurat_meta$annotation_level2[-cells_unhealthy] == "Treg_cl4")

boxplot(data_boxplot[subset_cells,1] ~ cells_organ_healthy[subset_cells], 
        main = "K6 (subset_cells)", ylab = "membership", outline = FALSE, las = 2,
        xlab = "", cex.axis = 0.7
)

boxplot(data_boxplot[subset_cells,2] ~ cells_organ_healthy[subset_cells], 
        main = "K37 (subset_cells)", ylab = "membership", outline = FALSE, las = 2,
        xlab = "", cex.axis = 0.7
)
```

### Condition: {#condition_annotation}

#### Variance explained: {#Variance_Explained_condition}

```{r eval = FALSE}
condition_broad <- ANOVA_factors(L_pm, seurat_meta$condition_broad)
condition_broad$factor_original <- which(flashier_snmf_summary$pve > 1e-4)[condition_broad$factor]
saveRDS(condition_broad, file = paste0(data_path, "condition_broad.rds"))
```

```{r}
condition_broad <- readRDS(paste0(data_path, "condition_broad.rds"))
plot(condition_broad$rank, condition_broad$stats, type = "o", xlab = "Rank", ylab = "R2", main = "condition_broad", ylim = c(0,1.1))
text(condition_broad$rank, condition_broad$stats, labels = condition_broad$factor_original, pos = 3, cex = 0.5)
abline(h = c(0.1,0.2,0.4), col = "red", lty = 2)
```


#### Maximal AUC: {#Maximal_AUC_condition}

```{r eval = FALSE}
condition_broad_AUC <- compute_auc_matrix(loading_mat = L_pm_old, group_info = seurat_meta$condition_broad)
rownames(condition_broad_AUC) <- sort(unique(seurat_meta$condition_broad))
saveRDS(condition_broad_AUC, file = paste0(data_path, "condition_broad_AUC.rds"))
```

```{r}
condition_broad_AUC <- readRDS(paste0(data_path, "condition_broad_AUC.rds"))
condition_broad_AUC_max <- apply(condition_broad_AUC, 2, max)
condition_broad_AUC_max_name <- apply(condition_broad_AUC, 2, function(x) rownames(condition_broad_AUC)[which.max(x)])
```

```{r}
o <- order(condition_broad_AUC_max, decreasing = TRUE)
table_condition_broad_AUC <- data.frame(
  Max_AUC = condition_broad_AUC_max[o],
  Annotation = condition_broad_AUC_max_name[o]
)
```

```{r}
table_condition_broad_AUC_pve <- table_condition_broad_AUC[rownames(table_condition_broad_AUC) %in% colnames(L_pm), ]
annotations <- unique(table_condition_broad_AUC_pve$Annotation)
colors <- grDevices::colorRampPalette(RColorBrewer::brewer.pal(9, "Set3"))(length(annotations))
names(colors) <- annotations
plot(1:nrow(table_condition_broad_AUC_pve), table_condition_broad_AUC_pve$Max_AUC, type = "o", xlab = "Rank by max AUC", ylab = "Max AUC", main = "Condition Broad", col = colors[table_condition_broad_AUC_pve$Annotation], pch = 16)
text(1:nrow(table_condition_broad_AUC_pve), table_condition_broad_AUC_pve$Max_AUC, labels = gsub("K", "", rownames(table_condition_broad_AUC_pve)), pos = 3, cex = 0.5, col = colors[table_condition_broad_AUC_pve$Annotation])
legend("topright", legend = names(colors), col = colors, pch = 16, cex = 1)
```

Highlight factors with maximal AUC at least $0.85$:

```{r}
highlights <- table_condition_broad_AUC[condition_broad_AUC_max[o] >= 0.85, ]
# produce a table
knitr::kable(highlights)
```

Structure plot:

```{r, results="hide", message=FALSE}
set.seed(12345)
condition_broad_factors <- as.numeric(gsub("K", rownames(highlights), replacement = ""))
fit2 <- L_pm_old[, condition_broad_factors, drop = FALSE]
colnames(fit2) <- paste0("K", condition_broad_factors)
condition_broad <- seurat_meta$condition_broad
cells <- sort(c(sample(which(condition_broad == "healthy"),5e4),
                which(condition_broad != "healthy")))
structure_plot(fit2[cells, , drop = FALSE],gap = 5,n = 3000,
                    grouping = condition_broad[cells]) +
  labs(y = "membership",color = "",fill = "") +
  ylim(0,2) +
  theme(legend.position = "bottom")
```



#### Further conditioning {#Further_conditioning}

We will redo the AUC analysis when we condition on CD8 cells only:

```{r}
cells_CD8 <- which(seurat_meta$annotation_level1 == "CD8")
L_pm_old_CD8 <- L_pm_old[cells_CD8, ]
L_pm_CD8 <- L_pm[cells_CD8, ]
condition_broad_CD8 <- seurat_meta$condition_broad[cells_CD8]
```

```{r eval = FALSE}
condition_broad_AUC_CD8 <- compute_auc_matrix(loading_mat = L_pm_old_CD8, 
                                              group_info = condition_broad_CD8)
rownames(condition_broad_AUC_CD8) <- sort(unique(condition_broad_CD8))
saveRDS(condition_broad_AUC_CD8, file = paste0(data_path, "condition_broad_AUC_CD8.rds"))
```

```{r}
condition_broad_AUC_CD8 <- readRDS(paste0(data_path, "condition_broad_AUC_CD8.rds"))
condition_broad_AUC_CD8_max <- apply(condition_broad_AUC_CD8, 2, max)
condition_broad_AUC_CD8_max_name <- apply(condition_broad_AUC_CD8, 2, function(x)
  rownames(condition_broad_AUC_CD8)[which.max(x)])
o <- order(condition_broad_AUC_CD8_max, decreasing = TRUE)
table_condition_broad_AUC_CD8 <- data.frame(Max_AUC = condition_broad_AUC_CD8_max[o], Annotation = condition_broad_AUC_CD8_max_name[o])
```
                                                                                             
```{r}
table_condition_broad_AUC_CD8_pve <- table_condition_broad_AUC_CD8
table_condition_broad_AUC_CD8_pve <- table_condition_broad_AUC_CD8_pve[rownames(table_condition_broad_AUC_CD8_pve) %in% colnames(L_pm), ]
annotations <- unique(table_condition_broad_AUC_CD8_pve$Annotation)
colors <- grDevices::colorRampPalette(RColorBrewer::brewer.pal(9,"Set3"))(length(annotations))
names(colors) <- annotations
plot(
  1:nrow(table_condition_broad_AUC_CD8_pve),
  table_condition_broad_AUC_CD8_pve$Max_AUC,
  type = "o",
  xlab = "Rank by max AUC",
  ylab = "Max AUC",
  main = "Condition Broad (CD8)",
  col = colors[table_condition_broad_AUC_CD8_pve$Annotation],
  pch = 16
)
text(
  1:nrow(table_condition_broad_AUC_CD8_pve),
  table_condition_broad_AUC_CD8_pve$Max_AUC,
  labels = gsub("K", "", rownames(table_condition_broad_AUC_CD8_pve)),
  pos = 3,
  cex = 0.5,
  col = colors[table_condition_broad_AUC_CD8_pve$Annotation]
)
legend(
  "topright",
  legend = names(colors),
  col = colors,
  pch = 16
  ,
  cex = 1
)
```

```{r}
highlights <- table_condition_broad_AUC_CD8[condition_broad_AUC_CD8_max[o] >= 0.85, ]
# produce a table
knitr::kable(highlights)
```

Structure plot:

```{r, results="hide", message=FALSE}
set.seed(12345)
condition_broad_CD8_factors <- as.numeric(gsub("K", rownames(highlights), replacement = ""))
fit2 <- L_pm_old_CD8[, condition_broad_CD8_factors, drop = FALSE]
colnames(fit2) <- paste0("K", condition_broad_CD8_factors)
cells <- sort(c(sample(which(condition_broad_CD8 == "healthy"),1e3),
                sample(which(condition_broad_CD8 == "virus"),1e3),
                sample(which(condition_broad_CD8 == "bacteria"),1e3),
                sample(which(condition_broad_CD8 == "autoimmune"),1e3),
                sample(which(condition_broad_CD8 == "cancer"),1e3),
                sample(which(condition_broad_CD8 == "mutiinfection"),1e3),
                which(condition_broad_CD8 != "healthy" & condition_broad_CD8 != "virus" & condition_broad_CD8 != "bacteria" & condition_broad_CD8 != "autoimmune" & condition_broad_CD8 != "cancer" & condition_broad_CD8 != "mutiinfection")))
structure_plot(fit2[cells, , drop = FALSE], gap = 5,n = 3000,
                    grouping = condition_broad_CD8[cells]) +
  labs(y = "membership",color = "",fill = "") +
  ylim(0,2) +
  theme(legend.position = "bottom")
```



### Condition_detailed: {#condition_detailed_annotation}

#### Variance explained: {#Variance_Explained_condition_detailed}

```{r}
# Focus on group with at least 3000 cells
selected_conditions <- names(which(table(seurat_meta$condition_detailed) >= 3000))
selected_cells <- which(seurat_meta$condition_detailed %in% selected_conditions)
```


```{r eval = FALSE}
condition_detailed <- ANOVA_factors(L_pm[selected_cells, ], seurat_meta$condition_detailed[selected_cells])
condition_detailed$factor_original <- which(flashier_snmf_summary$pve > 1e-4)[condition_detailed$factor]
saveRDS(condition_detailed, file = paste0(data_path, "condition_detailed.rds"))
```

```{r}
condition_detailed <- readRDS(paste0(data_path, "condition_detailed.rds"))
plot(condition_detailed$rank, condition_detailed$stats, type = "o", xlab = "Rank", ylab = "R2", main = "condition_detailed", ylim = c(0,1.1))
text(condition_detailed$rank, condition_detailed$stats, labels = condition_detailed$factor_original, pos = 3, cex = 0.5)
abline(h = c(0.1,0.2,0.4), col = "red", lty = 2)
```

#### Maximal AUC: {#Maximal_AUC_condition_detailed}

```{r eval = FALSE}
condition_detailed_AUC <- compute_auc_matrix(loading_mat = L_pm_old, group_info = seurat_meta$condition_detailed)

rownames(condition_detailed_AUC) <- sort(unique(seurat_meta$condition_detailed))

saveRDS(condition_detailed_AUC, file = paste0(data_path, "condition_detailed_AUC.rds"))
```


```{r}
condition_detailed_AUC <- readRDS(paste0(data_path, "condition_detailed_AUC.rds"))
condition_detailed_AUC_max <- apply(condition_detailed_AUC, 2, max)
condition_detailed_AUC_max_name <- apply(condition_detailed_AUC, 2, function(x) rownames(condition_detailed_AUC)[which.max(x)])
o <- order(condition_detailed_AUC_max, decreasing = TRUE)
table_condition_detailed_AUC <- data.frame(
  Max_AUC = condition_detailed_AUC_max[o],
  Annotation = condition_detailed_AUC_max_name[o]
)
```

```{r}
table_condition_detailed_AUC_pve <- table_condition_detailed_AUC[rownames(table_condition_detailed_AUC) %in% colnames(L_pm), ]

# Only focus on conditions with at least 3000 cells
table_condition_detailed_AUC_pve <- table_condition_detailed_AUC_pve[table_condition_detailed_AUC_pve$Annotation %in% selected_conditions, ]

annotations <- unique(table_condition_detailed_AUC_pve$Annotation)
colors <- grDevices::colorRampPalette(RColorBrewer::brewer.pal(9, "Set3"))(length(annotations))
names(colors) <- annotations
plot(1:nrow(table_condition_detailed_AUC_pve), table_condition_detailed_AUC_pve$Max_AUC, type = "o", xlab = "Rank by max AUC", ylab = "Max AUC", main = "Condition Detailed", col = colors[table_condition_detailed_AUC_pve$Annotation], pch = 16)
text(1:nrow(table_condition_detailed_AUC_pve), table_condition_detailed_AUC_pve$Max_AUC, labels = gsub("K", "", rownames(table_condition_detailed_AUC_pve)), pos = 3, cex = 0.7, col = colors[table_condition_detailed_AUC_pve$Annotation])
legend("topright", legend = names(colors), col = colors, pch = 16, cex = 0.85)
```

Highlight factors with maximal AUC at least $0.85$:

```{r}
highlights <- table_condition_detailed_AUC_pve[table_condition_detailed_AUC_pve$Max_AUC >= 0.85, ]
# produce a table
knitr::kable(highlights)
```

structure plot:

```{r, results="hide", message=FALSE}
set.seed(12345)
condition_detailed_factors <- as.numeric(gsub("K", rownames(highlights), replacement = ""))
fit2 <- L_pm_old[, condition_detailed_factors, drop = FALSE]
colnames(fit2) <- paste0("K", condition_detailed_factors)
condition_detailed <- seurat_meta$condition_detailed
# subsample only 10000 baseline cells
cells <- sort(c(sample(which(condition_detailed == "baseline"),1e4),
                which(condition_detailed != "baseline")))
structure_plot(fit2[cells, , drop = FALSE],gap = 5,n = 3000,
               
                    grouping = condition_detailed[cells]) +
  labs(y = "membership",color = "",fill = "") +
  ylim(0,2) +
  theme(legend.position = "bottom")
```


#### Further conditioning {#Further_conditioning_condition_detailed}

We will redo the AUC analysis when we condition on CD8 cells only:

```{r}
cells_CD8 <- which(seurat_meta$annotation_level1 == "CD8")
L_pm_old_CD8 <- L_pm_old[cells_CD8, ]
L_pm_CD8 <- L_pm[cells_CD8, ]
condition_detailed_CD8 <- seurat_meta$condition_detailed[cells_CD8]
```

```{r eval = FALSE}
condition_detailed_AUC_CD8 <- compute_auc_matrix(loading_mat = L_pm_old_CD8, 
                                              group_info = condition_detailed_CD8)
rownames(condition_detailed_AUC_CD8) <- sort(unique(condition_detailed_CD8))
saveRDS(condition_detailed_AUC_CD8, file = paste0(data_path, "condition_detailed_AUC_CD8.rds"))
```

```{r}
condition_detailed_AUC_CD8 <- readRDS(paste0(data_path, "condition_detailed_AUC_CD8.rds"))
condition_detailed_AUC_CD8_max <- apply(condition_detailed_AUC_CD8, 2, max)
condition_detailed_AUC_CD8_max_name <- apply(condition_detailed_AUC_CD8, 2, function(x)
  rownames(condition_detailed_AUC_CD8)[which.max(x)])
o <- order(condition_detailed_AUC_CD8_max, decreasing = TRUE)
table_condition_detailed_AUC_CD8 <- data.frame(Max_AUC = condition_detailed_AUC_CD8_max[o], Annotation = condition_detailed_AUC_CD8_max_name[o])
```
                                                                                             
```{r}
table_condition_detailed_AUC_CD8_pve <- table_condition_detailed_AUC_CD8
table_condition_detailed_AUC_CD8_pve <- table_condition_detailed_AUC_CD8_pve[rownames(table_condition_detailed_AUC_CD8_pve) %in% colnames(L_pm), ]
annotations <- unique(table_condition_detailed_AUC_CD8_pve$Annotation)
colors <- grDevices::colorRampPalette(RColorBrewer::brewer.pal(9,"Set3"))(length(annotations))
names(colors) <- annotations
plot(
  1:nrow(table_condition_detailed_AUC_CD8_pve),
  table_condition_detailed_AUC_CD8_pve$Max_AUC,
  type = "o",
  xlab = "Rank by max AUC",
  ylab = "Max AUC",
  main = "Condition Broad (CD8)",
  col = colors[table_condition_detailed_AUC_CD8_pve$Annotation],
  pch = 16
)
text(
  1:nrow(table_condition_detailed_AUC_CD8_pve),
  table_condition_detailed_AUC_CD8_pve$Max_AUC,
  labels = gsub("K", "", rownames(table_condition_detailed_AUC_CD8_pve)),
  pos = 3,
  cex = 0.5,
  col = colors[table_condition_detailed_AUC_CD8_pve$Annotation]
)
legend(
  "topright",
  legend = names(colors),
  col = colors,
  pch = 16
  ,
  cex = 1
)
```

```{r}
highlights <- table_condition_detailed_AUC_CD8[condition_detailed_AUC_CD8_max[o] >= 0.85, ]
# Remove terms less than 100
terms_less_100 <- names(table(condition_detailed_CD8))[table(condition_detailed_CD8) < 100]
highlights <- highlights[!highlights$Annotation %in% terms_less_100, ]
# produce a table
knitr::kable(highlights)
```





