---
title: Better Structure plot
author: Peter Carbonetto
output: workflowr::wflow_html
---

This is my attempt to make a better, more unified Structure plot from
the semi-NMF model fit with $K = 200$.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load the packages needed for this analysis:

```{r load-pkgs, message=FALSE}
library(Matrix)
library(fastTopics)
library(flashier)
library(ggplot2)
library(cowplot)
```

Load the meta-data and the semi-NMF results:

```{r load-data}
load("../data/flashier_snmf_ldf.RData")
```

This is a function implementing the iterative filtering procedure:

```{r filter-cells-by-total-membership-function}
filter_cells_by_total_membership <- function (L, max_val = 10, numiter = 10) {
  n <- nrow(L)
  rows <- 1:n
  for (iter in 1:numiter) {
    x <- rowSums(L)
    cat(sprintf("%d. Filtered out %d cells.\n",iter,sum(x > max_val)))
    i <- which(x <= max_val)
    L <- L[i,]
    rows <- rows[i]
    d <- apply(L,2,max)
    L <- scale_cols(L,1/d)
  }
  return(rows)
}
```

Here is the missing function `scale_cols`:

```{r scale-cols-function}
scale_cols <- function (A, b)
  t(t(A) * b)
```

Apply the iterative filtering procedure to the ${\bf L}$
matrix from the semi-NMF fit:

```{r filter-cells}
L <- fl_snmf_ldf$L
cells <- filter_cells_by_total_membership(L,numiter = 12)
sample_info <- sample_info[cells,]
L <- L[cells,]
```

And now renormalize the memberships:

```{r renormalize-memberships}
d <- apply(L,2,max)
L <- scale_cols(L,1/d)
```

Remove "miniverse" and "proliferative" cells:

```{r filter-cells-2}
sample_info <-
  transform(sample_info,
    miniverse = grepl("miniverse",sample_info$annotation_level2,fixed = TRUE),
    prolif    = grepl("prolif",sample_info$annotation_level2,fixed = TRUE))
cells <- which(with(sample_info,!(miniverse | prolif)))
sample_info <- sample_info[cells,]
L <- L[cells,]
```

This is the number of cells per major T cell type:

```{r major-cell-types}
table(sample_info$annotation_level1)
```

Let's assign the cells to the different Structure plots:

```{r assign-to-structure-plots}
sample_info$structure_plot <- 4
sample_info[sample_info$annotation_level1 == "CD4","structure_plot"]  <- 1
sample_info[sample_info$annotation_level1 == "CD8","structure_plot"]  <- 1
sample_info[sample_info$annotation_level1 == "Treg","structure_plot"] <- 2
sample_info[sample_info$annotation_level1 == "gdT","structure_plot"]  <- 3
```

Find a more informative labeling:

```{r annotation-for-structure-plots}
sample_info$label <- as.character(sample_info$annotation_level2)
sample_info[sample_info$annotation_level1 == "CD4","label"]     <- "CD4"
sample_info[sample_info$annotation_level1 == "CD8","label"]     <- "CD8"
sample_info[sample_info$annotation_level1 == "CD8aa","label"]   <- "CD8aa"
sample_info[sample_info$annotation_level1 == "DN","label"]      <- "DN"
sample_info[sample_info$annotation_level1 == "DP","label"]      <- "DP"
sample_info[sample_info$annotation_level1 == "gdT","label"]     <- "gdT"
sample_info[sample_info$annotation_level1 == "nonconv","label"] <- "nonconv"
cells <- which(with(sample_info,
                    annotation_level1 == "CD8" &
					is.element(organ_simplified,c("CNS","lung","spleen",
					                              "uterus")) &
			        (grepl("LCMV",condition_detailed,fixed = TRUE) |
			         grepl("dirty",condition_detailed,fixed = TRUE) |
			         grepl("MCMV",condition_detailed,fixed = TRUE) |
					 grepl("Toxo",condition_detailed,fixed = TRUE))))
sample_info[cells,"label"] <- "CD8(L10)"
cells <- which(with(sample_info,
                    annotation_level1 == "gdT" &
					is.element(organ_simplified,
 				  	  c("colon LP","colon epi","skin","small intestine LP",
					    "small intestine epi"))))
sample_info[cells,"label"] <- "gdT(L3)"
cells <- which(with(sample_info,
                    annotation_level1 == "CD8aa" &
					is.element(organ_simplified,
 				  	  c("colon LP","colon epi","skin","small intestine LP",
					    "small intestine epi"))))
sample_info[cells,"label"] <- "CD8aa(L3)"
cells <- which(with(sample_info,
                    annotation_level1 == "gdT" &
					condition_detailed == "baseline" &
					organ_simplified == "skin"))
sample_info[cells,"label"] <- "gdT(skin/b)"
cells <- which(with(sample_info,
                    annotation_level1 == "gdT" &
					condition_detailed != "baseline" &
					organ_simplified == "skin"))
sample_info[cells,"label"] <- "gdT(skin/nb)"
cells <- which(with(sample_info,
                    annotation_level1 == "gdT" &
					is.element(organ_simplified,
					           c("lung","submandibular gland","LN"))))
sample_info[cells,"label"] <- "gdT(lung)"
cells <- which(with(sample_info,
                    is.element(annotation_level1,c("CD4","Treg")) &
			        (grepl("NP_OVA",condition_detailed,fixed = TRUE) |
			         grepl("NOD",condition_detailed,fixed = TRUE) |
			         grepl("LCMV",condition_detailed,fixed = TRUE))))
sample_info[cells,"label"] <- "L12"
sample_info <- transform(sample_info,label = factor(label))
```

Create the Structure plots:

```{r create-structure-plots, results="hide", message=FALSE}
set.seed(1)
n <- 4
clusters <- sample_info$label
factor_colors <-
  c("gainsboro","gold","magenta","red",
    "darkred","deepskyblue","darkblue","forestgreen","cyan",
	"#F6768E","#00538A","limegreen","darkmagenta","#FF8E00",
	"olivedrab")
names(factor_colors) <- c("k171","k3","k5","k7",
                          "k8","k10","k13","k19","k22",
                          "k30","k37","k58","k62","k68","k12")
factors_to_show <- c("k171","k12","k58","k22","k13","k3",
                     "k19","k37","k62","k68",
					 "k5","k7","k8","k30","k10")
structure_plots <- vector("list",5)
names(structure_plots) <- paste0("p",1:n)
for (i in 1:n) {
  cells <- which(sample_info$structure_plot == i)
  p <- structure_plot(L[cells,factors_to_show],
                      gap = 5,n = 2000,
                      topics = factors_to_show,
					  colors = factor_colors[factors_to_show],
                      grouping = factor(clusters[cells]),
		              font.size = 7) +
    labs(y = "membership",color = "",fill = "") +
    theme(legend.position = "bottom",
      	  plot.margin = unit(c(0,0,-0.75,0),"cm"),
		  element_text(angle = 45,hjust = 1,size = 7))
  if (i < n) {
    p <- p +
	  guides(fill = "none") +
      theme(plot.margin = unit(c(0,0,-0.75,0),"cm"))
  } else {
    p <- p + theme(plot.margin = unit(c(0,0,0,0),"cm"))
  }
  structure_plots[[i]] <- p
}
```

Here are the Structure plots:

```{r structure-plots, fig.height=5.5, fig.width=7.25, results="hide", message=FALSE}
plot_grid(structure_plots$p1,
          structure_plots$p2,
          structure_plots$p3,
          structure_plots$p4,
          nrow = 4,ncol = 1,
		  rel_heights = c(2,2,2,4.5))
```

Factor 22 remains a factor that does not seem to align well with any
of the annotations.
