---
title: "bi_cross_validation"
author: "Ziang Zhang"
date: "2025-12-22"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Predicting CITE-seq using scRNA GEPs

```{r}
# Load necessary libraries
library(ggplot2)
library(ggrepel)
library(patchwork)
library(cowplot)
library(fastTopics)
library(dplyr)
library(pheatmap)
library(scales)
library(scattermore)
library(qs)

# data_path <- "../data/"
data_path <- "data/"
full <- readRDS(paste0(data_path, "protein_flash_selected_summary_lognorm.rds"))
partA <- qs::qread(paste0(data_path, "fit1_summary.qs"))
partB <- readRDS(paste0(data_path, "GEP_flash_selected_summary_testB.rds"))
partC <- readRDS(paste0(data_path, "protein_flash_selected_summary_train_C.rds"))

protein_mat_normalized_lognorm <- readRDS(paste0(data_path, "protein_mat_normalized_lognorm.rds")) 
```

The best fitted CITE-seq matrix using GEP loadings:

```{r}
# included_cells <- intersect(rownames(partB$L_pm), rownames(full$L_pm))
L_pm <- full$L_pm
F_pm <- full$F_pm
predicted_citeseq <- L_pm %*% t(F_pm)
true_citeseq <- protein_mat_normalized_lognorm
```

```{r}
all_test_cells <- intersect(rownames(predicted_citeseq), rownames(true_citeseq))
all_test_genes <- intersect(colnames(predicted_citeseq), colnames(true_citeseq))

predicted_citeseq <- predicted_citeseq[all_test_cells, all_test_genes]
true_citeseq <- true_citeseq[all_test_cells, all_test_genes]
```

```{r}
residual_mat <- true_citeseq - predicted_citeseq

# aggregate the error by protein to rMSE
rMSE_by_protein <- apply(residual_mat, 2, function(x) sqrt(mean(x^2)))
rMSE_by_protein <- data.frame(protein = names(rMSE_by_protein),
                                     rMSE = rMSE_by_protein)

# compute the relative error by dividing the protein-level standard deviation
rel_rMSE_by_protein <- rMSE_by_protein %>%
  left_join(data.frame(protein = colnames(true_citeseq),
                       SD = apply(true_citeseq, 2, sd)),
            by = "protein") %>%
  mutate(rel_rMSE = rMSE / SD)

# plot the distribution of relative error
ggplot(rel_rMSE_by_protein, aes(x = rel_rMSE)) +
  geom_histogram(bins = 50, fill = "lightblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Relative rMSE by Protein",
       x = "Relative rMSE",
       y = "Frequency")

# aggregate the error by cell
rMSE_by_cell <- apply(residual_mat, 1, function(x) sqrt(mean(x^2)))
rMSE_by_cell <- data.frame(cell = names(rMSE_by_cell),
                           rMSE = rMSE_by_cell)
# compute the relative error by dividing the cell-level standard deviation
rel_rMSE_by_cell <- rMSE_by_cell %>%
  left_join(data.frame(cell = rownames(true_citeseq),
                       SD = apply(true_citeseq, 1, sd)),
            by = "cell") %>%
  mutate(rel_rMSE = rMSE / SD)
# plot the distribution of relative error by cell
ggplot(rel_rMSE_by_cell, aes(x = rel_rMSE)) +
  geom_histogram(bins = 50, fill = "lightgreen", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Relative rMSE by Cell",
       x = "Relative rMSE",
       y = "Frequency")
```

What proteins were not well predicted?

```{r}
poorly_predicted_proteins <- rel_rMSE_by_protein %>%
  arrange(desc(rel_rMSE)) %>%
  head(10)

poorly_predicted_proteins
```

Pick a protein, show the scatter plot of true vs predicted:

```{r}
# protein_to_plot <- "IL2RA.CD25"
# protein_to_plot <- "FR4"
# protein_to_plot <- "GITR.CD357"
protein_to_plot <- "CD8B"
# protein_to_plot <- "TCRVG1.1"
ggplot(data.frame(True = true_citeseq[, protein_to_plot],
                  Predicted = predicted_citeseq[, protein_to_plot]),
       aes(x = True, y = Predicted)) +
  coord_cartesian(xlim = c(0, max(true_citeseq[, protein_to_plot], predicted_citeseq[, protein_to_plot])),
                  ylim = c(0, max(true_citeseq[, protein_to_plot], predicted_citeseq[, protein_to_plot]))) +
  geom_scattermore(alpha = 0.3, pointsize = 1) +
  theme_minimal() +
  labs(title = paste("True vs Predicted for", protein_to_plot),
       x = "True Expression",
       y = "Predicted Expression") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  # also show the R2 and lm line
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  annotate("text", x = Inf, y = Inf, label = paste0("R² = ",
                                                  round(cor(true_citeseq[, protein_to_plot],
                                                            predicted_citeseq[, protein_to_plot])^2, 3)),
           hjust = 1.1, vjust = 1.5, size = 5, color = "black")
```

Compute the correlation between true and predicted for all proteins:

```{r}
cor_by_protein <- sapply(colnames(true_citeseq), function(protein) {
  cor(true_citeseq[, protein], predicted_citeseq[, protein])
})
sort(cor_by_protein)
```

Show a plot of correlation by protein:

```{r}
ggplot(data.frame(protein = names(cor_by_protein),
                  correlation = cor_by_protein),
       aes(x = reorder(protein, correlation), y = correlation)) +
  geom_bar(stat = "identity", fill = "lightcoral") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Correlation between True and Predicted by Protein",
       x = "Protein",
       y = "Correlation") +
  # make the text size smaller
  theme(axis.text.y = element_text(size = 4.5))
```

Show a plot of relative rMSE by protein:

```{r}
ggplot(rel_rMSE_by_protein, aes(x = reorder(protein, -rel_rMSE), y = rel_rMSE)) +
  geom_bar(stat = "identity", fill = "lightseagreen") +
  theme_minimal() +
  labs(title = "Relative rMSE by Protein",
       x = "Protein",
       y = "Relative rMSE") +
  # make the x-lim 0 to 1
  coord_flip(ylim = c(0, 1)) +
  # make the text size smaller
  theme(axis.text.y = element_text(size = 4.5))
```

What types of cells were not well predicted?

```{r}
seurat_meta <- readRDS(file = paste0(data_path, "seurat_meta.rds"))
cells_meta <- seurat_meta[rownames(seurat_meta) %in% all_test_cells, ]
```

Plot the distribution of relative rMSE by cell type (cells_meta$annotation_level1):

```{r}
rel_rMSE_by_cell <- rel_rMSE_by_cell %>%
  left_join(cells_meta, by = c("cell" = "cellID"))
ggplot(rel_rMSE_by_cell, aes(x = annotation_level1, y = rel_rMSE)) +
  geom_boxplot(fill = "lightyellow", outliers = FALSE) +
  theme_minimal() +
  labs(title = "Distribution of Relative rMSE by Cell Type",
       x = "Cell Type",
       y = "Relative rMSE") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


## Gene-Gene Analysis

```{r eval=FALSE}
first_half_summary <- qs::qread(paste0(data_path, "fit1_summary.qs"))
second_half_summary <- qs::qread(paste0(data_path, "fit2_summary.qs"))
L_pm <- second_half_summary$L_pm
F_pm <- first_half_summary$F_pm
true_expression <- second_half


Y_hat <- predicted_expression <- L_pm %*% t(F_pm[1, , drop = FALSE])
Y <- (true_expression[rownames(L_pm), 1])
rMSE <- sqrt(mean((Y-Y_hat)^2))

compute_gene_rMSE_all <- function(true_expression, L_pm, F_pm,
                                  genes = NULL,
                                  gene_block = 500L,
                                  na_rm = TRUE) {
  # Align cells
  cells <- rownames(L_pm)
  if (is.null(cells)) stop("L_pm must have rownames (cell IDs).")
  if (is.null(rownames(true_expression))) stop("true_expression must have rownames (cell IDs).")
  
  cells_use <- intersect(cells, rownames(true_expression))
  if (length(cells_use) == 0L) stop("No overlapping cell IDs between L_pm and true_expression.")
  if (length(cells_use) < length(cells)) {
    L_pm <- L_pm[cells_use, , drop = FALSE]
  }
  Y_all <- true_expression[cells_use, , drop = FALSE]
  
  # Decide which genes to compute
  if (is.null(genes)) {
    if (!is.null(colnames(Y_all)) && !is.null(rownames(F_pm))) {
      genes <- intersect(colnames(Y_all), rownames(F_pm))
    } else {
      # fallback: assume ordering matches by position
      if (ncol(Y_all) != nrow(F_pm)) {
        stop("Cannot infer gene mapping: need colnames(true_expression) and rownames(F_pm), or matching dimensions.")
      }
      genes <- seq_len(ncol(Y_all))
    }
  }
  
  # Pull the right Y and F in the same order
  if (is.character(genes)) {
    Y_all <- Y_all[, genes, drop = FALSE]
    F_use <- F_pm[genes, , drop = FALSE]
    gene_names <- genes
  } else {
    Y_all <- Y_all[, genes, drop = FALSE]
    F_use <- F_pm[genes, , drop = FALSE]
    gene_names <- colnames(Y_all)
    if (is.null(gene_names) && !is.null(rownames(F_use))) gene_names <- rownames(F_use)
  }
  
  m <- ncol(Y_all)
  rmse <- numeric(m)
  
  for (j0 in seq.int(1L, m, by = gene_block)) {
    j1 <- min(m, j0 + gene_block - 1L)
    
    # predicted block: n_cells x block_genes
    pred_blk <- tcrossprod(L_pm, F_use[j0:j1, , drop = FALSE])
    resid <- Y_all[, j0:j1, drop = FALSE] - pred_blk
    
    if (na_rm) {
      ok <- !is.na(resid)
      resid[!ok] <- 0
      sse  <- colSums(resid * resid)
      nobs <- colSums(ok)
    } else {
      sse  <- colSums(resid * resid)
      nobs <- rep.int(nrow(resid), ncol(resid))
    }
    
    rmse[j0:j1] <- sqrt(sse / pmax(nobs, 1))
    rm(pred_blk, resid); gc()
  }
  
  if (!is.null(gene_names)) names(rmse) <- gene_names
  rmse
}

rmse_vec <- compute_gene_rMSE_all(true_expression, L_pm, F_pm, gene_block = 300)
summary(rmse_vec)
```


```{r}
gene_summary <- readRDS(paste0(data_path, "gene_factor_validation.rds"))
# gene_summary <- readRDS(paste0(data_path, "gene_factor_validation_sanity.rds"))
```


```{r}
gene_summary %>%
  # order by rel_mse from low to high
  arrange(rel_mse) %>%
  # only show top 40 genes
  head(40) %>%
  ggplot(aes(x = reorder(gene, -rel_mse), y = rel_mse)) +
  geom_bar(stat = "identity", fill = "lightseagreen") +
  theme_minimal() +
  labs(title = "Relative rMSE by gene",
       x = "gene",
       y = "Relative rMSE") +
  # make the x-lim 0 to 1
  coord_flip(ylim = c(0, 1)) +
  # make the text size smaller +
  theme(axis.text.y = element_text(size = 4.5))
``` 

Plot R2 by gene:

```{r}
gene_summary %>%
  # order by R2 from high to low
  arrange(desc(R2)) %>%
  # only show top 40 genes
  head(40) %>%
  ggplot(aes(x = reorder(gene, R2), y = R2)) +
  geom_bar(stat = "identity", fill = "lightcoral") +
  theme_minimal() +
  labs(title = "R² by gene",
       x = "gene",
       y = "R²") +
  # make the x-lim 0 to 1
  coord_flip(ylim = c(0, 1)) +
  # make the text size smaller +
  theme(axis.text.y = element_text(size = 4.5))
```


