---
title: Examining the "extreme values" in the semi-NMF with K=200
author: Peter Carbonetto
output: workflowr::wflow_html
---

There is a curious phenomenon in the $K = 200$ semi-NMF results which
produces lopsided, or "extreme", values in the factors. As a point of
comparison, the NMF results with $K = 20$ does not show this pattern.
Here we expose this issue and suggest an extra step to filter out a
small fraction of cells with these extreme values.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load the packages needed for this analysis:

```{r load-pkgs, message=FALSE}
library(Matrix)
library(fastTopics)
library(flashier)
library(ggplot2)
library(cowplot)
```

Load the results of the flashier NMF analysis with $K = 20$:

```{r load-nmf}
load("../outputs/fl_nmf_k=20.RData")
colnames(fl_nmf_ldf$L) <- paste0("k",1:20)
colnames(fl_nmf_ldf$F) <- paste0("k",1:20)
```

In the LDF decomposition of the NMF (with "infinity-norm scaling"),
this is the ${\bf D}$ matrix:

```{r nmf-D-histogram, fig.height=2, fig.width=2}
pdat <- data.frame(d = fl_nmf_ldf$D)
ggplot(pdat,aes(x = d)) +
  geom_histogram(bins = 12,color = "black",fill = "black") +
  labs(x = "") +
  theme_cowplot(font_size = 12)
```

This next histogram is simply a histogram of the "total memberships",
i.e., the row sums of the ${\bf L}$ matrix in the LDF decomposition:

```{r nmf-L-histogram, fig.height=2, fig.width=3}
pdat <- data.frame(x = rowSums(fl_nmf_ldf$L))
ggplot(pdat,aes(x = x)) +
  geom_histogram(bins = 32,color = "black",fill = "black") +
  labs(x = "") +
  theme_cowplot(font_size = 12)
```

Notice that the peak of the distribution is hovering around 1 which is
somewhat reassuring.

Let's now compare these histograms to what we get from the LDF
decomposition of the semi-NMF fit.

```{r load-snmf}
load("../data/flashier_snmf_ldf.RData")
```
The ${\bf D}$ matrix contains some very large values:

```{r snmf-D-histogram, fig.height=2, fig.width=2.5}
pdat <- data.frame(d = fl_snmf_ldf$D)
ggplot(pdat,aes(x = d)) +
  geom_histogram(bins = 24,color = "black",fill = "black") +
  labs(x = "") +
  theme_cowplot(font_size = 12)
```

And a small proportion of the cells have very large total memberships
(e.g., >10):

```{r snmf-L-histogram, fig.height=2, fig.width=3}
pdat <- data.frame(x = rowSums(fl_snmf_ldf$L))
ggplot(pdat,aes(x = x)) +
  geom_histogram(bins = 32,color = "black",fill = "black") +
  labs(x = "") +
  theme_cowplot(font_size = 12)
```

This skewed distribution leads to unappealing Structure plots such as
this:

```{r structure-plot-by-cell-type, fig.height=2.5, fig.width=7.25, results="hide", message=FALSE}
L <- fl_snmf_ldf$L
set.seed(1)
cell_type_factors <- c("k3","k7","k22","k27","k29","k30","k58","k68",
                       "k101","k127")
cell_type <- sample_info$annotation_level1
cells <- which(cell_type == "CD4" | cell_type == "CD8")
cells <- sample(cells,1e5)
cells <- sort(c(cells,which(cell_type != "CD4" & cell_type != "CD8")))
structure_plot(L[cells,cell_type_factors],gap = 10,n = 2000,
                    grouping = cell_type[cells]) +
  labs(y = "membership",color = "",fill = "") +
  theme(legend.position = "bottom",legend.direction = "horizontal")
```

For example factor 58 which is corresponds quitely closely to CD8+
cells is not well visualized in this Structure plot.

To address this issue, I suggest an iterative process in which (i) a
small proportion of cells are filtered out by their total memberships
then (ii) the LDF factorization is adjusted accordingly. This is my
procedure:

```{r filter-cells-by-total-membership-function}
filter_cells_by_total_membership <- function (L, max_val = 10, numiter = 10) {
  n <- nrow(L)
  rows <- 1:n
  for (iter in 1:numiter) {
    x <- rowSums(L)
	cat(sprintf("%d. Filtered out %d cells.\n",iter,sum(x > max_val)))
	i <- which(x <= max_val)
	L <- L[i,]
	rows <- rows[i]
	d <- apply(L,2,max)
	L <- scale_cols(L,1/d)
  }
  return(rows)
}
```

Here is the missing function `scale_cols`:

```{r scale-cols-function}
scale_cols <- function (A, b)
  t(t(A) * b)
```

Now let's apply this iterative filtering procedure to the ${\bf L}$
matrix from the semi-NMF fit:

```{r filter-cells}
L <- fl_snmf_ldf$L
D <- fl_snmf_ldf$D
cells <- filter_cells_by_total_membership(L,numiter = 12)
sample_info <- sample_info[cells,]
L <- L[cells,]
```

Notice that this iterative filtering eventually "converged"; that is,
no more cells were removed.

And now renormalize the memberships:

```{r renormalize-memberships}
d <- apply(L,2,max)
L <- scale_cols(L,1/d)
fl_snmf_ldf$L <- L
fl_snmf_ldf$D <- d * fl_snmf_ldf$D
```

With this filtering step, the new Structure plot is much more balanced:

```{r structure-plot-by-cell-type-improved, fig.height=2.5, fig.width=7.25, results="hide", message=FALSE}
set.seed(1)
cell_type <- sample_info$annotation_level1
cell_type_factors <- c("k127","k58","k3","k22","k29",
                       "k68","k7","k30","k101","k27")
cells <- which(cell_type == "CD4" | cell_type == "CD8")
cells <- sample(cells,1e5)
cells <- sort(c(cells,which(cell_type != "CD4" & cell_type != "CD8")))
structure_plot(L[cells,cell_type_factors],gap = 10,n = 2000,
                    grouping = cell_type[cells],
					topics = 1:length(cell_type_factors)) +
  labs(y = "membership",color = "",fill = "") +
  theme(legend.position = "bottom",legend.direction = "horizontal")
```

Likewise, the ${\bf D}$ matrix is now considerably more balanced than
before:

```{r snmf-D-histogram-new, fig.height=2, fig.width=2.5}
pdat <- data.frame(d = fl_snmf_ldf$D)
ggplot(pdat,aes(x = d)) +
  geom_histogram(bins = 24,color = "black",fill = "black") +
  labs(x = "") +
  theme_cowplot(font_size = 12)
```

And less surprisingly so are the "total memberships":

```{r snmf-L-histogram-new, fig.height=2, fig.width=3}
pdat <- data.frame(x = rowSums(fl_snmf_ldf$L))
ggplot(pdat,aes(x = x)) +
  geom_histogram(bins = 32,color = "black",fill = "black") +
  labs(x = "") +
  theme_cowplot(font_size = 12)
```
