---
title: "Figure 1 for the manuscript (version 1)"
author: "Ziang Zhang"
date: "2025-11-06"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r}
library(ggplot2)
library(fastTopics)
library(dplyr)
library(ggrepel)
data_path <- "../data/"
code_path <- "../code/"
data_path <- "data/"
code_path <- "code/"
source(paste0(code_path, "ROC.R"))
seurat_meta <- readRDS(paste0(data_path, "seurat_meta.rds"))
flashier_snmf_summary <- readRDS(paste0(data_path, "flashier_snmf_summary.rds"))
```


## Figure 1

Part (A): Scatterplot of maximum absolute coefficients for level1 versus organ.

Align the cells to remove the extra cells in the flashier result:

```{r}
cells_flashier <- rownames(flashier_snmf_summary$L_pm)
cells_seurat <- seurat_meta$cellID
which(!cells_flashier %in% cells_seurat)
L_pm <- flashier_snmf_summary$L_pm[cells_flashier %in% cells_seurat, ]
all(rownames(L_pm) == cells_seurat)
```

Focus on healthy cells that are not thymocytes:

```{r}
healthy_cells <- which(seurat_meta$condition_broad == "healthy")
L_pm <- L_pm[healthy_cells, ]
seurat_meta <- seurat_meta[healthy_cells, ]

cells_thymocyte <- which(seurat_meta$annotation_level1 == "thymocyte")
L_pm <- L_pm[-cells_thymocyte, ]
seurat_meta <- seurat_meta[-cells_thymocyte, ]
```

Filter out organs and cell types with too few cells:

```{r}
selected_organ <- table(seurat_meta$organ)
selected_organ <- names(selected_organ[selected_organ >= 3000])
selected_level1 <- table(seurat_meta$annotation_level1)
selected_level1 <- names(selected_level1[selected_level1 >= 1000])
selected_cells <- which(seurat_meta$organ %in% selected_organ & seurat_meta$annotation_level1 %in% selected_level1)
seurat_meta_selected <- seurat_meta[selected_cells, ]
# redefine the levels
seurat_meta_selected$organ <- factor(seurat_meta_selected$organ)
seurat_meta_selected$annotation_level1 <- factor(seurat_meta_selected$annotation_level1)
```

Write a function that computes the coefficients of organ and cell type for each factor:

```{r}
compute_coef_per_factor <- function(factor_id){
  Y <- L_pm[selected_cells, factor_id]
  organ <- seurat_meta_selected$organ
  level1 <- seurat_meta_selected$annotation_level1
  model_full <- lm(Y ~ organ + level1)
  coefs <- coef(model_full)
  # Extract coefficients for organ
  organ_coefs <- coefs[grep("^organ", names(coefs))]
  # Extract coefficients for annotation_level1
  level1_coefs <- coefs[grep("^level1", names(coefs))]
  
  return(
    list(
      factor_id = factor_id,
      organ_coefs = organ_coefs,
      level1_coefs = level1_coefs
    )
  )
}
```

```{r eval=FALSE}
factor_coef_list <- lapply(1:ncol(L_pm), compute_coef_per_factor)
saveRDS(factor_coef_list, file = paste0(data_path, "conditional_analysis_factor_level1_organ.rds"))
```


Plot the maximum absolute coefficient for each factor:

```{r}
factor_coef_list <- readRDS(file = paste0(data_path, "conditional_analysis_factor_level1_organ.rds"))

max_coef_df <- data.frame(
  Factor = integer(),
  Max_Abs_Coefficient_Organ = numeric(),
  Max_Abs_Coefficient_Level1 = numeric()
)

for (i in 1:length(factor_coef_list)) {
  coef_df <- factor_coef_list[[i]]
  max_abs_organ <- max(abs(coef_df$organ_coefs))
  max_abs_Level1 <- max(abs(coef_df$level1_coefs))
  max_coef_df <- rbind(max_coef_df, data.frame(
    Factor = i,
    Max_Abs_Coefficient_Organ = max_abs_organ,
    Max_Abs_Coefficient_Level1 = max_abs_Level1
  ))
}

```


```{r}
top_n <- 10  # how many points to annotate above and below y = x

df <- max_coef_df %>%
  mutate(
    residual = Max_Abs_Coefficient_Level1 - Max_Abs_Coefficient_Organ, # >0: above y=x; <0: below
    abs_res  = abs(residual)
  )

# --- Select top-N above and below y = x ---
label_above <- df %>%
  filter(is.finite(residual), residual > 0) %>%
  slice_max(residual, n = top_n, with_ties = FALSE)

label_below <- df %>%
  filter(is.finite(residual), residual < 0) %>%
  slice_min(residual, n = top_n, with_ties = FALSE)

# --- Base plot builder (no labels yet) ---
base_plot <- ggplot(df, aes(x = Max_Abs_Coefficient_Organ, y = Max_Abs_Coefficient_Level1)) +
  geom_point(alpha = 0.3, size = 1.8) +                                # background points
  geom_point(data = label_above, color = "#1f78b4", alpha = 0.6, size = 1.8) +       # highlight above y=x
  geom_point(data = label_below, color = "#e31a1c", alpha = 0.6, size = 1.8) +       # highlight below y=x
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  # # make both y and x axis on log scale
  # scale_x_log10() +
  # scale_y_log10() +
  coord_cartesian(xlim = c(0, 0.4), ylim = c(0, 0.4), expand = TRUE) +  # keep your limits
  labs(
    x = "Max Absolute Coefficient (Organ)",
    y = "Max Absolute Coefficient (Level-1)",
    title = "Max Absolute Coefficient: Organ vs Level-1"
  ) +
  theme_minimal()

# --- Combined plot: annotate both top-N above and below y = x ---
p_all <- base_plot +
  geom_text_repel(
    data = label_above,
    aes(label = paste0("Factor ", Factor)),
    color = "#1f78b4", size = 3, max.overlaps = Inf,
    box.padding = 0.3, point.padding = 0.2, min.segment.length = 0
  ) +
  geom_text_repel(
    data = label_below,
    aes(label = paste0("Factor ", Factor)),
    color = "#e31a1c", size = 3, max.overlaps = Inf,
    box.padding = 0.3, point.padding = 0.2, min.segment.length = 0
  ) +
  labs(subtitle = paste0("Annotating top ", top_n, " above (blue) and top ", top_n, " below (red) y = x"))

p_all
```


Also do a similar scatterplot for max AUC:

```{r}
level_1_AUC <- readRDS(paste0(data_path, "level_1_AUC.rds"))
level_1_AUC_max <- apply(level_1_AUC, 2, max)
level_1_AUC_max_name <- apply(level_1_AUC, 2, function(x) rownames(level_1_AUC)[which.max(x)])
o <- order(level_1_AUC_max, decreasing = TRUE)
table_level_1_AUC <- data.frame(
  Factor = colnames(level_1_AUC)[o],
  Max_AUC = level_1_AUC_max[o],
  Annotation = level_1_AUC_max_name[o]
)

organ_AUC <- readRDS(paste0(data_path, "organ_AUC.rds"))
# remove CNS and synovial fluid from organ_AUC
organ_AUC <- organ_AUC[!rownames(organ_AUC) %in% c("CNS","synovial fluid"), ]
organ_AUC_max <- apply(organ_AUC, 2, max)
organ_AUC_max_name <- apply(organ_AUC, 2, function(x) rownames(organ_AUC)[which.max(x)])
o <- order(organ_AUC_max, decreasing = TRUE)
table_organ_AUC <- data.frame(
  Max_AUC = organ_AUC_max[o],
  Annotation = organ_AUC_max_name[o]
)

# merge the two AUC tables by factor
max_AUC_df <- data.frame(
  Factor = table_level_1_AUC$Factor,
  annotation_Level1 = table_level_1_AUC$Annotation,
  annotation_Organ = organ_AUC_max_name[match(table_level_1_AUC$Factor, names(organ_AUC_max))],
  Max_AUC_Organ = organ_AUC_max[match(table_level_1_AUC$Factor, names(organ_AUC_max))],
  Max_AUC_Level1 = table_level_1_AUC$Max_AUC
)
```

Scatterplot of max AUC for organ vs level1:

```{r}
top_n <- 8

df <- max_AUC_df %>%
  mutate(
    residual = Max_AUC_Level1 - Max_AUC_Organ,
    abs_res  = abs(residual)
  )

df$annotation_Level1 <- gsub("all_CD8aa", "CD8aa", df$annotation_Level1)

label_above <- df %>%
  filter(is.finite(residual), residual > 0) %>%
  slice_max(Max_AUC_Level1, n = top_n, with_ties = FALSE) %>%
  mutate(nudge_x = -0.035)

label_below <- df %>%
  filter(is.finite(residual), residual < 0) %>%
  slice_min(-Max_AUC_Organ, n = top_n, with_ties = FALSE) %>%
  mutate(nudge_x = 0.035)

base_plot <- ggplot(df, aes(Max_AUC_Organ, Max_AUC_Level1)) +
  geom_point(alpha = 0.3, size = 1.8) +
  geom_point(data = label_above, color = "#1f78b4", alpha = 0.8, size = 1.8) +
  geom_point(data = label_below, color = "#e31a1c", alpha = 0.8, size = 1.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  coord_cartesian(xlim = c(0.46, 1.04), ylim = c(0.5, 1.02), expand = FALSE) +
  labs(
    x = "Max AUC (Organ)",
    y = "Max AUC (Level-1)",
    title = "Max AUC: Organ vs Level-1",
    subtitle = sprintf("Annotating top %d above (blue) and top %d below (red) y = x", top_n, top_n)
  ) +
  theme_minimal(base_size = 13)

p_all <- base_plot +
  geom_text_repel(
    data = label_above,
    aes(label = paste0(Factor, ": ", annotation_Level1)),
    color = "#1f78b4",
    direction = "y",
    nudge_x = label_above$nudge_x,
    segment.color = "#1f78b4",
    arrow = arrow(length = unit(0.008, "npc"), type = "closed", angle = 20),  # smaller arrow
    force = 1.2, force_pull = 0.2,
    box.padding = 0.35, point.padding = 0.15,
    max.time = 2, max.iter = 5e3,
    min.segment.length = 0.01,
    segment.alpha = 0.7
  ) +
  geom_text_repel(
    data = label_below,
    aes(label = paste0(Factor, ": ", annotation_Organ)),
    color = "#e31a1c",
    direction = "y",
    nudge_x = label_below$nudge_x,
    segment.color = "#e31a1c",
    arrow = arrow(length = unit(0.008, "npc"), type = "closed", angle = 20),  # smaller arrow
    force = 1.2, force_pull = 0.2,
    box.padding = 0.35, point.padding = 0.15,
    max.time = 2, max.iter = 5e3,
    min.segment.length = 0.01,
    segment.alpha = 0.7
  )

p_all

ggsave(
  filename = "figures/Figure1/Figure1_MaxAUC_Organ_vs_Level1_arrow_small.png",
  plot = p_all, width = 8, height = 8, dpi = 300
)
```

Let's prepare a structure plot for the selected factors"

```{r}
### Filter cells with high total membership
filter_cells_by_total_membership <- function (L, max_val = 10, numiter = 10) {
  n <- nrow(L)
  rows <- 1:n
  for (iter in 1:numiter) {
    x <- rowSums(L)
    cat(sprintf("%d. Filtered out %d cells.\n",iter,sum(x > max_val)))
    i <- which(x <= max_val)
    L <- L[i,]
    rows <- rows[i]
    d <- apply(L,2,max)
    L <- scale_cols(L,1/d)
  }
  return(rows)
}
scale_cols <- function (A, b) t(t(A) * b)
D <- diag(1 / apply(L_pm, 2, function(x) max(x)))
L <- L_pm %*% D
cells <- filter_cells_by_total_membership(L,numiter = 12)
seurat_meta_filtered <- seurat_meta[cells,]
L_pm_filtered <- L_pm[cells,]

### Normalizing loadings
d <- apply(L_pm_filtered,2,max)
L_pm_filtered <- scale_cols(L_pm_filtered,1/d)
D <- (1/d) * D
```

```{r}
### Highlighted cell types
set.seed(12345)
### These are all factors that passed the validation test.
highlights <- c("K22","K30","K58","K68")
color_coding <- ZemmourLib::immgent_colors$level1
Level1_match <- c("DN", "nonconv", "CD8", "Treg")
cell_type_factors <- as.numeric(gsub("K", highlights, replacement = ""))
fit2 <- L_pm_filtered[, cell_type_factors, drop = FALSE]
colnames(fit2) <- paste0("k", cell_type_factors)

cell_type <- seurat_meta_filtered$annotation_level1
# cells <- which(cell_type == "CD4" | cell_type == "CD8" | seurat_meta_filtered$organ_simplified == "spleen")
cells <- which(cell_type == "CD4" | cell_type == "CD8")
cells <- sample(cells,1e5)
cells <- sort(c(cells,which(cell_type != "CD4" & cell_type != "CD8")))

structure_plot(fit2[cells,],gap = 20,n = 3000,
               colors = color_coding[Level1_match],
               grouping = cell_type[cells]) +
  labs(y = "membership",color = "",fill = "") +
  guides(
    fill = guide_legend(nrow = 1),
    color = guide_legend(nrow = 1)
  ) +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.box = "horizontal",
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13)
  )

ggsave(filename = "figures/Figure1/Figure1_StructurePlot_SelectedFactors.png",
       width = 8, height = 5)
```

Showing some boxplot to highlight the loading differences:

```{r}
plot_factor_boxplot_level1 <- function(factor_id, 
                                L_pm_filtered, 
                                seurat_meta_filtered,
                                outline = FALSE,
                                log = FALSE,
                                color_coding) {
  df <- data.frame(
    Loading = L_pm_filtered[, factor_id],
    CellType = seurat_meta_filtered$annotation_level1
  )
  
  median_order <- df %>%
    group_by(CellType) %>%
    summarize(med = median(Loading, na.rm = TRUE)) %>%
    arrange(desc(med)) %>%        
    pull(CellType)
  
  df$CellType <- factor(df$CellType, levels = median_order)
  
  p <- ggplot(df, aes(x = CellType, y = Loading, fill = CellType)) +
    geom_boxplot(outliers = outline, outlier.size = 0.4, width = 0.6, alpha = 0.8, color = "gray40") +
    scale_fill_manual(values = color_coding) +
    # scale_y_log10() +
    labs(
      title = "",
      x = "",
      y = paste0("Loading K", factor_id)
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
      axis.text.x = element_text(size = 10, angle = 45, hjust = 1, vjust = 1),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 14, face = "bold"),
      legend.position = "none",
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank()
    )
  
  if (log) {
    p <- p + scale_y_log10()
  }
  
  return(p)
}

p1 <- plot_factor_boxplot_level1(factor_id = 68, L_pm_filtered = L_pm_filtered, seurat_meta_filtered = seurat_meta_filtered, color_coding = color_coding)
p2 <- plot_factor_boxplot_level1(factor_id = 58, L_pm_filtered = L_pm_filtered, seurat_meta_filtered = seurat_meta_filtered, color_coding = color_coding)
p3 <- plot_factor_boxplot_level1(factor_id = 30, L_pm_filtered = L_pm_filtered, seurat_meta_filtered = seurat_meta_filtered, color_coding = color_coding)
p4 <- plot_factor_boxplot_level1(factor_id = 22, L_pm_filtered = L_pm_filtered, seurat_meta_filtered = seurat_meta_filtered, color_coding = color_coding)
p_combined <- gridExtra::grid.arrange(p1, p2, p3, p4, ncol = 2)
ggsave(filename = "figures/Figure1/Figure1_Boxplot_SelectedFactors.png",
       plot = p_combined,
       width = 8, height = 8)
```


Showing an UMAP plot to highlight the loading differences:

```{r}
umap_result <- readRDS(paste0(data_path, "umap_result.rds"))
colnames(umap_result) <- c("UMAP_1", "UMAP_2")
umap_result <- umap_result[rownames(L_pm_filtered),]
```

```{r}
df_umap <- as.data.frame(umap_result)
df_umap$CellType <- seurat_meta_filtered$annotation_level1
stopifnot(nrow(df_umap) == length(seurat_meta_filtered$annotation_level1))

highlighted_type <- "nonconv"

set.seed(1)
bg_n <- 120000
bg_idx <- if (nrow(df_umap) > bg_n) sample(seq_len(nrow(df_umap)), bg_n) else seq_len(nrow(df_umap))
bg_df <- df_umap[bg_idx, ]

type_df <- df_umap[df_umap$CellType == highlighted_type, , drop = FALSE]

p_type <- ggplot(bg_df, aes(UMAP_1, UMAP_2)) +
  geom_point(color = "grey85", size = 0.1, alpha = 0.05) +              
  geom_point(data = type_df,
             color = unname(color_coding[highlighted_type]),                   
             size = 0.25, alpha = 0.7) +                                
  coord_equal() +
  labs(title = paste0(highlighted_type, " on UMAP"), x = "UMAP 1", y = "UMAP 2") +
  theme_minimal() +
  theme(panel.grid = element_blank())

p_type
```

```{r}
plot_loadings_on_umap <- function(umap, loading, factor_num, base_color, size = 0.5) {
  df <- as.data.frame(umap)
  stopifnot(nrow(df) == length(loading))
  df$loading <- loading
  
  ggplot(df, aes(x = UMAP_1, y = UMAP_2)) +
    geom_point(
      aes(alpha = loading),
      color = base_color,
      size  = size,
      show.legend = FALSE
    ) +
    scale_alpha(range = c(0, 0.1)) +
    labs(
      title = paste0("Factor ", factor_num),
      x = "UMAP 1", y = "UMAP 2"
    ) +
    theme_minimal() +
    theme(
      panel.background = element_rect(fill = "white", color = NA),
      panel.grid = element_blank()
    )
}

factor_color_map <- c(
  "22" = unname(color_coding["CD8"]),
  "30" = unname(color_coding["nonconv"]),
  "58" = unname(color_coding["CD4"]),
  "68" = unname(color_coding["Treg"])
)

selected_factors <- c(22, 30, 58, 68) 
plots <- list()
sample_n <- 80000

for (f in selected_factors) {
  sample_index <- sample(
    seq_len(nrow(L_pm_filtered)),
    size    = min(sample_n, nrow(L_pm_filtered)),
    replace = FALSE
  )
  
  loading <- L_pm_filtered[sample_index, f]
  base_col <- unname(factor_color_map[as.character(f)])
  
  p <- plot_loadings_on_umap(
    umap       = umap_result[sample_index, ],
    loading    = loading,
    factor_num = f,
    base_color = base_col,
    size       = 0.1
  )
  
  plots[[length(plots) + 1]] <- p
}

combined_plot <- patchwork::wrap_plots(plots, ncol = 2, nrow = 2)

ggsave(
  filename = "figures/Figure1/Figure1_UMAP_SelectedFactors.png",
  plot     = combined_plot,
  width    = 8,
  height   = 8,
  dpi      = 300
)
```


```{r}
plot_loadings_on_umap_v2 <- function(umap, loading, factor_num,
                                     size = 0.25,
                                     show_bg = TRUE,
                                     bg_alpha = 0.05,
                                     bg_color = "grey90",
                                     clip_q = 0.99,         
                                     top_q = NULL,          
                                     viridis_option = "viridis",
                                     viridis_direction = 1) {
  stopifnot(is.matrix(umap) || is.data.frame(umap))
  df <- as.data.frame(umap)
  stopifnot(all(c("UMAP_1","UMAP_2") %in% colnames(df)))
  stopifnot(nrow(df) == length(loading))

  q_hi <- stats::quantile(loading, probs = clip_q, na.rm = TRUE)
  loading_clip <- pmin(loading, q_hi)

  if (!is.null(top_q)) {
    stopifnot(top_q > 0 && top_q < 1)
    thr <- stats::quantile(loading_clip, probs = top_q, na.rm = TRUE)
    loading_plot <- ifelse(loading_clip >= thr, loading_clip, NA_real_)
  } else {
    loading_plot <- loading_clip
  }

  df$loading_plot <- loading_plot

  p <- ggplot2::ggplot() +
    {
      if (show_bg) ggplot2::geom_point(
        data = df, ggplot2::aes(UMAP_1, UMAP_2),
        color = bg_color, size = size, alpha = bg_alpha
      )
    } +
    ggplot2::geom_point(
      data = df, ggplot2::aes(UMAP_1, UMAP_2, color = loading_plot),
      size = size, alpha = 0.9, na.rm = TRUE   
    ) +
    ggplot2::scale_color_viridis_c(
      option = viridis_option,
      direction = viridis_direction,
      limits = c(0, q_hi),
      oob = scales::squish,
      breaks = scales::pretty_breaks(4),
      name = paste0("K", factor_num, " loading")
    ) +
    ggplot2::coord_equal() +
    ggplot2::labs(
      title = paste0("Factor ", factor_num),
      x = "", y = ""
    ) +
    ggplot2::theme_minimal() +
    ggplot2::theme(
      panel.grid = ggplot2::element_blank()
    )

  return(p)
}

selected_factors <- c(22, 30, 58, 68)
sample_n <- 80000

set.seed(1)
sample_index <- sample(seq_len(nrow(L_pm_filtered)), size = min(sample_n, nrow(L_pm_filtered)))
umap_sub <- umap_result[sample_index, , drop = FALSE]
L_sub    <- L_pm_filtered[sample_index, selected_factors, drop = FALSE]

plots <- lapply(seq_along(selected_factors), function(i){
  f <- selected_factors[i]
  plot_loadings_on_umap_v2(
    umap    = umap_sub,
    loading = L_sub[, i],
    factor_num = f,
    size = 0.1,
    show_bg = TRUE,
    bg_alpha = 0.06,
    clip_q = 0.99,
    top_q = NULL,             
    viridis_option = "viridis",
    viridis_direction = 1
  )
})

combined <- patchwork::wrap_plots(plots, ncol = 2)
ggsave(
  filename = "figures/Figure1/Figure1_UMAP_SelectedFactors_v2.png",
  plot     = combined,
  width    = 8,
  height   = 8,
  dpi      = 300
)
```







Protein annotation plot:

```{r}
protein_flash_summary_lognorm <- readRDS(file = paste0(data_path, "protein_flash_selected_summary_lognorm.rds"))

D_lognorm <- diag(1 / apply(protein_flash_summary_lognorm$F_pm, 2, function(x) (max(x) - min(x))))
protein_flash_summary_lognorm$F_pm <- protein_flash_summary_lognorm$F_pm %*% D_lognorm
protein_flash_summary_lognorm$L_pm <- protein_flash_summary_lognorm$L_pm %*% solve(D_lognorm)
F_pm <- protein_flash_summary_lognorm$F_pm
colnames(F_pm) <- paste0("F", 1:ncol(flashier_snmf_summary$F_pm))[ flashier_snmf_summary$pve > 1e-4]
```

```{r}
kset <- paste0("F", c(68, 58, 30, 22))
F_pm <- F_pm[, kset]
annotation_heatmap(F_pm,n = 3,
                   dim = kset,
                   show_dims = kset,
                   select_features = "both",
                   feature_sign = "both",
                   zero_value = 0.1,
                   compare_dims = kset,
                   font_size = 15) +
  theme(plot.title = element_text(face = "plain",size = 10))
ggsave(filename = "figures/Figure1/Figure1_ProteinAnnotationPlot_SelectedFactors.png",
       width = 6, height = 7)
```





