---
title: "Figures and tables for the manuscript"
author: "Ziang Zhang"
date: "2025-11-06"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r}
library(ggplot2)
library(fastTopics)
library(dplyr)
library(ggrepel)
library(tidyr)
library(tibble)
library(cowplot)
library(ZemmourLib)
library(patchwork)

data_path <- "../data/"
code_path <- "../code/"
table_path <- "../tables/"
data_path <- "data/"
code_path <- "code/"
table_path <- "tables/"
source(paste0(code_path, "ROC.R"))
seurat_meta <- readRDS(paste0(data_path, "seurat_meta.rds"))
flashier_snmf_summary <- readRDS(paste0(data_path, "flashier_snmf_summary.rds"))
```

```{r}
filter_cells_by_total_membership <- function (L, max_val = 10, numiter = 10) {
  n <- nrow(L)
  rows <- 1:n
  for (iter in 1:numiter) {
    x <- rowSums(L)
    cat(sprintf("%d. Filtered out %d cells.\n",iter,sum(x > max_val)))
    i <- which(x <= max_val)
    L <- L[i,]
    rows <- rows[i]
    d <- apply(L,2,max)
    L <- scale_cols(L,1/d)
  }
  return(rows)
}
scale_cols <- function (A, b)
  t(t(A) * b)
```

Align the cells to remove the extra cells in the flashier result:

```{r}
cells_flashier <- rownames(flashier_snmf_summary$L_pm)
cells_seurat <- seurat_meta$cellID
L_pm <- flashier_snmf_summary$L_pm[cells_flashier %in% cells_seurat, ]
colnames(L_pm) <- paste0("K", 1:ncol(L_pm))
all(rownames(L_pm) == cells_seurat)
```


## Table: overview of GPs

In this table, we want to show the maxAUC for each GP for level 1, level 2, organ, condition_detailed.

When computing max AUC, we should remove thymocyte cells.
The max of AUC should only be taken to categories that have at least 500 cells.

```{r}
cells_thymocyte <- which(seurat_meta$annotation_level1 == "thymocyte")
L_pm_no_thymocytes <- L_pm[-cells_thymocyte, ]
seurat_meta_no_thymocytes <- seurat_meta[-cells_thymocyte, ]
```

```{r eval=FALSE}
level_1_AUC_list <- compute_auc_threshold_matrix(loading_mat = L_pm_no_thymocytes, group_info = seurat_meta_no_thymocytes$annotation_level1)
rownames(level_1_AUC_list$auc) <- sort(unique(seurat_meta_no_thymocytes$annotation_level1))
rownames(level_1_AUC_list$threshold) <- sort(unique(seurat_meta_no_thymocytes$annotation_level1))
saveRDS(level_1_AUC_list, file = paste0(data_path, "level_1_AUC_list_figure.rds"))
```

```{r}
level_1_AUC_list <- readRDS(paste0(data_path, "level_1_AUC_list_figure.rds"))
level_1_AUC <- level_1_AUC_list$auc
level_1_threshold <- level_1_AUC_list$threshold
```

```{r}
level_1_AUC_fac_row <- t(level_1_AUC)

level_1_AUC_fac_row_df <- as.data.frame(level_1_AUC_fac_row) |>
  rownames_to_column(var = "Factor")  

write.csv(level_1_AUC_fac_row_df, paste0(table_path,"level1_AUC_factor_rows.csv"), row.names = FALSE)
```

```{r eval=FALSE}
level_2_AUC_list <- compute_auc_threshold_matrix(loading_mat = L_pm_no_thymocytes, group_info = seurat_meta_no_thymocytes$annotation_level2)
rownames(level_2_AUC_list$auc) <- sort(unique(seurat_meta_no_thymocytes$annotation_level2))
rownames(level_2_AUC_list$threshold) <- sort(unique(seurat_meta_no_thymocytes$annotation_level2))
saveRDS(level_2_AUC_list, file = paste0(data_path, "level_2_AUC_list_figure.rds"))
```

```{r}
level_2_AUC_list <- readRDS(paste0(data_path, "level_2_AUC_list_figure.rds"))
level_2_AUC <- level_2_AUC_list$auc
level_2_threshold <- level_2_AUC_list$threshold

# filter out annotations with less than 500 cells
which_less_500 <- sapply(rownames(level_2_AUC), function(x){
  sum(seurat_meta$annotation_level2 == x) < 500
})
level_2_AUC <- level_2_AUC[!which_less_500, ]
level_2_threshold <- level_2_threshold[!which_less_500, ]
```

```{r}
level_2_AUC_fac_row <- t(level_2_AUC)

level_2_AUC_fac_row_df <- as.data.frame(level_2_AUC_fac_row) |>
  rownames_to_column(var = "Factor")  

write.csv(level_2_AUC_fac_row_df, paste0(table_path,"level_2_AUC_factor_rows.csv"), row.names = FALSE)
```

```{r}
level_2_within_level_1_AUC_list <- readRDS(paste0(data_path, "level_2_within_level_1_AUC_list_figure.rds"))
level_2_AUC_within_level1 <- level_2_within_level_1_AUC_list$auc
level_2_threshold_within_level1 <- level_2_within_level_1_AUC_list$threshold

# filter out annotations with less than 500 cells
which_less_500 <- sapply(rownames(level_2_AUC_within_level1), function(x){
  sum(seurat_meta$annotation_level2 == x) < 500
})
level_2_AUC_within_level1 <- level_2_AUC_within_level1[!which_less_500, ]
level_2_threshold_within_level1 <- level_2_threshold_within_level1[!which_less_500, ]
```

```{r}
level_2_AUC_within_level1_fac_row <- t(level_2_AUC_within_level1)

level_2_AUC_within_level1_fac_row <- as.data.frame(level_2_AUC_within_level1_fac_row) |>
  rownames_to_column(var = "Factor")  

write.csv(level_2_AUC_within_level1_fac_row, paste0(table_path,"level_2_AUC_within_level1_factor_rows.csv"), row.names = FALSE)
```





```{r eval=FALSE}
condition_detailed_AUC_list <- compute_auc_threshold_matrix(loading_mat = L_pm_no_thymocytes, group_info = seurat_meta_no_thymocytes$condition_detailed)
rownames(condition_detailed_AUC_list$auc) <- sort(unique(seurat_meta_no_thymocytes$condition_detailed))
rownames(condition_detailed_AUC_list$threshold) <- sort(unique(seurat_meta_no_thymocytes$condition_detailed))
saveRDS(condition_detailed_AUC_list, file = paste0(data_path, "condition_detailed_AUC_list_figure.rds"))
```

```{r}
condition_detailed_AUC_list <- readRDS(paste0(data_path, "condition_detailed_AUC_list_figure.rds"))
condition_detailed_AUC <- condition_detailed_AUC_list$auc
condition_detailed_threshold <- condition_detailed_AUC_list$threshold

# filter out annotations with less than 500 cells
which_less_500 <- sapply(rownames(condition_detailed_AUC), function(x){
  sum(seurat_meta$condition_detailed == x) < 500
})
condition_detailed_AUC <- condition_detailed_AUC[!which_less_500, ]
condition_detailed_threshold <- condition_detailed_threshold[!which_less_500, ]

condition_detailed_AUC_fac_row <- t(condition_detailed_AUC)
condition_detailed_AUC_fac_row_df <- as.data.frame(condition_detailed_AUC_fac_row) |>
  rownames_to_column(var = "Factor")  
write.csv(condition_detailed_AUC_fac_row_df, paste0(table_path,"condition_detailed_AUC_factor_rows.csv"), row.names = FALSE)
```

```{r eval=FALSE}
organ_AUC_list <- compute_auc_threshold_matrix(loading_mat = L_pm_no_thymocytes, group_info = seurat_meta_no_thymocytes$organ)
rownames(organ_AUC_list$auc) <- sort(unique(seurat_meta_no_thymocytes$organ))
rownames(organ_AUC_list$threshold) <- sort(unique(seurat_meta_no_thymocytes$organ))
saveRDS(organ_AUC_list, file = paste0(data_path, "organ_AUC_list_figure.rds"))
```

```{r}
organ_AUC_list <- readRDS(paste0(data_path, "organ_AUC_list_figure.rds"))
organ_AUC <- organ_AUC_list$auc
organ_threshold <- organ_AUC_list$threshold

# filter out annotations with less than 500 cells
which_less_500 <- sapply(rownames(organ_AUC), function(x){
  sum(seurat_meta$organ == x) < 500
})
organ_AUC <- organ_AUC[!which_less_500, ]
organ_threshold <- organ_threshold[!which_less_500, ]

organ_AUC_fac_row <- t(organ_AUC)
organ_AUC_fac_row_df <- as.data.frame(organ_AUC_fac_row) |>
  rownames_to_column(var = "Factor")
write.csv(organ_AUC_fac_row_df, paste0(table_path,"organ_AUC_factor_rows.csv"), row.names = FALSE)
```


```{r eval=FALSE}
igt_AUC_list <- compute_auc_threshold_matrix(loading_mat = L_pm_no_thymocytes, group_info = seurat_meta_no_thymocytes$IGT)
rownames(igt_AUC_list$auc) <- sort(unique(seurat_meta_no_thymocytes$IGT))
rownames(igt_AUC_list$threshold) <- sort(unique(seurat_meta_no_thymocytes$IGT))
saveRDS(igt_AUC_list, file = paste0(data_path, "igt_AUC_list_figure.rds"))
```

```{r}
igt_AUC_list <- readRDS(paste0(data_path, "igt_AUC_list_figure.rds"))
igt_AUC <- igt_AUC_list$auc
igt_threshold <- igt_AUC_list$threshold
# filter out annotations with less than 500 cells
which_less_500 <- sapply(rownames(igt_AUC), function(x){
  sum(seurat_meta$IGT == x) < 500
})
igt_AUC <- igt_AUC[!which_less_500, ]
igt_threshold <- igt_threshold[!which_less_500, ]
igt_AUC_fac_row <- t(igt_AUC)
igt_AUC_fac_row_df <- as.data.frame(igt_AUC_fac_row) |>
  rownames_to_column(var = "Factor")
write.csv(igt_AUC_fac_row_df, paste0(table_path,"igt_AUC_factor_rows.csv"), row.names = FALSE)
```



## Figure: Heatmaps for each GP

First, look for variation across level 1 annotations:

Focus on GPs with max AUC > 0.65:

```{r}
## 1. select high-AUC factors by name
max_AUC_level1 <- apply(level_1_AUC, 2, max)
factors_high_AUC_level1 <- which(max_AUC_level1 > 0.65)
factor_names_high <- colnames(level_1_AUC)[factors_high_AUC_level1]

colnames(L_pm_no_thymocytes) <- paste0("K", 1:ncol(L_pm_no_thymocytes))
L_pm_level1_high_AUC <- L_pm_no_thymocytes[, factor_names_high, drop = FALSE]
annotations_level1 <- seurat_meta$annotation_level1[-cells_thymocyte]

# Take out cells for visualization
d <- apply(L_pm_no_thymocytes, 2, max)
L_pm_no_thymocytes_normalized <- scale_cols(L_pm_no_thymocytes, 1/d)
cells_all <- filter_cells_by_total_membership(L_pm_no_thymocytes_normalized, numiter = 15)
L_pm_no_thymocytes_normalized <- L_pm_no_thymocytes_normalized[cells_all, ]
d <- apply(L_pm_no_thymocytes_normalized, 2, max)
L_pm_no_thymocytes_normalized <- scale_cols(L_pm_no_thymocytes_normalized, 1/d)
L_pm_no_thymocytes_normalized_high_AUC <- L_pm_no_thymocytes_normalized[, factor_names_high, drop = FALSE]
```

```{r}
compute_prop_above_threshold <- function(
  loading_mat,
  group_info,
  threshold_mat,
  remove_na = TRUE,
  compare = c(">", ">=", "<", "<=")
) {
  compare <- match.arg(compare)

  # Coerce to matrix
  loading_mat <- as.matrix(loading_mat)

  # Basic checks
  if (nrow(loading_mat) != length(group_info)) {
    stop("`loading_mat` rows must have same length as `group_info`.")
  }
  if (is.null(rownames(threshold_mat)) || is.null(colnames(threshold_mat))) {
    stop("`threshold_mat` must have both rownames (groups) and colnames (loadings).")
  }
  if (is.null(colnames(loading_mat))) {
    stop("`loading_mat` must have column names that match threshold_mat's colnames.")
  }

  # Align loadings with threshold columns
  common_loadings <- intersect(colnames(loading_mat), colnames(threshold_mat))
  if (length(common_loadings) == 0L) {
    stop("No overlapping loading names between `loading_mat` and `threshold_mat`.")
  }
  loading_mat   <- loading_mat[, common_loadings, drop = FALSE]
  threshold_mat <- threshold_mat[, common_loadings, drop = FALSE]

  # Groups to use: intersection of group_info and threshold rownames
  all_groups      <- unique(as.character(group_info))
  threshold_groups <- rownames(threshold_mat)
  use_groups      <- intersect(threshold_groups, all_groups)
  if (length(use_groups) == 0L) {
    stop("No overlapping groups between `group_info` and `rownames(threshold_mat)`.")
  }

  # Subset threshold_mat to the groups we actually have
  threshold_mat <- threshold_mat[use_groups, , drop = FALSE]

  # Prepare output
  prop_mat <- matrix(
    NA_real_,
    nrow = nrow(threshold_mat),
    ncol = ncol(threshold_mat),
    dimnames = dimnames(threshold_mat)
  )

  # Helper for comparison
  cmp_fun <- switch(
    compare,
    ">"  = function(x, thr) x >  thr,
    ">=" = function(x, thr) x >= thr,
    "<"  = function(x, thr) x <  thr,
    "<=" = function(x, thr) x <= thr
  )

  # Loop over groups and loadings
  for (g in use_groups) {
    idx_g <- which(as.character(group_info) == g)

    if (length(idx_g) == 0L) next

    for (k in common_loadings) {
      x <- loading_mat[idx_g, k]

      if (remove_na) {
        keep <- !is.na(x)
        x    <- x[keep]
      }

      if (length(x) == 0L || is.na(threshold_mat[g, k])) {
        prop_mat[g, k] <- NA_real_
      } else {
        thr <- threshold_mat[g, k]
        prop_mat[g, k] <- mean(cmp_fun(x, thr))
      }
    }
  }

  prop_mat
}
```


```{r}
## 1. compute proportion of cells above threshold for each (level1, factor)
prop_mat <- compute_prop_above_threshold(
  loading_mat  = L_pm_level1_high_AUC,
  group_info   = seurat_meta_no_thymocytes$annotation_level1,
  threshold_mat = level_1_threshold[, factor_names_high, drop = FALSE],
  compare      = ">"
)

## 2. compute median loading for each (level1, factor)
median_loading_level1 <- sapply(
  sort(unique(annotations_level1)),
  function(x) {
    apply(
      L_pm_no_thymocytes_normalized_high_AUC[annotations_level1[cells_all] == x, , drop = FALSE],
      2,
      median
    )
  }
)

# median_loading_level1: rows = Factor, cols = Annotation_Level1
median_loading_level1_df <- as.data.frame(median_loading_level1) |>
  rownames_to_column(var = "Factor") |>
  pivot_longer(
    cols      = -Factor,
    names_to  = "Annotation_Level1",
    values_to = "Median_Loading"
  )

## 3. proportion of cells in each level1 annotation, based on prop_mat
cell_prop_df <- data.frame(Annotation_Level1 = rownames(prop_mat)) |>
  left_join(
    as.data.frame(prop_mat) |>
      rownames_to_column(var = "Annotation_Level1") |>
      pivot_longer(
        cols      = -Annotation_Level1,
        names_to  = "Factor",
        values_to = "prop_cells"
      ),
    by = c("Annotation_Level1")
  )

# by Factor and Annotation_Level1
median_loading_level1_df <- median_loading_level1_df |>
  left_join(cell_prop_df, by = c("Annotation_Level1", "Factor"))

## 4. dot heatmap: y = Factor (rows), x = Level1 (columns)
median_loading_level1_df <- median_loading_level1_df |>
  mutate(
    Factor_num = as.numeric(sub("K", "", Factor))
  ) |>
  arrange(Factor_num) |>
  mutate(
    Factor = factor(Factor, levels = unique(Factor))
  )
```

```{r}
ggplot(median_loading_level1_df,
       aes(x = Annotation_Level1, y = Factor)) +
  geom_point(aes(
    size  = prop_cells,
    color = Median_Loading,
    alpha = Median_Loading
  )) +
  scale_color_gradient(
    low  = "white",
    high = "red"
  ) +
  scale_alpha_continuous(range = c(0, 5), guide = "none") +  
  scale_size_continuous(range = c(2, 8)) +
  theme_minimal() +
  labs(
    title = "Median Loadings for Level 1 Annotations",
    x = "Level 1 Annotations",
    y = "Factors",
    size  = "Proportion of cells",
    color = "Median loading"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(
  filename = "figures/Figures_Collection/Median_Loadings_Level1.png",
  width = 8, height = 8, dpi = 300
)

```

Create a similar heatplot for AUC:

```{r}
level_1_AUC_high_AUC <- level_1_AUC[, factors_high_AUC_level1, drop = FALSE]
level_1_AUC_high_AUC_df <- as.data.frame(level_1_AUC_high_AUC) |>
  rownames_to_column(var = "Annotation_Level1") |>
  pivot_longer(
    cols      = -Annotation_Level1,
    names_to  = "Factor",
    values_to = "AUC"
  )
level_1_AUC_high_AUC_df <- level_1_AUC_high_AUC_df |>
  mutate(
    Factor_num = as.numeric(sub("K", "", Factor))
  ) |>
  arrange(Factor_num) |>
  mutate(
    Factor = factor(Factor, levels = unique(Factor))
  )

# proportion of cells in each level1 annotation
# join by Annotation_Level1 and Factor
level_1_AUC_high_AUC_df <- level_1_AUC_high_AUC_df |>
  left_join(cell_prop_df, by = c("Annotation_Level1", "Factor"))
```

```{r}
ggplot(level_1_AUC_high_AUC_df,
       aes(x = Annotation_Level1, y = Factor)) +
  geom_point(aes(
    size  = prop_cells,
    color = AUC,
    alpha = AUC
  )) +
  scale_color_gradient(
    low  = "white",
    high = "blue"
  ) +
  scale_alpha_continuous(range = c(0, 0.7), guide = "none") +  
  scale_size_continuous(range = c(2, 8)) +
  theme_minimal() +
  labs(
    title = "AUC for Level 1 Annotations",
    x = "Level 1 Annotations",
    y = "Factors",
    size  = "Proportion of cells",
    color = "AUC"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(
  filename = "figures/Figures_Collection/AUC_Level1.png",
  width = 8, height = 8, dpi = 300
)
```


Next, look for variation across level 2 annotations:

Focus on GPs with max AUC > 0.95:

```{r}
max_AUC_level2 <- apply(level_2_AUC, 2, max)
factors_high_AUC_level2 <- which(max_AUC_level2 > 0.95)
factor_names_high <- colnames(level_1_AUC)[factors_high_AUC_level2]
colnames(L_pm_no_thymocytes) <- paste0("K", 1:ncol(L_pm_no_thymocytes))
L_pm_level2_high_AUC <- L_pm_no_thymocytes[, factor_names_high, drop = FALSE]
annotations_level2 <- seurat_meta$annotation_level2[-cells_thymocyte]
annotations_level2 <- as.character(annotations_level2)

# Don't need to renormalize, just use previous normalized matrix
L_pm_no_thymocytes_normalized_high_AUC <- L_pm_no_thymocytes_normalized[, factor_names_high, drop = FALSE]
```

```{r}
median_loading_level2 <- sapply(
  sort(unique(annotations_level2)),
  function(x) {
    apply(
      L_pm_no_thymocytes_normalized_high_AUC[annotations_level2[cells_all] == x, , drop = FALSE],
      2,
      median
    )
  }
)

# median_loading_level2: rows = Factor, cols = Annotation_Level2
median_loading_level2_df <- as.data.frame(median_loading_level2) |>
  rownames_to_column(var = "Factor") |>
  pivot_longer(
    cols      = -Factor,
    names_to  = "Annotation_Level2",
    values_to = "Median_Loading"
  )

# compute based on level_2_threshold

prop_mat <- compute_prop_above_threshold(
  loading_mat  = L_pm_level2_high_AUC,
  group_info   = seurat_meta_no_thymocytes$annotation_level2,
  threshold_mat = level_2_threshold[, factor_names_high, drop = FALSE],
  compare      = ">"
)

cell_prop_df <- data.frame(Annotation_Level2 = rownames(prop_mat)) |>
  left_join(
    as.data.frame(prop_mat) |>
      rownames_to_column(var = "Annotation_Level2") |>
      pivot_longer(
        cols      = -Annotation_Level2,
        names_to  = "Factor",
        values_to = "prop_cells"
      ),
    by = c("Annotation_Level2")
  )

median_loading_level2_df <- median_loading_level2_df |>
  left_join(cell_prop_df, by = c("Annotation_Level2", "Factor"))
```

To simplify plotting, only consider level 2 annotations that have at least a median loading > 0.4 in at least one of the selected factors:

```{r}
level2_to_keep <- median_loading_level2_df |>
  group_by(Annotation_Level2) |>
  summarize(max_median_loading = max(Median_Loading)) |>
  filter(max_median_loading > 0.4) |>
  pull(Annotation_Level2)

median_loading_level2_df <- median_loading_level2_df |>
  filter(Annotation_Level2 %in% level2_to_keep)
```


```{r}
ggplot(median_loading_level2_df,
       aes(x = Annotation_Level2, y = Factor)) +
  geom_point(aes(
    size  = prop_cells,
    color = Median_Loading,
    alpha = Median_Loading
  )) +
  scale_color_gradient(
    low  = "white",
    high = "red"
  ) +
  scale_alpha_continuous(range = c(0, 5), guide = "none") +  
  scale_size_continuous(range = c(2, 8)) +
  theme_minimal() +
  labs(
    title = "Median Loadings for Level 2 Annotations",
    x = "Level 2 Annotations",
    y = "Factors",
    size  = "Proportion of cells",
    color = "Median loading"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(
  filename = "figures/Figures_Collection/Median_Loadings_Level2.png",
  width = 10, height = 10, dpi = 500
)
```

```{r}
level_2_AUC_high_AUC <- level_2_AUC[, factors_high_AUC_level2, drop = FALSE]
level_2_AUC_high_AUC_df <- as.data.frame(level_2_AUC_high_AUC) |>
  rownames_to_column(var = "Annotation_Level2") |>
  pivot_longer(
    cols      = -Annotation_Level2,
    names_to  = "Factor",
    values_to = "AUC"
  )
level_2_AUC_high_AUC_df <- level_2_AUC_high_AUC_df |>
  mutate(
    Factor_num = as.numeric(sub("K", "", Factor))
  ) |>
  arrange(Factor_num) |>
  mutate(
    Factor = factor(Factor, levels = unique(Factor))
  )
level_2_AUC_high_AUC_df <- level_2_AUC_high_AUC_df |>
  left_join(cell_prop_df, by = c("Annotation_Level2", "Factor"))
```

Keep only level 2 annotations with AUC at least 0.95 in at least one of the selected factors:
```{r}
level2_to_keep_auc <- level_2_AUC_high_AUC_df |>
  group_by(Annotation_Level2) |>
  summarize(max_auc = max(AUC)) |>
  filter(max_auc > 0.98) |>
  pull(Annotation_Level2)

level_2_AUC_high_AUC_df <- level_2_AUC_high_AUC_df |>
  filter(Annotation_Level2 %in% level2_to_keep_auc)
```


```{r}
ggplot(level_2_AUC_high_AUC_df,
       aes(x = Annotation_Level2, y = Factor)) +
  geom_point(aes(
    size  = prop_cells,
    color = AUC,
    alpha = AUC
  )) +
  scale_color_gradient(
    low  = "white",
    high = "blue"
  ) +
  scale_alpha_continuous(range = c(0, 5), guide = "none") +  
  scale_size_continuous(range = c(2, 8)) +
  theme_minimal() +
  labs(
    title = "AUC for Level 2 Annotations",
    x = "Level 2 Annotations",
    y = "Factors",
    size  = "Proportion of cells",
    color = "AUC"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
ggsave(
  filename = "figures/Figures_Collection/AUC_Level2.png",
  width = 10, height = 10, dpi = 500
)
```

Next, look for variation across organ annotations:

Focus on GPs with max AUC > 0.9:

```{r}
max_AUC_organ <- apply(organ_AUC, 2, max)
factors_high_AUC_organ <- which(max_AUC_organ > 0.9)
factor_names_high <- colnames(level_1_AUC)[factors_high_AUC_organ]

colnames(L_pm) <- paste0("K", 1:ncol(L_pm))
L_pm_organ_high_AUC <- L_pm[, factor_names_high, drop = FALSE]
annotations_organ <- seurat_meta$annotation_organ[-cells_thymocyte]

# Take out cells for visualization
d <- apply(L_pm, 2, max)
L_pm_normalized <- scale_cols(L_pm, 1/d)
cells_all <- filter_cells_by_total_membership(L_pm_normalized, numiter = 15)
L_pm_normalized <- L_pm_normalized[cells_all, ]
d <- apply(L_pm_normalized, 2, max)
L_pm_normalized <- scale_cols(L_pm_normalized, 1/d)
L_pm_normalized_high_AUC <- L_pm_normalized[, factor_names_high, drop = FALSE]
```

```{r}
annotations_organ <- seurat_meta$organ
median_loading_organ <- sapply(
  sort(unique(annotations_organ)),
  function(x) {
    apply(
      L_pm_normalized_high_AUC[annotations_organ[cells_all] == x, , drop = FALSE],
      2,
      median
    )
  }
)

median_loading_organ_df <- as.data.frame(median_loading_organ) |>
  rownames_to_column(var = "Factor") |>
  pivot_longer(
    cols      = -Factor,
    names_to  = "Annotation_organ",
    values_to = "Median_Loading"
  )

cell_prop_df <- 

median_loading_organ_df <- median_loading_organ_df |>
  left_join(cell_prop_df, by = "Annotation_organ")


median_loading_organ_df <- median_loading_organ_df |>
  mutate(
    Factor_num = as.numeric(sub("K", "", Factor))
  ) |>
  arrange(Factor_num) |>
  mutate(
    Factor = factor(Factor, levels = unique(Factor))
  )
```

```{r}
organ_AUC_high_AUC <- organ_AUC[, factors_high_AUC_organ, drop = FALSE]
organ_AUC_high_AUC_df <- as.data.frame(organ_AUC_high_AUC) |>
  rownames_to_column(var = "Annotation_organ") |>
  pivot_longer(
    cols      = -Annotation_organ,
    names_to  = "Factor",
    values_to = "AUC"
  )
organ_AUC_high_AUC_df <- organ_AUC_high_AUC_df |>
  mutate(
    Factor_num = as.numeric(sub("K", "", Factor))
  ) |>
  arrange(Factor_num) |>
  mutate(
    Factor = factor(Factor, levels = unique(Factor))
  )

organ_AUC_high_AUC_df <- organ_AUC_high_AUC_df |>
  left_join(cell_prop_df, by = "Annotation_organ")
```

```{r}
# plot median loading heatmap for organ
ggplot(median_loading_organ_df,
       aes(x = Annotation_organ, y = Factor)) +
  geom_point(aes(
    size  = prop_cells,
    color = Median_Loading,
    alpha = Median_Loading
  )) +
  scale_color_gradient(
    low  = "white",
    high = "red"
  ) +
  scale_alpha_continuous(range = c(0, 5), guide = "none") +  
  scale_size_continuous(range = c(2, 5)) +
  theme_minimal() +
  labs(
    x = "Organ",
    y = "Factors",
    size  = "Proportion of cells",
    color = "Median loading"
  ) +
theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 2))

ggsave(
  filename = "figures/Figures_Collection/Median_Loadings_Organ.png",
  width = 10, height = 10, dpi = 500
)
```

```{r}
ggplot(organ_AUC_high_AUC_df,
       aes(x = Annotation_organ, y = Factor)) +
  geom_point(aes(
    size  = prop_cells,
    color = AUC,
    alpha = AUC
  )) +
  scale_color_gradient(
    low  = "white",
    high = "blue"
  ) +
  scale_alpha_continuous(range = c(0, 5), guide = "none") +  
  scale_size_continuous(range = c(2, 5)) +
  theme_minimal() +
  labs(
    title = "AUC for Organ Annotations",
    x = "Organ Annotations",
    y = "Factors",
    size  = "Proportion of cells",
    color = "AUC"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 2))

ggsave(
  filename = "figures/Figures_Collection/AUC_Organ.png",
  width = 10, height = 10, dpi = 500
)
```


Next, let's look at condition_detailed

```{r}
condition_detailed_AUC <- readRDS(paste0(data_path, "condition_detailed_AUC.rds"))

# only consider condition_detailed with more than 500 cells
annotations_condition <- seurat_meta$condition_detailed
which_less_500_condition <- sapply(sort(unique(annotations_condition)), function(x){
  sum(annotations_condition == x) < 500
})

condition_detailed_AUC_filtered <- condition_detailed_AUC[!which_less_500_condition, ]

max_AUC_condition <- apply(condition_detailed_AUC_filtered, 2, max)
factors_high_AUC_condition <- which(max_AUC_condition > 0.85)
factor_names_high_condition <- colnames(condition_detailed_AUC)[factors_high_AUC_condition]
```

```{r}
L_pm_normalized_high_AUC_condition <- L_pm_normalized[, factor_names_high_condition,
                                                    drop = FALSE]
```

```{r}
annotations_condition <- seurat_meta$condition_detailed
# compute median loading for each (condition_detailed, factor)
median_loading_condition <- sapply(
  sort(unique(annotations_condition)),
  function(x) {
    apply(
      L_pm_normalized_high_AUC_condition[annotations_condition[cells_all] == x, , drop = FALSE],
      2,
      median
    )
  }
)
# median_loading_condition: rows = Factor, cols = Annotation_condition
median_loading_condition_df <- as.data.frame(median_loading_condition) |>
  rownames_to_column(var = "Factor") |>
  pivot_longer(
    cols      = -Factor,
    names_to  = "Annotation_condition",
    values_to = "Median_Loading"
  )
# proportion of cells in each condition_detailed
cell_prop_condition_df <- data.frame(Annotation_condition = annotations_condition) |>
  count(Annotation_condition, name = "n_cells") |>
  mutate(prop_cells = n_cells / sum(n_cells))
median_loading_condition_df <- median_loading_condition_df |>
  left_join(cell_prop_condition_df, by = "Annotation_condition")
# dot heatmap: y = Factor (rows), x = condition_detailed (columns)
median_loading_condition_df <- median_loading_condition_df |>
  mutate(
    Factor_num = as.numeric(sub("K", "", Factor))
  ) |>
  arrange(Factor_num) |>
  mutate(
    Factor = factor(Factor, levels = unique(Factor))
  )
```


```{r}
ggplot(median_loading_condition_df,
       aes(x = Annotation_condition, y = Factor)) +
  geom_point(aes(
    size  = prop_cells,
    color = Median_Loading,
    alpha = Median_Loading    
  )) +
  scale_color_gradient(
    low  = "white",
    high = "red"
  ) +
  scale_alpha_continuous(range = c(0, 5), guide = "none") +  
  scale_size_continuous(range = c(2, 5)) +
  theme_minimal() +
  labs(
    x = "Condition Detailed",
    y = "Factors",
    size  = "Proportion of cells",
    color = "Median loading"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 4))

ggsave(
  filename = "figures/Figures_Collection/Median_Loadings_Condition_Detailed.png",
  width = 12, height = 14, dpi = 500
)
```

```{r}
# AUC for condition_detailed
condition_detailed_AUC_high_AUC <- condition_detailed_AUC_filtered[, factors_high_AUC_condition, drop = FALSE]
condition_detailed_AUC_high_AUC_df <- as.data.frame(condition_detailed_AUC_high_AUC) |>
  rownames_to_column(var = "Annotation_condition_detailed") |>
  pivot_longer(
    cols      = -Annotation_condition_detailed,
    names_to  = "Factor",
    values_to = "AUC"
  )
condition_detailed_AUC_high_AUC_df <- condition_detailed_AUC_high_AUC_df |>
  mutate(
    Factor_num = as.numeric(sub("K", "", Factor))
  ) |>
  arrange(Factor_num) |>
  mutate(
    Factor = factor(Factor, levels = unique(Factor))
  )
condition_detailed_AUC_high_AUC_df$Annotation_condition <- condition_detailed_AUC_high_AUC_df$Annotation_condition_detailed
condition_detailed_AUC_high_AUC_df <- condition_detailed_AUC_high_AUC_df |>
  left_join(cell_prop_condition_df, by = "Annotation_condition")
```

```{r}
ggplot(condition_detailed_AUC_high_AUC_df,
       aes(x = Annotation_condition, y = Factor)) +
  geom_point(aes(
    size  = prop_cells,
    color = AUC,
    alpha = AUC
  )) +
  scale_color_gradient(
    low  = "white",
    high = "blue"
  ) +
  scale_alpha_continuous(range = c(0, 5), guide = "none") +  
  scale_size_continuous(range = c(2, 5)) +
  theme_minimal() +
  labs(
    title = "AUC for Condition Detailed",
    x = "Condition Detailed",
    y = "Factors",
    size  = "Proportion of cells",
    color = "AUC"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 4))

ggsave(
  filename = "figures/Figures_Collection/AUC_Condition_Detailed.png",
  width = 12, height = 14, dpi = 500
)
```



## Table: View GPs in terms of max_AUC

We want to create table with rows denoting GPs, and columns denoting max_AUC in level-1, level-2, organ, and condition_detailed annotation.

```{r}
max_AUC_level1 <- apply(level_1_AUC, 2, max)
max_AUC_level2 <- apply(level_2_AUC_within_level1, 2, max)
max_AUC_organ <- apply(organ_AUC, 2, max)
max_AUC_condition_detailed <- apply(condition_detailed_AUC, 2, max)
gp_overview_df <- data.frame(
  Factor = colnames(L_pm),
  Max_AUC_Level1 = max_AUC_level1,
  Max_AUC_Level2 = max_AUC_level2,
  Max_AUC_Organ = max_AUC_organ,
  Max_AUC_Condition_Detailed = max_AUC_condition_detailed
)
write.csv(gp_overview_df, paste0(table_path, "GPs_Max_AUC_Overview.csv"), row.names = FALSE)
```

We also want to produce a table that shows in what categories for each time of meta-information does each GP predict well in terms of AUC > 0.895:

```{r}
summarize_auc_hits <- function(auc_list, threshold = 0.895) {
  # auc_list: list(level1 = level_1_AUC, level2 = level_2_AUC, ...)

  all_factors <- Reduce(union, lapply(auc_list, colnames))
  all_factors <- sort(all_factors)
  
  res <- data.frame(
    factor = all_factors,
    stringsAsFactors = FALSE
  )
  
  for (meta_name in names(auc_list)) {
    mat <- auc_list[[meta_name]]
    
    if (is.null(colnames(mat)) || is.null(rownames(mat))) {
      stop("Matrix in auc_list[['", meta_name, "']] must have rownames and colnames.")
    }
    
    meta_hits <- vapply(all_factors, function(k) {
      if (!k %in% colnames(mat)) {
        return(NA_character_)  
      }
      vals <- mat[, k]
      ok  <- which(!is.na(vals) & vals >= threshold)
      if (!length(ok)) {
        return("")  
      }
      paste(rownames(mat)[ok], collapse = "; ")
    }, character(1))
    
    res[[meta_name]] <- meta_hits
  }
  
  res
}
auc_list <- list(
  Level1 = level_1_AUC,
  Level2 = level_2_AUC_within_level1,
  Organ = organ_AUC,
  Condition_Detailed = condition_detailed_AUC,
  igt = igt_AUC
)
gp_auc_hits_df <- summarize_auc_hits(auc_list, threshold = 0.895)
gp_auc_hits_df <- gp_auc_hits_df[
  order(as.integer(sub("^K", "", gp_auc_hits_df$factor))),
]
# output to csv
write.csv(gp_auc_hits_df, paste0(table_path, "GPs_AUC_Hits_Overview.csv"), row.names = FALSE)
```



## Figure: Structure Plot:

Here we want to create a structure plot where each level-1 is matched to a GP that has the highest AUC for that level-1 annotation.

```{r}
match_category_to_factor <- function(auc_matrix) {
  n_factors    <- ncol(auc_matrix)
  n_categories <- nrow(auc_matrix)

  category_to_factor <- rep(NA_character_, n_categories)
  names(category_to_factor) <- rownames(auc_matrix)

  factor_assigned <- rep(FALSE, n_factors)
  names(factor_assigned) <- colnames(auc_matrix)

  category_score <- apply(auc_matrix, 1, max)    
  category_order <- order(category_score, decreasing = TRUE)
  ordered_categories <- rownames(auc_matrix)[category_order]

  for (cat in ordered_categories) {
    aucs <- auc_matrix[cat, ]

    aucs[factor_assigned] <- -Inf

    best_factor_idx <- which.max(aucs)
    best_factor_name <- names(aucs)[best_factor_idx]

    category_to_factor[cat] <- best_factor_name
    factor_assigned[best_factor_idx] <- TRUE
  }

  return(category_to_factor)
}
level1_category_to_factor <- match_category_to_factor(level_1_AUC)
color_coding <- ZemmourLib::immgent_colors$level1
```

```{r}
fit2 <- L_pm_no_thymocytes_normalized[, level1_category_to_factor, drop = FALSE]
cell_type <- seurat_meta[rownames(fit2), "annotation_level1"]
cells <- which(cell_type == "CD4" | cell_type == "CD8")
cells <- sample(cells,1e5)
cells <- sort(c(cells,which(cell_type != "CD4" & cell_type != "CD8")))

structure_plot(fit2[cells,],gap = 20,n = 3000,
               colors = color_coding[names(level1_category_to_factor)],
               grouping = cell_type[cells]) +
  labs(y = "membership",color = "",fill = "") +
  guides(
    fill = guide_legend(nrow = 1),
    color = guide_legend(nrow = 1)
  ) +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.box = "horizontal",
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13)
  )

ggsave(
  filename = "figures/Figures_Collection/Structure_Plot_Level1.png",
  width = 10, height = 6, dpi = 300
)
```








## Figure: Scatterplot of level2 versus organ

We will first load the result from healthy cells:

```{r eval=FALSE}
healthy_cells <- which(seurat_meta$condition_broad == "healthy")
L_pm <- L_pm[healthy_cells, ]
seurat_meta <- seurat_meta[healthy_cells, ]

selected_organ <- table(seurat_meta$organ)
selected_organ <- names(selected_organ[selected_organ >= 3000])
selected_level2 <- table(seurat_meta$annotation_level2)
selected_level2 <- names(selected_level2[selected_level2 >= 1000])
selected_cells <- which(seurat_meta$organ %in% selected_organ & seurat_meta$annotation_level2 %in% selected_level2)
seurat_meta_selected <- seurat_meta[selected_cells, ]
seurat_meta_selected$organ <- factor(seurat_meta_selected$organ)
seurat_meta_selected$annotation_level2 <- factor(seurat_meta_selected$annotation_level2)

compute_coef_per_factor <- function(factor_id){
  Y <- L_pm[selected_cells, factor_id]
  organ <- seurat_meta_selected$organ
  level2 <- seurat_meta_selected$annotation_level2
  model_full <- lm(Y ~ organ + level2)
  coefs <- coef(model_full)
  # Extract coefficients for organ
  organ_coefs <- coefs[grep("^organ", names(coefs))]
  # Extract coefficients for annotation_level2
  level2_coefs <- coefs[grep("^level2", names(coefs))]
  
  return(
    list(
      factor_id = factor_id,
      organ_coefs = organ_coefs,
      level2_coefs = level2_coefs
    )
  )
}

factor_coef_list <- lapply(1:ncol(L_pm), compute_coef_per_factor)
saveRDS(factor_coef_list, file = paste0(data_path, "conditional_analysis_factor.rds"))
```

```{r}
factor_coef_list <- readRDS(file = paste0(data_path, "conditional_analysis_factor.rds"))
```

```{r}
max_coef_df <- data.frame(
  Factor = integer(),
  Max_Abs_Coefficient_Organ = numeric(),
  Max_Abs_Coefficient_Level2 = numeric()
)

for (i in 1:length(factor_coef_list)) {
  coef_df <- factor_coef_list[[i]]
  max_abs_organ <- max(abs(coef_df$organ_coefs))
  max_abs_level2 <- max(abs(coef_df$level2_coefs))
  max_coef_df <- rbind(max_coef_df, data.frame(
    Factor = i,
    Max_Abs_Coefficient_Organ = max_abs_organ,
    Max_Abs_Coefficient_Level2 = max_abs_level2
  ))
}
```

```{r}
top_n <- 5  # how many points to annotate above and below y = x

df <- max_coef_df %>%
  mutate(
    residual = Max_Abs_Coefficient_Level2 - Max_Abs_Coefficient_Organ, # >0: above y=x; <0: below
    abs_res  = abs(residual)
  )

# --- Select top-N above and below y = x ---
label_above <- df %>%
  filter(is.finite(residual), residual > 0) %>%
  slice_max(Max_Abs_Coefficient_Level2, n = top_n, with_ties = FALSE)

label_below <- df %>%
  filter(is.finite(residual), residual < 0) %>%
  slice_min(-Max_Abs_Coefficient_Organ, n = top_n, with_ties = FALSE)

# --- Base plot builder (no labels yet) ---
base_plot <- ggplot(df, aes(x = Max_Abs_Coefficient_Organ, y = Max_Abs_Coefficient_Level2)) +
  geom_point(alpha = 0.3, size = 1.8) +                                # background points
  geom_point(data = label_above, color = "#1f78b4", alpha = 0.6, size = 1.8) +       # highlight above y=x
  geom_point(data = label_below, color = "#e31a1c", alpha = 0.6, size = 1.8) +       # highlight below y=x
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  coord_cartesian(xlim = c(0, 0.6), ylim = c(0, 0.6), expand = TRUE) +  # keep your limits
  labs(
    x = "Max Absolute Coefficient (Organ)",
    y = "Max Absolute Coefficient (Level-2)",
    title = "Max Absolute Coefficient: Organ vs Level-2"
  ) +
  theme_minimal()

# --- Combined plot: annotate both top-N above and below y = x ---
p_all <- base_plot +
  geom_text_repel(
    data = label_above,
    aes(label = paste0("Factor ", Factor)),
    color = "#1f78b4", size = 3, max.overlaps = Inf,
    box.padding = 0.3, point.padding = 0.2, min.segment.length = 0
  ) +
  geom_text_repel(
    data = label_below,
    aes(label = paste0("Factor ", Factor)),
    color = "#e31a1c", size = 3, max.overlaps = Inf,
    box.padding = 0.3, point.padding = 0.2, min.segment.length = 0
  ) +
  labs(subtitle = paste0("Annotating top ", top_n, " above (blue) and top ", top_n, " below (red) y = x"))

p_all

ggsave(
  filename = "figures/Figures_Collection/Scatterplot_Organ_vs_Level2.png",
  plot = p_all,
  width = 7, height = 6, dpi = 300
)
```

Produce some truncated boxplots for selected factors:

```{r}
healthy_cells <- which(seurat_meta$condition_broad == "healthy")
L_pm_healthy <- L_pm[healthy_cells, ]
seurat_meta_healthy <- seurat_meta[healthy_cells, ]

# Take out cells for visualization
d <- apply(L_pm_healthy, 2, max); L_pm_healthy_normalized <- scale_cols(L_pm_healthy, 1/d)
cells_all <- filter_cells_by_total_membership(L_pm_healthy_normalized, numiter = 15)
L_pm_healthy_normalized <- L_pm_healthy_normalized[cells_all, ]
d <- apply(L_pm_healthy_normalized, 2, max); L_pm_healthy_normalized <- scale_cols(L_pm_healthy_normalized, 1/d)
seurat_meta_healthy_normalized <- seurat_meta_healthy[rownames(L_pm_healthy_normalized), ]
```

For level 2:
```{r}
num_truncated <- 15
factor_id_vec <- c(8, 13, 93, 107)
p_list <- list()
for (factor_id in factor_id_vec) {
  df_boxplot <- data.frame(Loading = L_pm_healthy_normalized[, factor_id], Level2  = seurat_meta_healthy_normalized$annotation_level2)
  
  level2_stats <- df_boxplot %>%
    group_by(Level2) %>%
    summarise(median_loading = median(Loading, na.rm = TRUE),
              .groups = "drop")
  
  top_level2 <- level2_stats %>%
    slice_max(median_loading, n = num_truncated, with_ties = FALSE) %>%
    arrange(desc(median_loading))
  
  df_boxplot_top <- df_boxplot %>%
    filter(Level2 %in% top_level2$Level2) %>%
    mutate(Level2 = factor(Level2, levels = top_level2$Level2))
  
  p_new <- ggplot(df_boxplot_top, aes(x = Level2, y = Loading)) +
    geom_boxplot(outlier.shape = NA) +
    coord_cartesian(ylim = c(0, 1)) +
    labs(title = paste0("Factor ", factor_id),
         x = "Level-2 Annotation",
         y = "Normalized Loading") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  p_list[[as.character(factor_id)]] <- p_new
}
p_combined <- wrap_plots(p_list, ncol = 2)
p_combined
ggsave(
  filename = "figures/Figures_Collection/Boxplots_Level2_Selected_Factors.png",
  plot = p_combined,
  width = 12, height = 12, dpi = 400
)
```

For organ:
```{r}
factor_id_vec <- c(9, 37, 135, 177)
p_list <- list()
for (factor_id in factor_id_vec) {
  df_boxplot <- data.frame(Loading = L_pm_healthy_normalized[, factor_id], Organ  = seurat_meta_healthy_normalized$organ)
  
  Organ_stats <- df_boxplot %>%
    group_by(Organ) %>%
    summarise(median_loading = median(Loading, na.rm = TRUE),
              .groups = "drop")
  
  top_Organ <- Organ_stats %>%
    slice_max(median_loading, n = num_truncated, with_ties = FALSE) %>%
    arrange(desc(median_loading))
  
  df_boxplot_top <- df_boxplot %>%
    filter(Organ %in% top_Organ$Organ) %>%
    mutate(Organ = factor(Organ, levels = top_Organ$Organ))
  
  p_new <- ggplot(df_boxplot_top, aes(x = Organ, y = Loading)) +
    geom_boxplot(outlier.shape = NA) +
    coord_cartesian(ylim = c(0, 1)) +
    labs(title = paste0("Factor ", factor_id),
         x = "Organ Annotation",
         y = "Normalized Loading") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  p_list[[as.character(factor_id)]] <- p_new
}
p_combined <- wrap_plots(p_list, ncol = 2)
p_combined

ggsave(
  filename = "figures/Figures_Collection/Boxplots_Organ_Selected_Factors.png",
  plot = p_combined,
  width = 12, height = 12, dpi = 400
)
```


## Figure: Structure plot for level2

Within each level-1 category, we will produce a structure plot to visualize some highlighted factors across level-2 annotations.

```{r}
# load the AUC matrix for level2:
level_2_AUC <- readRDS(paste0(data_path, "level_2_AUC_figure.rds"))

# filter out annotations with less than 500 cells
which_less_500 <- sapply(rownames(level_2_AUC), function(x){
  sum(seurat_meta$annotation_level2 == x) < 500
})
level_2_AUC <- level_2_AUC[!which_less_500, ]

colnames(L_pm_normalized) <- paste0("K", 1:ncol(L_pm_normalized))
seurat_meta_filtered <- seurat_meta[rownames(L_pm_normalized), ]

all_level1 <- unique(seurat_meta$annotation_level1)
all_level1 <- all_level1[-which(all_level1 == "thymocyte")]
for (level1_cat in all_level1) {
  cells_in_cat <- which(seurat_meta_filtered$annotation_level1 == level1_cat)
  L_pm_cat <- L_pm_normalized[cells_in_cat, ]
  seurat_meta_cat <- seurat_meta_filtered[cells_in_cat, ]
  
  level2_in_cat <- unique(seurat_meta_cat$annotation_level2)
  level2_in_cat <- level2_in_cat[order(level2_in_cat)]
  
  # match level2 categories to factors
  selected_level2 <- grep(paste0(level1_cat, "_"), rownames(level_2_AUC), value = TRUE)
  level2_category_to_factor <- match_category_to_factor(level_2_AUC[selected_level2, ])

  # filter out cells in selected level2 categories
  cells_selected_level2 <- which(seurat_meta_cat$annotation_level2 %in% selected_level2)
  L_pm_cat <- L_pm_cat[cells_selected_level2, ]
  seurat_meta_cat <- seurat_meta_cat[cells_selected_level2, ]
  
  fit_cat <- L_pm_cat[, level2_category_to_factor, drop = FALSE]
  cell_type_cat <- seurat_meta_cat$annotation_level2
  
  # re-level cell_type_cat to remove unused levels
  cell_type_cat <- factor(cell_type_cat, levels = (selected_level2))

  # Make sure all the level2 categories have at least 500 cells
  level2_counts <- table(cell_type_cat)
  level2_selected <- names(level2_counts[level2_counts >= 500])
  cells <- which(cell_type_cat %in% level2_selected)
  
  structure_plot(fit_cat[cells,], gap = 20,n = 5000,
               grouping = cell_type_cat[cells]) +
  labs(y = "membership",color = "",fill = "") +
  guides(
    fill = guide_legend(nrow = 1),
    color = guide_legend(nrow = 1)
  ) +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.box = "horizontal",
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13)
  )

ggsave(
  filename = paste0("figures/Figures_Collection/Structure_Plot_", level1_cat, "_Level2.png"),
  width = 15, height = 6, dpi = 300
)
}
```




## Figure: Correlation plot of GPs

The correlation map of all GPs in all cells

```{r}
cor_matrix <- cor(L_pm, method = "pearson")
colnames(L_pm) <- paste0("K", seq_len(ncol(L_pm)))
colnames(cor_matrix) <- paste0("K", seq_len(ncol(cor_matrix)))
rownames(cor_matrix) <- paste0("K", seq_len(nrow(cor_matrix)))
```

```{r}
plot_gp_corr_by_level1 <- function(
  cor_full,        # filtered full correlation matrix (e.g., after high_corr_indices)
  L_pm,            # loading matrix: cells x GPs
  seurat_meta,     # metadata with annotation_level1, same row order as L_pm
  level1_type = c("all"),
  clustering_method = "complete",
  main_prefix = "Correlation map of GP loadings",
  filename = NULL,
  width = 8,
  height = 8,
  show_names = TRUE
) {
  # ---- Basic checks ----
  if (nrow(L_pm) != nrow(seurat_meta)) {
    stop("L_pm and seurat_meta must have the same number of rows (cells).")
  }
  if (is.null(seurat_meta$annotation_level1)) {
    stop("seurat_meta must contain a column named 'annotation_level1'.")
  }
  if (is.null(colnames(cor_full)) || is.null(rownames(cor_full))) {
    stop("cor_full must have rownames and colnames (GP names).")
  }
  if (is.null(colnames(L_pm))) {
    stop("L_pm must have column names (GP names) to match cor_full.")
  }
  
  # Make sure level1_type is scalar character
  level1_type <- as.character(level1_type)[1]
  
  ## ------------------------------------------------------------
  ## 1. Run pheatmap on full correlation ONCE to get the ordering
  ##    (this uses clustering; plot is suppressed with silent=TRUE)
  ## ------------------------------------------------------------
  res_all <- pheatmap(
    cor_full,
    clustering_method = clustering_method,
    show_rownames     = show_names,
    show_colnames     = show_names,
    color             = colorRampPalette(c("blue", "white", "red"))(100),
    main              = paste0(main_prefix, " (all cells, for ordering)"),
    treeheight_row    = 0,
    treeheight_col    = 0,
    silent            = TRUE        # don't actually draw this plot
  )
  
  gp_order <- res_all$tree_row$order
  gp_names <- rownames(cor_full)[gp_order]
  
  ## ------------------------------------------------------------
  ## 2. Construct the correlation matrix to plot for this level1
  ## ------------------------------------------------------------
  if (identical(level1_type, "all")) {
    # Just use the full correlation, reordered
    cor_plot  <- cor_full[gp_names, gp_names, drop = FALSE]
    main_title <- paste0(main_prefix, " (all cells)")
  } else {
    # Subset cells by level1_type
    cell_idx <- which(seurat_meta$annotation_level1 == level1_type)
    if (length(cell_idx) == 0) {
      stop("No cells found with annotation_level1 == ", level1_type)
    }
    
    L_sub <- L_pm[cell_idx, , drop = FALSE]
    
    # Keep the same GPs as in cor_full
    common_gps <- intersect(colnames(cor_full), colnames(L_sub))
    if (length(common_gps) == 0) {
      stop("No overlapping GP names between cor_full and L_pm.")
    }
    # enforce same order as gp_names
    common_gps <- gp_names[gp_names %in% common_gps]
    L_sub <- L_sub[, common_gps, drop = FALSE]
    
    # Correlation for this subset of cells
    cor_sub <- cor(L_sub, method = "pearson")
    
    # Reorder using the same gp_names ordering
    cor_plot  <- cor_sub[gp_names, gp_names, drop = FALSE]
    main_title <- paste0(main_prefix, " in ", level1_type, " cells")
  }
  
  ## ------------------------------------------------------------
  ## 3. Draw heatmap (no clustering here; we already fixed order)
  ## ------------------------------------------------------------
  if (is.null(filename)) {
    res <- pheatmap(
      cor_plot,
      cluster_rows   = FALSE,
      cluster_cols   = FALSE,
      show_rownames  = show_names,
      show_colnames  = show_names,
      color          = colorRampPalette(c("blue", "white", "red"))(100),
      main           = main_title,
      treeheight_row = 0,
      treeheight_col = 0
    )
  } else {
    res <- pheatmap(
      cor_plot,
      cluster_rows   = FALSE,
      cluster_cols   = FALSE,
      show_rownames  = show_names,
      show_colnames  = show_names,
      color          = colorRampPalette(c("blue", "white", "red"))(100),
      main           = main_title,
      treeheight_row = 0,
      treeheight_col = 0,
      filename       = filename,
      width          = width,
      height         = height
    )
  }
  
  invisible(list(
    cor_plot    = cor_plot,
    gp_order    = gp_order,
    gp_names    = gp_names,
    level1_type = level1_type
  ))
}
```


```{r}
# 1) All cells
plot_gp_corr_by_level1(
  cor_full    = cor_matrix,
  L_pm        = L_pm,
  seurat_meta = seurat_meta,
  width = 25,
  height = 25,
  level1_type = "all",
  filename    = "figures/Figures_Collection/Correlation_Map_All_GPs.png"
)

# 2) CD4
plot_gp_corr_by_level1(
  cor_full    = cor_matrix,
  L_pm        = L_pm,
  width = 25,
  height = 25,
  seurat_meta = seurat_meta,
  level1_type = "CD4",
  filename    = "figures/Figures_Collection/Correlation_Map_CD4_GPs.png"
)

# 3) CD8
plot_gp_corr_by_level1(
  cor_full    = cor_matrix,
  L_pm        = L_pm,
  seurat_meta = seurat_meta,
  level1_type = "CD8",
  width = 25,
  height = 25,
  filename    = "figures/Figures_Collection/Correlation_Map_CD8_GPs.png"
)

# 4) Treg
plot_gp_corr_by_level1(
  cor_full    = cor_matrix,
  L_pm        = L_pm,
  seurat_meta = seurat_meta,
  level1_type = "Treg",
  width = 25,
  height = 25,
  filename    = "figures/Figures_Collection/Correlation_Map_Treg_GPs.png"
)

# 5) DN
plot_gp_corr_by_level1(
  cor_full    = cor_matrix,
  L_pm        = L_pm,
  seurat_meta = seurat_meta,
  level1_type = "DN",
  width = 25,
  height = 25,
  filename    = "figures/Figures_Collection/Correlation_Map_DN_GPs.png"
)

# 6) DP
plot_gp_corr_by_level1(
  cor_full    = cor_matrix,
  L_pm        = L_pm,
  seurat_meta = seurat_meta,
  level1_type = "DP",
  width = 25,
  height = 25,
  filename    = "figures/Figures_Collection/Correlation_Map_DP_GPs.png"
)

# 7) nonconv
plot_gp_corr_by_level1(
  cor_full    = cor_matrix,
  L_pm        = L_pm,
  seurat_meta = seurat_meta,
  level1_type = "nonconv",
  width = 25,
  height = 25,
  filename    = "figures/Figures_Collection/Correlation_Map_nonconv_GPs.png"
)

# 8) gdT
plot_gp_corr_by_level1(
  cor_full    = cor_matrix,
  L_pm        = L_pm,
  seurat_meta = seurat_meta,
  level1_type = "gdT",
  width = 25,
  height = 25,
  filename    = "figures/Figures_Collection/Correlation_Map_gdT_GPs.png"
)

# 9) CD8aa
plot_gp_corr_by_level1(
  cor_full    = cor_matrix,
  L_pm        = L_pm,
  seurat_meta = seurat_meta,
  level1_type = "CD8aa",
  width = 25,
  height = 25,
  filename    = "figures/Figures_Collection/Correlation_Map_CD8aa_GPs.png"
)
```

Let's redo the plot focusing on high-pve factors only:

```{r}
high_pve_factors <- paste0("K", which(flashier_snmf_summary$pve >= 1e-4))
cor_matrix_high_pve <- cor_matrix[high_pve_factors, high_pve_factors]
L_pm_high_pve <- L_pm[, high_pve_factors, drop = FALSE]
```

```{r}
# 1) All cells
plot_gp_corr_by_level1(
  cor_full    = cor_matrix_high_pve,
  L_pm        = L_pm_high_pve,
  seurat_meta = seurat_meta,
  width = 10,
  height = 10,
  level1_type = "all",
  filename    = "figures/Figures_Collection/Correlation_Map_highPVE_GPs.png"
)

# 2) CD4
plot_gp_corr_by_level1(
  cor_full    = cor_matrix_high_pve,
  L_pm        = L_pm_high_pve,
  width = 10,
  height = 10,
  seurat_meta = seurat_meta,
  level1_type = "CD4",
  filename    = "figures/Figures_Collection/Correlation_Map_CD4_highPVE_GPs.png"
)

# 3) CD8
plot_gp_corr_by_level1(
  cor_full    = cor_matrix_high_pve,
  L_pm        = L_pm_high_pve,
  seurat_meta = seurat_meta,
  level1_type = "CD8",
  width = 10,
  height = 10,
  filename    = "figures/Figures_Collection/Correlation_Map_CD8_highPVE_GPs.png"
)

# 4) Treg
plot_gp_corr_by_level1(
  cor_full    = cor_matrix_high_pve,
  L_pm        = L_pm_high_pve,
  seurat_meta = seurat_meta,
  level1_type = "Treg",
  width = 10,
  height = 10,
  filename    = "figures/Figures_Collection/Correlation_Map_Treg_highPVE_GPs.png"
)

# 5) DN
plot_gp_corr_by_level1(
  cor_full    = cor_matrix_high_pve,
  L_pm        = L_pm_high_pve,
  seurat_meta = seurat_meta,
  level1_type = "DN",
  width = 10,
  height = 10,
  filename    = "figures/Figures_Collection/Correlation_Map_DN_highPVE_GPs.png"
)

# 6) DP
plot_gp_corr_by_level1(
  cor_full    = cor_matrix_high_pve,
  L_pm        = L_pm_high_pve,
  seurat_meta = seurat_meta,
  level1_type = "DP",
  width = 10,
  height = 10,
  filename    = "figures/Figures_Collection/Correlation_Map_DP_highPVE_GPs.png"
)

# 7) nonconv
plot_gp_corr_by_level1(
  cor_full    = cor_matrix_high_pve,
  L_pm        = L_pm_high_pve,
  seurat_meta = seurat_meta,
  level1_type = "nonconv",
  width = 10,
  height = 10,
  filename    = "figures/Figures_Collection/Correlation_Map_nonconv_highPVE_GPs.png"
)

# 8) gdT
plot_gp_corr_by_level1(
  cor_full    = cor_matrix_high_pve,
  L_pm        = L_pm_high_pve,
  seurat_meta = seurat_meta,
  level1_type = "gdT",
  width = 10,
  height = 10,
  filename = "figures/Figures_Collection/Correlation_Map_gdT_highPVE_GPs.png"
)

# 9) CD8aa
plot_gp_corr_by_level1(
  cor_full    = cor_matrix_high_pve,
  L_pm        = L_pm_high_pve,
  seurat_meta = seurat_meta,
  level1_type = "CD8aa",
  width = 10,
  height = 10,
  filename = "figures/Figures_Collection/Correlation_Map_CD8aa_highPVE_GPs.png"
)
```


## Figure: Scatterplot of GP correlations across level1 categories

We will compare the correlation between each pair of GPs in two different level-1 categories.

```{r}
library(ggplot2)
library(ggrepel)

# Compare pairwise GP correlations between two level-1 categories
plot_gp_corr_scatter <- function(
  level1_x,              # level-1 category for x-axis (e.g., "CD4")
  level1_y,              # level-1 category for y-axis (e.g., "Treg")
  L_pm,                  # loading matrix: cells x GPs
  seurat_meta,           # metadata with annotation_level1
  cor_full = NULL,       # optional: filtered full correlation to define GP set/order
  diff_threshold = 0,    # keep only pairs with |corr_y - corr_x| >= diff_threshold
  highlight_k = 0,       # number of most different pairs to label
  xlim = c(-1, 1),
  ylim = c(-1, 1),
  point_alpha = 0.5,
  title_prefix = "Pairwise GP correlations"
) {
  # Basic checks
  if (nrow(L_pm) != nrow(seurat_meta)) {
    stop("L_pm and seurat_meta must have the same number of rows (cells).")
  }
  if (is.null(seurat_meta$annotation_level1)) {
    stop("seurat_meta must contain a column named 'annotation_level1'.")
  }
  if (is.null(colnames(L_pm))) {
    stop("L_pm must have column names (GP names).")
  }
  
  # Select cells for each level-1 category
  idx_x <- which(seurat_meta$annotation_level1 == level1_x)
  idx_y <- which(seurat_meta$annotation_level1 == level1_y)
  if (length(idx_x) == 0) stop("No cells found with annotation_level1 == ", level1_x)
  if (length(idx_y) == 0) stop("No cells found with annotation_level1 == ", level1_y)
  
  L_x <- L_pm[idx_x, , drop = FALSE]
  L_y <- L_pm[idx_y, , drop = FALSE]
  
  # Correlation matrices in each category
  cor_x <- cor(L_x, method = "pearson")
  cor_y <- cor(L_y, method = "pearson")
  
  # Decide which GPs to keep
  if (!is.null(cor_full)) {
    gp_keep <- intersect(colnames(cor_full), intersect(colnames(cor_x), colnames(cor_y)))
  } else {
    gp_keep <- intersect(colnames(cor_x), colnames(cor_y))
  }
  if (length(gp_keep) < 2) stop("Not enough overlapping GPs to form pairs.")
  
  cor_x_sub <- cor_x[gp_keep, gp_keep, drop = FALSE]
  cor_y_sub <- cor_y[gp_keep, gp_keep, drop = FALSE]
  
  # Flatten upper triangle (i < j) into a data frame of GP pairs
  upper_idx <- which(upper.tri(cor_x_sub), arr.ind = TRUE)
  cor_df <- data.frame(
    GP1      = rownames(cor_x_sub)[upper_idx[, "row"]],
    GP2      = colnames(cor_x_sub)[upper_idx[, "col"]],
    cor_x    = cor_x_sub[upper_idx],
    cor_y    = cor_y_sub[upper_idx],
    stringsAsFactors = FALSE
  )
  
  # Difference between correlations
  cor_df$diff     <- cor_df$cor_y - cor_df$cor_x
  cor_df$abs_diff <- abs(cor_df$diff)
  
  # Optional filter: keep only pairs with |Δcor| >= diff_threshold
  if (!is.null(diff_threshold) && diff_threshold > 0) {
    cor_df <- subset(cor_df, abs_diff >= diff_threshold)
  }
  
  if (nrow(cor_df) == 0) {
    stop("No GP pairs left after applying diff_threshold filter.")
  }
  
  # Highlight the most different k pairs
  cor_df$label <- ""
  if (!is.null(highlight_k) && highlight_k > 0) {
    highlight_k <- min(highlight_k, nrow(cor_df))
    top_idx <- order(-cor_df$abs_diff)[seq_len(highlight_k)]
    cor_df$label[top_idx] <- paste(cor_df$GP1[top_idx], cor_df$GP2[top_idx], sep = "--")
  }
  
  # Build plot
  p <- ggplot(cor_df, aes(x = cor_x, y = cor_y)) +
    geom_point(alpha = point_alpha) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
    coord_equal(xlim = xlim, ylim = ylim) +
    labs(
      x = paste0("Correlation in ", level1_x, " cells"),
      y = paste0("Correlation in ", level1_y, " cells")
    ) +
    theme_minimal()
  
  # Add labels for highlighted pairs
  if (any(cor_df$label != "")) {
    p <- p + 
      ggrepel::geom_text_repel(
        data = subset(cor_df, label != ""),
        aes(label = label),
        size = 3,
        max.overlaps = Inf
      )
  }
  
  return(list(
    plot   = p,
    data   = cor_df,
    x_name = level1_x,
    y_name = level1_y
  ))
}
```

```{r}
res_treg_cd4 <- plot_gp_corr_scatter(
  level1_x      = "CD4",
  level1_y      = "Treg",
  L_pm          = L_pm_high_pve,
  seurat_meta   = seurat_meta,
  cor_full      = cor_matrix_high_pve,   
  diff_threshold = 0,           
  highlight_k    = 5           
)

res_treg_cd4$plot
```

```{r}
res_treg_cd8 <- plot_gp_corr_scatter(
  level1_x      = "CD8",
  level1_y      = "Treg",
  L_pm          = L_pm_high_pve,
  seurat_meta   = seurat_meta,
  cor_full      = cor_matrix_high_pve,   
  diff_threshold = 0,           
  highlight_k    = 5           
)

res_treg_cd8$plot
```


```{r}
res_treg_DN <- plot_gp_corr_scatter(
  level1_x      = "DN",
  level1_y      = "Treg",
  L_pm          = L_pm_high_pve,
  seurat_meta   = seurat_meta,
  cor_full      = cor_matrix_high_pve,   
  diff_threshold = 0,           
  highlight_k    = 5           
)

res_treg_DN$plot
```


```{r}
res_treg_DP <- plot_gp_corr_scatter(
  level1_x      = "DP",
  level1_y      = "Treg",
  L_pm          = L_pm_high_pve,
  seurat_meta   = seurat_meta,
  cor_full      = cor_matrix_high_pve,   
  diff_threshold = 0,           
  highlight_k    = 5           
)

res_treg_DP$plot
```

```{r}
# produce a 4-panel figure
p_combined <- wrap_plots(
  res_treg_cd4$plot + labs(subtitle = "Treg vs CD4"),
  res_treg_cd8$plot + labs(subtitle = "Treg vs CD8"),
  res_treg_DN$plot + labs(subtitle = "Treg vs DN"),
  res_treg_DP$plot + labs(subtitle = "Treg vs DP"),
  ncol = 2
)

ggsave(
  filename = "figures/Figures_Collection/Scatterplot_Treg_vs_Others.png",
  plot = p_combined,
  width = 12, height = 10, dpi = 300
)
```


```{r}
res_treg_CD8aa <- plot_gp_corr_scatter(
  level1_x      = "CD8aa",
  level1_y      = "Treg",
  L_pm          = L_pm_high_pve,
  seurat_meta   = seurat_meta,
  cor_full      = cor_matrix_high_pve,   
  diff_threshold = 0,           
  highlight_k    = 5           
)

res_treg_CD8aa$plot
```


```{r}
res_treg_nonconv <- plot_gp_corr_scatter(
  level1_x      = "nonconv",
  level1_y      = "Treg",
  L_pm          = L_pm_high_pve,
  seurat_meta   = seurat_meta,
  cor_full      = cor_matrix_high_pve,   
  diff_threshold = 0,           
  highlight_k    = 5           
)

res_treg_nonconv$plot
```

```{r}
res_treg_gdT <- plot_gp_corr_scatter(
  level1_x      = "gdT",
  level1_y      = "Treg",
  L_pm          = L_pm_high_pve,
  seurat_meta   = seurat_meta,
  cor_full      = cor_matrix_high_pve,   
  diff_threshold = 0,           
  highlight_k    = 5           
)
res_treg_gdT$plot
```

```{r}
# finally, also compare CD8aa vs gdT
res_CD8aa_gdT <- plot_gp_corr_scatter(
  level1_x      = "CD8aa",
  level1_y      = "gdT",
  L_pm          = L_pm_high_pve,
  seurat_meta   = seurat_meta,
  cor_full      = cor_matrix_high_pve,   
  diff_threshold = 0,           
  highlight_k    = 5           
)
res_CD8aa_gdT$plot
```


```{r}
# produce a 4-panel figure
p_combined_2 <- wrap_plots(
  res_treg_CD8aa$plot + labs(subtitle = "Treg vs CD8aa"),
  res_treg_nonconv$plot + labs(subtitle = "Treg vs nonconv"),
  res_treg_gdT$plot + labs(subtitle = "Treg vs gdT"),
  res_CD8aa_gdT$plot + labs(subtitle = "gdT vs CD8aa"),
  ncol = 2
)
ggsave(
  filename = "figures/Figures_Collection/Scatterplot_GP_cors_2.png",
  plot = p_combined_2,
  width = 12, height = 8, dpi = 300
)
```



