---
title: "Validating the EBMF result using halves of the data"
author: "Ziang Zhang"
date: "2025-10-15"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---


```{r}
source("https://raw.githubusercontent.com/stephenslab/single-cell-jamboree/refs/heads/main/code/group_factors.R")
library(ggplot2)
library(dplyr)
library(fastTopics)
library(qs)
library(cowplot)
library(ggrepel)
data_path <- "../data/"
code_path <- "../code/"
# data_path <- "data/"
# code_path <- "code/"
source(paste0(code_path, "ROC.R"))
seurat_meta <- readRDS(paste0(data_path, "seurat_meta.rds"))
# fit1_summary.qs
flashier_snmf_summary <- readRDS(paste0(data_path, "flashier_snmf_summary.rds"))
flashier_snmf_summary1 <- qs::qread(paste0(data_path, "fit1_summary.qs"))
flashier_snmf_summary2 <- qs::qread(paste0(data_path, "fit2_summary.qs"))
```

```{r}
F_all <- flashier_snmf_summary$F_pm
F_1 <- flashier_snmf_summary1$F_pm
F_2 <- flashier_snmf_summary2$F_pm

L_all <- flashier_snmf_summary$L_pm
L_1 <- flashier_snmf_summary1$L_pm
L_2 <- flashier_snmf_summary2$L_pm
```

## Match the factors between the two havles

To match factor in F_1 to F_2, we will compute the correlation between each pair of factors:

```{r}
# Normalize each column to have unit variance
F_1 <- scale(F_1, center = FALSE, scale = TRUE)
F_2 <- scale(F_2, center = FALSE, scale = TRUE)
```

```{r}
# Subset to have the common set of genes
common_genes <- intersect(rownames(F_1), rownames(F_2))
F_1 <- F_1[common_genes, ]
F_2 <- F_2[common_genes, ]
```

```{r}
# compute column-wise correlation
cor_F <- cor(F_1, F_2)
# find the best match for each factor in F_1
best_match_F <- apply(cor_F, 1, which.max)
best_cor_F <- apply(cor_F, 1, max)
# Create a data frame to store the results
match_results_F <- data.frame(
  Factor_F1 = 1:ncol(F_1),
  Best_Match_F2 = best_match_F,
  Correlation = best_cor_F
)
```

```{r}
hist(cor_F, breaks = 50, main = "Histogram of Correlations between F1 and F2", xlab = "Correlation")
```

```{r}
image((cor_F)>0.5)
```

```{r}
# Doing a bar plot to visualize the overlap of matched factors for each factor in F_1
# correlation on y-axis and factor in F_1 on x-axis
plot_F <- ggplot(match_results_F, aes(x = Factor_F1, y = Correlation)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  labs(title = "Best Match Correlation for Each Factor in F_1",
       x = "Factor in F_1",
       y = "Best Match Correlation with F_2") +
  theme_minimal()
print(plot_F)
```


Let's focus on the non-overlapped matched factors:

```{r}
# if there exists overlap of matched F_2 factors, we will only keep the one with the highest correlation
match_results_F_filtered <- match_results_F %>%
  group_by(Best_Match_F2) %>%
  slice_max(order_by = Correlation, n = 1, with_ties = FALSE) %>%
  arrange(Factor_F1) %>%
  ungroup()
```

We will only consider the factors with correlation > 0.5:

```{r}
selected_factors_F <- match_results_F_filtered %>%
  filter(Correlation > 0.5)
nrow(selected_factors_F)
```

```{r}
plot_F_filtered <- ggplot(selected_factors_F, aes(x = Factor_F1, y = Correlation)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  labs(title = "Best Match Correlation for Selected Factors in F_1",
       x = "Factor in F_1",
       y = "Best Match Correlation with F_2") +
  theme_minimal()
print(plot_F_filtered)
```

## Match the factor from the whole data analysis

Now, let's match the factors from the whole data analysis (F_all) to the two halves (F_1 and F_2).

```{r}
# Normalize each column to have unit variance
F_all <- scale(F_all, center = FALSE, scale = TRUE)
```

```{r}
# Subset to have the common set of genes
common_genes_all_1 <- intersect(rownames(F_all), rownames(F_1))
common_genes_all_2 <- intersect(rownames(F_all), rownames(F_2))
common_genes_all <- intersect(common_genes_all_1, common_genes_all_2)
F_all <- F_all[common_genes_all, ]
F_1 <- F_1[common_genes_all, ]
F_2 <- F_2[common_genes_all, ]
```

```{r}
# compute column-wise correlation
cor_F_all_1 <- cor(F_all, F_1)
cor_F_all_2 <- cor(F_all, F_2)

# find the best match for each factor in F_all
best_match_F_all_1 <- apply(cor_F_all_1, 1, which.max)
best_cor_F_all_1 <- apply(cor_F_all_1, 1, max)
best_match_F_all_2 <- apply(cor_F_all_2, 1, which.max)
best_cor_F_all_2 <- apply(cor_F_all_2, 1, max)
# Create a data frame to store the results
match_results_F_all <- data.frame(
  Factor_F_all = 1:ncol(F_all),
  Best_Match_F1 = best_match_F_all_1,
  Correlation_F1 = best_cor_F_all_1,
  Best_Match_F2 = best_match_F_all_2,
  Correlation_F2 = best_cor_F_all_2
)
match_results_F_all$MaxCorr <- pmax(match_results_F_all$Correlation_F1, match_results_F_all$Correlation_F2)
match_results_F_all[c(1,2,3,5,22,58, 68, 171), 1:5]
```


```{r}
par(mfrow=c(1,2))
hist(cor_F_all_1, breaks = 50, main = "Corr: F_all and F_1", xlab = "Correlation")
hist(cor_F_all_2, breaks = 50, main = "Corr: F_all and F_2", xlab = "Correlation")
par(mfrow=c(1,1))
```

```{r}
# select factors with correlation > 0.5
selected_factors_F_all <- match_results_F_all %>%
  filter(MaxCorr > 0.5)
nrow(selected_factors_F_all)
```

We will only keep the non-overlapped matched factors with the highest correlation.
If two factors from F_all are matched to the same factor in F_1 or F_2, we will only keep the one with the highest correlation.

```{r}
selected_factors_F_all_unique <- selected_factors_F_all %>%
  arrange(desc(MaxCorr)) %>%
  distinct(Best_Match_F1, .keep_all = TRUE) %>%  
  distinct(Best_Match_F2, .keep_all = TRUE) %>%
  arrange(Factor_F_all)
nrow(selected_factors_F_all_unique)
```

Bar plot to visualize the correlation between the matched factors and each factor in F_all:

```{r}
plot_F1 <- ggplot(selected_factors_F_all_unique, aes(x = Factor_F_all, y = Correlation_F1)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  labs(title = "Best Match Correlation for Each Factor in F_all vs F_1",
       x = "Factor in F_all",
       y = "Best Match Correlation with F_1") +
  theme_minimal()

plot_F2 <- ggplot(selected_factors_F_all_unique, aes(x = Factor_F_all, y = Correlation_F2)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  labs(title = "Best Match Correlation for Each Factor in F_all vs F_2",
       x = "Factor in F_all",
       y = "Best Match Correlation with F_2") +
  theme_minimal()

gridExtra::grid.arrange(plot_F1, plot_F2, nrow = 2)
```


How many large pve factors are there in the whole data analysis?

```{r}
sum(which(flashier_snmf_summary$pve > 1e-4) %in% selected_factors_F_all$Factor_F_all)/sum(flashier_snmf_summary$pve > 1e-4)

sum(which(flashier_snmf_summary$pve > 1e-4) %in% selected_factors_F_all_unique$Factor_F_all)/sum(flashier_snmf_summary$pve > 1e-4)
```

Structure plot:

```{r}
# normalize by the 99 percent quantile in each column
D_all <- diag(1/apply(L_all, 2, function(x) quantile(x, 0.99)))
L_all <- L_all %*% D_all
D_1 <- diag(1/apply(L_1, 2, function(x) quantile(x, 0.99)))
L_1 <- L_1 %*% D_1
D_2 <- diag(1/apply(L_2, 2, function(x) quantile(x, 0.99)))
L_2 <- L_2 %*% D_2
```


```{r, results="hide", message=FALSE}
## structure plot for all the data
set.seed(12345)
annotation_sub <- seurat_meta$annotation_level1[match(rownames(L_all), seurat_meta$cellID)]
cells <- which(annotation_sub == "CD4" | annotation_sub == "CD8")
cells <- sample(cells,1e5)
cells <- sort(c(cells,which(annotation_sub != "CD4" & annotation_sub != "CD8")))
selected_all_Factors <- c(22, 29, 30, 58, 68, 171)
L_all_plot <- L_all[cells,selected_all_Factors]
colnames(L_all_plot) <- paste0("K",selected_all_Factors)
p_all <- structure_plot(L_all_plot, gap = 10, n = 6000,
                      topics = paste0("K",selected_all_Factors),
                    grouping = annotation_sub[cells]) +
  labs(y = "membership",color = "",fill = "") +
  ylim(0,3) +
  theme(legend.position = "bottom")

## structure plot for the first half data
selected_1_Factors <- selected_factors_F_all$Best_Match_F1[match(selected_all_Factors,selected_factors_F_all$Factor_F_all)]
annotation_sub1 <- seurat_meta$annotation_level1[match(rownames(L_1), seurat_meta$cellID)]
cells1 <- which(annotation_sub1 == "CD4" | annotation_sub1 == "CD8")
cells1 <- sample(cells1,5e4)
cells1 <- sort(c(cells1,which(annotation_sub1 != "CD4" & annotation_sub1 != "CD8")))
L_1_plot <- L_1[cells1, selected_1_Factors]
colnames(L_1_plot) <- paste0("K",selected_1_Factors)
p_1 <- structure_plot(L_1_plot, gap = 10, n = 6000,
                      topics = paste0("K",selected_1_Factors),
                    grouping = annotation_sub1[cells1]) +
  labs(y = "membership",color = "",fill = "") +
  ylim(0,3) +
  theme(legend.position = "bottom")
                    
## structure plot for the second half data
selected_2_Factors <- selected_factors_F_all$Best_Match_F2[match(selected_all_Factors, selected_factors_F_all$Factor_F_all)]
annotation_sub2 <- seurat_meta$annotation_level1[match(rownames(L_2), seurat_meta$cellID)]
cells2 <- which(annotation_sub2 == "CD4" | annotation_sub2 == "CD8")
cells2 <- sample(cells2,5e4)
cells2 <- sort(c(cells2,which(annotation_sub2 != "CD4" & annotation_sub2 !=  "CD8")))
L_2_plot <- L_2[cells2, selected_2_Factors]
colnames(L_2_plot) <- paste0("K",selected_2_Factors)
p_2 <- structure_plot(L_2_plot, gap = 10, n = 6000,
                      topics = paste0("K",selected_2_Factors),
                    grouping = annotation_sub2[cells2]) +
  labs(y = "membership",color = "",fill = "") +
  ylim(0,3) +
  theme(legend.position = "bottom")
```

```{r, fig.width=10, fig.height=10}
plot_grid(p_all, p_1, p_2, nrow = 3)
```
