---
title: "Validating the EBMF result using halves of the data"
author: "Ziang Zhang"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Validation Analysis

This time, we repeat the validation analysis using the halved-analysis without backfitting.

```{r}
library(ggplot2)
library(dplyr)
library(fastTopics)
library(qs)
library(cowplot)
library(ggrepel)
data_path <- "../data/"
code_path <- "../code/"
# data_path <- "data/"
# code_path <- "code/"

seurat_meta <- readRDS(paste0(data_path, "seurat_meta.rds"))
flashier_snmf_summary <- readRDS(paste0(data_path, "flashier_snmf_summary.rds"))
flashier_snmf_summary1 <- qs::qread(paste0(data_path, "fit1_summary_noback.qs"))
flashier_snmf_summary2 <- qs::qread(paste0(data_path, "fit2_summary_noback.qs"))
```


Define a function to compute cosine similarity:

```{r}
compute_cosine_sim_matrix <- function(L1, L2){
  norms1 <- apply(L1, 2, function(x){sqrt(sum(x^2))})
  norms1[norms1 == 0] <- Inf
  
  norms2 <- apply(L2, 2, function(x){sqrt(sum(x^2))})
  norms2[norms2 == 0] <- Inf
  
  L1_normalized <- t(t(L1)/norms1)
  L2_normalized <- t(t(L2)/norms2)
  
  #compute matrix of cosine similarities
  cosine_sim_matrix <- crossprod(L1_normalized, L2_normalized)
  
  return(cosine_sim_matrix)
}
```

Normalize the factor matrices:

```{r}
F_all <- flashier_snmf_summary$F_pm
F_1 <- flashier_snmf_summary1$F_pm
F_2 <- flashier_snmf_summary2$F_pm

L_all <- flashier_snmf_summary$L_pm
L_1 <- flashier_snmf_summary1$L_pm
L_2 <- flashier_snmf_summary2$L_pm
```

```{r}
F_1 <- scale(F_1, center = FALSE, scale = TRUE)
F_2 <- scale(F_2, center = FALSE, scale = TRUE)
F_all <- scale(F_all, center = FALSE, scale = TRUE)
```

Now, let's match the factors from the whole data analysis (F_all) to the two halves (F_1 and F_2).
This time, we will be using the Hungarian algorithm to find the optimal matching.

```{r}
# Subset to have the common set of genes
common_genes_all_1 <- intersect(rownames(F_all), rownames(F_1))
common_genes_all_2 <- intersect(rownames(F_all), rownames(F_2))
common_genes_all <- intersect(common_genes_all_1, common_genes_all_2)
F_all <- F_all[common_genes_all, ]
F_1 <- F_1[common_genes_all, ]
F_2 <- F_2[common_genes_all, ]
```

```{r}
cos_1 <- compute_cosine_sim_matrix(F_all, F_1)
assignment_problem_1 <- RcppHungarian::HungarianSolver(-1*cos_1)
cos_2 <- compute_cosine_sim_matrix(F_all, F_2)
assignment_problem_2 <- RcppHungarian::HungarianSolver(-1*cos_2)

# summarize the results into a data frame
match_results <- data.frame(
  Factor_All = 1:ncol(F_all),
  Best_Match_F1 = assignment_problem_1$pairs[,2],
  Best_Match_F2 = assignment_problem_2$pairs[,2],
  cosine_F1 = cos_1[assignment_problem_1$pairs],
  cosine_F2 = cos_2[assignment_problem_2$pairs]
)

match_results[c(1,2,3,5,22,58, 68, 171), 1:5]
```


We should also take a look at the cosine similarity between the matched factors in F_1 and F_2.

```{r}
cos_f1_f1 <- compute_cosine_sim_matrix(F_1[, match_results$Best_Match_F1], F_2[, match_results$Best_Match_F2])
match_results$cosine_F1_F2 <- diag(cos_f1_f1)

match_results[c(1,2,3,5,22,58, 68, 171), c(2:6)]
```

Plot the scatterplot of the cosine similarities, and see how the similarities vary with PVE:

```{r}
match_results$PVE_All <- log10(flashier_snmf_summary$pve[match_results$Factor_All])
```

```{r}
p1 <- ggplot(match_results, aes(x = cosine_F1, y = cosine_F2, color = PVE_All)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Cosine Similarity (Whole Data vs 1st Half)", y = "Cosine Similarity (Whole Data vs 2nd Half)") +
  theme_minimal()

p2 <- ggplot(match_results, aes(x = cosine_F1, y = cosine_F1_F2, color = PVE_All)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Cosine Similarity (Whole Data vs 1st Half)", y = "Cosine Similarity (1st Half vs 2nd Half)") +
  theme_minimal()


# Whole Data - 2nd Half against 1st Half - 2nd Half
p3 <- ggplot(match_results, aes(x = cosine_F2, y = cosine_F1_F2, color = PVE_All)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Cosine Similarity (Whole Data vs 2nd Half)", y = "Cosine Similarity (1st Half vs 2nd Half)") +
  theme_minimal()

plot_grid(p1, p2, p3, nrow = 2)
```


This table should be a useful reference, let's save it as a csv file.

```{r eval=FALSE}
write.csv(match_results, file = paste0(data_path, "factor_matching_results_noback.csv"), row.names = FALSE)
```


Let's find the factors in F_all that have a good match (similarity > 0.5) in either F_1 or F_2.

```{r}
# match_results$MaxCorr <- pmax(match_results$Correlation_F1, match_results$Correlation_F2)
match_results$MaxCosine <- pmax(match_results$cosine_F1, match_results$cosine_F2)

# select factors with similarity > 0.5 and between F_1 and F_2 > 0.5
selected_factors_F_all <- match_results %>%
  filter(MaxCosine > 0.5 & cosine_F1_F2 > 0.5)
  # filter(MaxCorr > 0.5 & Correlation_F1_F2 > 0.5)
nrow(selected_factors_F_all)
```

What are the final selected factors in F_all?

```{r}
selected_factors_F_all$Factor_All
```

Structure plot:

```{r}
filter_cells_by_total_membership <- function (L, max_val = 10, numiter = 10) {
  n <- nrow(L)
  rows <- 1:n
  for (iter in 1:numiter) {
    x <- rowSums(L)
    cat(sprintf("%d. Filtered out %d cells.\n",iter,sum(x > max_val)))
    i <- which(x <= max_val)
    L <- L[i,]
    rows <- rows[i]
    d <- apply(L,2,max)
    L <- scale_cols(L,1/d)
  }
  return(rows)
}
scale_cols <- function (A, b)
  t(t(A) * b)
```


```{r}
# Take out cells for visualization
D_all <- diag(1 / apply(L_all, 2, function(x) max(x)))
L_all <- L_all %*% D_all
cells_all <- filter_cells_by_total_membership(L_all, numiter = 12)
L_all <- L_all[cells_all,]
d <- apply(L_all,2,max)
L_all <- scale_cols(L_all,1/d)
D_all <- d * D_all
seurat_meta_all <- seurat_meta[match(rownames(L_all), seurat_meta$cellID),]

D_1 <- diag(1 / apply(L_1, 2, function(x) max(x)))
L_1 <- L_1 %*% D_1
cells_1 <- filter_cells_by_total_membership(L_1, numiter = 12)
L_1 <- L_1[cells_1,]
d <- apply(L_1,2,max)
L_1 <- scale_cols(L_1,1/d)
D_1 <- d * D_1
seurat_meta_1 <- seurat_meta[match(rownames(L_1), seurat_meta$cellID),]


D_2 <- diag(1 / apply(L_2, 2, function(x) max(x)))
L_2 <- L_2 %*% D_2
cells_2 <- filter_cells_by_total_membership(L_2, numiter = 12)
L_2 <- L_2[cells_2,]
d <- apply(L_2,2,max)
L_2 <- scale_cols(L_2,1/d)
D_2 <- d * D_2
seurat_meta_2 <- seurat_meta[match(rownames(L_2), seurat_meta$cellID),]
```

```{r, results="hide", message=FALSE}
set.seed(12345)
annotation_sub <- seurat_meta_all$annotation_level1
cells <- which(annotation_sub == "CD4" | annotation_sub == "CD8")
cells <- sample(cells,1e5)
cells <- sort(c(cells,which(annotation_sub != "CD4" & annotation_sub != "CD8")))
selected_all_Factors <- c(22, 30, 58, 68, 171)
L_all_plot <- L_all[cells,selected_all_Factors]
colnames(L_all_plot) <- paste0("K",selected_all_Factors)
p_all <- structure_plot(L_all_plot, gap = 10, n = 6000,
                      topics = paste0("K",selected_all_Factors),
                    grouping = annotation_sub[cells]) +
  labs(y = "membership",color = "",fill = "") +
  theme(legend.position = "bottom")

## structure plot for the first half data
selected_1_Factors <- match_results$Best_Match_F1[match(selected_all_Factors,match_results$Factor_All)]
annotation_sub1 <- seurat_meta_1$annotation_level1
cells1 <- which(annotation_sub1 == "CD4" | annotation_sub1 == "CD8")
cells1 <- sample(cells1,5e4)
cells1 <- sort(c(cells1,which(annotation_sub1 != "CD4" & annotation_sub1 != "CD8")))
L_1_plot <- L_1[cells1, selected_1_Factors]
colnames(L_1_plot) <- paste0("K",selected_1_Factors)
p_1 <- structure_plot(L_1_plot, gap = 10, n = 6000,
                      topics = paste0("K",selected_1_Factors),
                    grouping = annotation_sub1[cells1]) +
  labs(y = "membership",color = "",fill = "") +
  # ylim(0,0.2) +
  theme(legend.position = "bottom")
                    
## structure plot for the second half data
selected_2_Factors <- match_results$Best_Match_F2[match(selected_all_Factors, match_results$Factor_All)]
annotation_sub2 <- seurat_meta_2$annotation_level1
cells2 <- which(annotation_sub2 == "CD4" | annotation_sub2 == "CD8")
cells2 <- sample(cells2,5e4)
cells2 <- sort(c(cells2,which(annotation_sub2 != "CD4" & annotation_sub2 !=  "CD8")))
L_2_plot <- L_2[cells2, selected_2_Factors]
colnames(L_2_plot) <- paste0("K",selected_2_Factors)
p_2 <- structure_plot(L_2_plot, gap = 10, n = 6000,
                      topics = paste0("K",selected_2_Factors),
                    grouping = annotation_sub2[cells2]) +
  labs(y = "membership",color = "",fill = "") +
  # ylim(0,0.2) +
  theme(legend.position = "bottom")
```

```{r}
plot_grid(p_all, p_1, p_2, nrow = 3)
```


## Outlier Analysis

I am also interested to know whether the loading from each half analysis (without backfitting) produces similar outlier cells as the whole data analysis.

To make the comparison clear, let's scale by median.

```{r}
L_all <- flashier_snmf_summary$L_pm
d <- apply(L_all,2,median)
L_all <- scale_cols(L_all,1/d)

L_1 <- flashier_snmf_summary1$L_pm
d <- apply(L_1,2,median)
L_1 <- scale_cols(L_1,1/d)

L_2 <- flashier_snmf_summary2$L_pm
d <- apply(L_2,2,median)
L_2 <- scale_cols(L_2,1/d)
```

Let's define outlier cells as those that have loading greater than median + 3 * IQR for each factor.

```{r}
find_outlier_cells <- function(L){
  outlier_list <- list()
  for (k in 1:ncol(L)){
    loadings <- L[,k]
    threshold <- median(loadings) + 3 * IQR(loadings)
    outlier_cells <- which(loadings > threshold)
    outlier_list[[k]] <- outlier_cells
  }
  return(outlier_list)
}

outliers_all <- find_outlier_cells(L_all)
outliers_1 <- find_outlier_cells(L_1)
outliers_2 <- find_outlier_cells(L_2)
```

Keep track of the number of outlier cells for each factor:

```{r}
num_outliers_all <- sapply(outliers_all, length)
prop_outliers_all <- num_outliers_all / nrow(L_all)

num_outliers_1 <- sapply(outliers_1, length)
prop_outliers_1 <- num_outliers_1 / nrow(L_1)

num_outliers_2 <- sapply(outliers_2, length)
prop_outliers_2 <- num_outliers_2 / nrow(L_2)
```


Summarize the distribution of outlier proportions:

```{r}
summary(prop_outliers_all)
summary(prop_outliers_1)
summary(prop_outliers_2)
```


Turning off the backfitting does not decrease the number of outlier cells detected.
Let's see if the extent of outlierness (loading values) are similar between the whole data and the halves.


Take a look at the boxplot for some matched factors:

```{r}
par(mfrow = c(2,2))
for (i in 1:4) {
  boxplot(L_all[,selected_factors_F_all$Factor_All[i]], L_1[,selected_factors_F_all$Best_Match_F1[i]], L_2[,selected_factors_F_all$Best_Match_F2[i]], names = c("All","1st Half","2nd Half"), main = paste0("Factor ", selected_factors_F_all$Factor_All[i], " Loadings"))
}
```
