---
title: "Validating the EBMF result using halves of the data"
author: "Ziang Zhang"
date: "2025-10-16"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r}
library(ggplot2)
library(dplyr)
library(fastTopics)
library(qs)
library(cowplot)
library(ggrepel)
# data_path <- "../data/"
# code_path <- "../code/"
data_path <- "data/"
code_path <- "code/"
seurat_meta <- readRDS(paste0(data_path, "seurat_meta.rds"))
flashier_snmf_summary <- readRDS(paste0(data_path, "flashier_snmf_summary.rds"))
flashier_snmf_summary1 <- qs::qread(paste0(data_path, "fit1_summary.qs"))
flashier_snmf_summary2 <- qs::qread(paste0(data_path, "fit2_summary.qs"))
```

Define a function to compute cosine similarity:

```{r}
compute_cosine_sim_matrix <- function(L1, L2){
  norms1 <- apply(L1, 2, function(x){sqrt(sum(x^2))})
  norms1[norms1 == 0] <- Inf
  
  norms2 <- apply(L2, 2, function(x){sqrt(sum(x^2))})
  norms2[norms2 == 0] <- Inf
  
  L1_normalized <- t(t(L1)/norms1)
  L2_normalized <- t(t(L2)/norms2)
  
  #compute matrix of cosine similarities
  cosine_sim_matrix <- crossprod(L1_normalized, L2_normalized)
  
  return(cosine_sim_matrix)
}
```

(Optional: some function to nonlinearly scale the cosine similarity for better hungarian matching)
```{r}
# Base S(t): quartic with change-point at c (S(0)=0, S(1)=1, S'(0)=S'(1)=0, S(c)=c)
S_quartic_cp <- function(t, c = 0.3){
  A <- (2*c - 1)/(c^2 - c)
  B <- (-2*c^2 - 2*c + 2)/(c^2 - c)
  C <- (3*c^2 - c - 1)/(c^2 - c)
  A*t^4 + B*t^3 + C*t^2
}

# baseline map: mix linear and S, then restore sign
f_cubic_cp <- function(x, c = 0.3, lambda = 1) {
  stopifnot(lambda >= 0, lambda <= 1, c > 0, c < 1)
  t <- abs(x)
  S <- S_quartic_cp(t, c)
  sign(x) * ((1 - lambda)*t + lambda*S)
}

# --------------------------------
# Amplifier A: relative-odds power
# --------------------------------
amplify_odds <- function(u, c = 0.3, q = 3){
  oc <- c/(1 - c)               # odds(c)
  r  <- (u/(1 - u)) / oc        # relative odds vs c
  o2 <- oc * r^q                # powered odds
  o2/(1 + o2)                   # back to [0,1]
}

f_strong_odds <- function(x, c = 0.3, lambda = 1, q = 3){
  stopifnot(lambda >= 0, lambda <= 1, c > 0, c < 1, q > 0)
  t  <- abs(x)
  S  <- S_quartic_cp(t, c)
  u  <- (1 - lambda)*t + lambda*S   # if you want max effect, set lambda=1
  u2 <- amplify_odds(u, c = c, q = q)
  sign(x) * u2
}

x <- seq(-1, 1, by = 0.01)
plot(x, f_cubic_cp(x, c = 0.4, lambda = 1), type = "l", col = "blue", ylim = c(-1,1))
lines(x, f_strong_odds(x, c = 0.4, lambda = 1, q = 2), col = "red")
lines(x, f_strong_odds(x, c = 0.4, lambda = 1, q = 4), col = "purple")
lines(x,x, col = "black", lty = 2)
legend("topleft", legend = c("f_cubic_cp","f_strong_odds (q=2)","f_strong_odds (q=4)","identity"), col = c("blue","red","purple","black"), lty = c(1,1,1,2))

```


Normalize the factor matrices:

```{r}
F_all <- flashier_snmf_summary$F_pm
F_1 <- flashier_snmf_summary1$F_pm
F_2 <- flashier_snmf_summary2$F_pm

L_all <- flashier_snmf_summary$L_pm
L_1 <- flashier_snmf_summary1$L_pm
L_2 <- flashier_snmf_summary2$L_pm
```

```{r}
F_1 <- scale(F_1, center = FALSE, scale = TRUE)
F_2 <- scale(F_2, center = FALSE, scale = TRUE)
F_all <- scale(F_all, center = FALSE, scale = TRUE)
```

Now, let's match the factors from the whole data analysis (F_all) to the two halves (F_1 and F_2).
This time, we will be using the Hungarian algorithm to find the optimal matching.

```{r}
# Subset to have the common set of genes
common_genes_all_1 <- intersect(rownames(F_all), rownames(F_1))
common_genes_all_2 <- intersect(rownames(F_all), rownames(F_2))
common_genes_all <- intersect(common_genes_all_1, common_genes_all_2)
F_all <- F_all[common_genes_all, ]
F_1 <- F_1[common_genes_all, ]
F_2 <- F_2[common_genes_all, ]
```

```{r}
cos_1 <- compute_cosine_sim_matrix(F_all, F_1)
# cos_1 <- cor(F_all, F_1)

# optionally apply nonlinear scaling
# cos_1_f <- f_cubic_cp(cos_1, c = 0.3, lambda = 1)
# cos_1_f <- f_strong_odds(cos_1, c = 0.3, lambda = 1, q = 4)
# assignment_problem_1 <- RcppHungarian::HungarianSolver(-1*cos_1_f)
assignment_problem_1 <- RcppHungarian::HungarianSolver(-1*cos_1)

cos_2 <- compute_cosine_sim_matrix(F_all, F_2)
# cos_2 <- cor(F_all, F_2)

# optionally apply nonlinear scaling
# cos_2_f <- f_cubic_cp(cos_2, c = 0.3, lambda = 1)
# cos_2_f <- f_strong_odds(cos_2, c = 0.3, lambda = 1, q = 4)
# assignment_problem_2 <- RcppHungarian::HungarianSolver(-1*cos_2_f)
assignment_problem_2 <- RcppHungarian::HungarianSolver(-1*cos_2)

# summarize the results into a data frame
match_results <- data.frame(
  Factor_All = 1:ncol(F_all),
  Best_Match_F1 = assignment_problem_1$pairs[,2],
  Best_Match_F2 = assignment_problem_2$pairs[,2],
  cosine_F1 = cos_1[assignment_problem_1$pairs],
  cosine_F2 = cos_2[assignment_problem_2$pairs]
)

match_results[c(1,2,3,5,22,58, 68, 171), 1:5]
match_results[1:10,1:5]
```


We should also take a look at the cosine similarity between the matched factors in F_1 and F_2.

```{r}
cos_f1_f2 <- compute_cosine_sim_matrix(F_1[, match_results$Best_Match_F1], F_2[, match_results$Best_Match_F2])
match_results$cosine_F1_F2 <- diag(cos_f1_f2)

match_results[c(1,2,3,5,22,58, 68, 171), c(2:6)]
```


This table should be a useful reference, let's save it as a csv file.

```{r eval=FALSE}
write.csv(match_results, file = paste0(data_path, "factor_matching_results.csv"), row.names = FALSE)
```

Take a look at the distribution of the similarities:

```{r}
par(mfrow=c(1,3))
hist(match_results$cosine_F1, breaks = 50, main = "Cosine: F_all and F_1", xlab = "Cosine Similarity")
hist(match_results$cosine_F2, breaks = 50, main = "Cosine: F_all and F_2", xlab = "Cosine Similarity")
hist(match_results$cosine_F1_F2, breaks = 50, main = "Cosine: F_1 and F_2", xlab = "Cosine Similarity")
par(mfrow=c(1,1))
```

Let's find the factors in F_all that have a good match (similarity > 0.5) in either F_1 or F_2.

```{r}
match_results$MaxCosine <- pmax(match_results$cosine_F1, match_results$cosine_F2)
match_results$MinCosine <- pmin(match_results$cosine_F1, match_results$cosine_F2)

# select factors with similarity > 0.5 and between F_1 and F_2 > 0.5
selected_factors_F_all <- match_results %>%
  filter(MinCosine > 0.5 & cosine_F1_F2 > 0.5)
  # filter(MaxCorr > 0.5 & Correlation_F1_F2 > 0.5)
nrow(selected_factors_F_all)
```

```{r eval=FALSE}
# if MinCosine > 0.5 and cosine_F1_F2 > 0.5 then is strongly validated
match_results$strongly_validated <- (match_results$MinCosine > 0.5)& (match_results$cosine_F1_F2 > 0.5)
# if MinCosine > 0.4 and cosine_F1_F2 > 0.4 then is moderately validated
match_results$moderately_validated <- (match_results$MinCosine > 0.4)& (match_results$cosine_F1_F2 > 0.4)
# if MaxCosine > 0.4 then is weakly validated
match_results$weakly_validated <- (match_results$MaxCosine > 0.4)

# if the max cosine with either F_1 or F_2 > 0.4 for some factor then is eventually validated
opt_F1 <- apply(cos_1, 1, which.max)
opt_F2 <- apply(cos_2, 1, which.max)
optimized_cos1 <- cos_1[cbind(1:nrow(cos_1), opt_F1)]; optimized_cos2 <- cos_2[cbind(1:nrow(cos_2), opt_F2)]
optimized_max_cos <- pmax(optimized_cos1, optimized_cos2)
match_results$eventually_validated <- optimized_max_cos > 0.4


sum(match_results$strongly_validated)
sum(match_results$moderately_validated)
sum(match_results$weakly_validated)
sum(match_results$eventually_validated)

write.csv(match_results, file = paste0(data_path, "factor_matching_results_filtered.csv"), row.names = FALSE)
```

What are the final selected factors in F_all?

```{r}
selected_factors_F_all$Factor_All
```


Bar plot to visualize the similarity between the matched factors and each factor in F_all:

```{r}
plot_F1 <- ggplot(selected_factors_F_all, aes(x = Factor_All, y = cosine_F1)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  labs(title = "Best Match Cosine Similarity for Each Factor in F_all vs F_1",
       x = "Factor in F_all",
       y = "Best Match Cosine Similarity with F_1") +
  theme_minimal()

plot_F2 <- ggplot(selected_factors_F_all, aes(x = Factor_All, y = cosine_F2)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  labs(title = "Best Match Cosine Similarity for Each Factor in F_all vs F_2",
       x = "Factor in F_all",
       y = "Best Match Cosine Similarity with F_2") +
  theme_minimal()

gridExtra::grid.arrange(plot_F1, plot_F2, nrow = 2)
```


How many large pve factors are there in the whole data analysis?

```{r}
sum(which(flashier_snmf_summary$pve > 1e-4) %in% selected_factors_F_all$Factor_All)/sum(flashier_snmf_summary$pve > 1e-4)
```


Structure plot:

```{r}
filter_cells_by_total_membership <- function (L, max_val = 10, numiter = 10) {
  n <- nrow(L)
  rows <- 1:n
  for (iter in 1:numiter) {
    x <- rowSums(L)
    cat(sprintf("%d. Filtered out %d cells.\n",iter,sum(x > max_val)))
    i <- which(x <= max_val)
    L <- L[i,]
    rows <- rows[i]
    d <- apply(L,2,max)
    L <- scale_cols(L,1/d)
  }
  return(rows)
}
scale_cols <- function (A, b)
  t(t(A) * b)
```


```{r}
# Take out cells for visualization
D_all <- diag(1 / apply(L_all, 2, function(x) max(x)))
L_all <- L_all %*% D_all
cells_all <- filter_cells_by_total_membership(L_all, numiter = 12)
L_all <- L_all[cells_all,]
d <- apply(L_all,2,max)
L_all <- scale_cols(L_all,1/d)
D_all <- d * D_all
seurat_meta_all <- seurat_meta[match(rownames(L_all), seurat_meta$cellID),]

D_1 <- diag(1 / apply(L_1, 2, function(x) max(x)))
L_1 <- L_1 %*% D_1
cells_1 <- filter_cells_by_total_membership(L_1, numiter = 12)
L_1 <- L_1[cells_1,]
d <- apply(L_1,2,max)
L_1 <- scale_cols(L_1,1/d)
D_1 <- d * D_1
seurat_meta_1 <- seurat_meta[match(rownames(L_1), seurat_meta$cellID),]

D_2 <- diag(1 / apply(L_2, 2, function(x) max(x)))
L_2 <- L_2 %*% D_2
cells_2 <- filter_cells_by_total_membership(L_2, numiter = 12)
L_2 <- L_2[cells_2,]
d <- apply(L_2,2,max)
L_2 <- scale_cols(L_2,1/d)
D_2 <- d * D_2
seurat_meta_2 <- seurat_meta[match(rownames(L_2), seurat_meta$cellID),]
```


```{r, results="hide", message=FALSE}
## structure plot for all the data
set.seed(12345)
annotation_sub <- seurat_meta_all$annotation_level1
cells <- which(annotation_sub == "CD4" | annotation_sub == "CD8")
cells <- sample(cells,1e5)
cells <- sort(c(cells,which(annotation_sub != "CD4" & annotation_sub != "CD8")))
selected_all_Factors <- c(7, 22, 30, 58, 68, 171)
L_all_plot <- L_all[cells,selected_all_Factors]
colnames(L_all_plot) <- paste0("K",selected_all_Factors)
p_all <- structure_plot(L_all_plot, gap = 10, n = 6000,
                      topics = paste0("K",selected_all_Factors),
                    grouping = annotation_sub[cells]) +
  labs(y = "membership",color = "",fill = "") +
  # ylim(0,0.2) +
  theme(legend.position = "bottom")

## structure plot for the first half data
selected_1_Factors <- match_results$Best_Match_F1[match(selected_all_Factors,match_results$Factor_All)]

# Visually, the best matching for factor 7 seems to be factor 7 in the first half data...
selected_1_Factors[1] <- which.max(cos_1[7,]);


annotation_sub1 <- seurat_meta_1$annotation_level1
cells1 <- which(annotation_sub1 == "CD4" | annotation_sub1 == "CD8")
cells1 <- sample(cells1,5e4)
cells1 <- sort(c(cells1,which(annotation_sub1 != "CD4" & annotation_sub1 != "CD8")))
L_1_plot <- L_1[cells1, selected_1_Factors]
colnames(L_1_plot) <- paste0("K",selected_1_Factors)
p_1 <- structure_plot(L_1_plot, gap = 10, n = 6000,
                      topics = paste0("K",selected_1_Factors),
                    grouping = annotation_sub1[cells1]) +
  labs(y = "membership",color = "",fill = "") +
  # ylim(0,0.2) +
  theme(legend.position = "bottom")
                    
## structure plot for the second half data
selected_2_Factors <- match_results$Best_Match_F2[match(selected_all_Factors, match_results$Factor_All)]
annotation_sub2 <- seurat_meta_2$annotation_level1
cells2 <- which(annotation_sub2 == "CD4" | annotation_sub2 == "CD8")
cells2 <- sample(cells2,5e4)
cells2 <- sort(c(cells2,which(annotation_sub2 != "CD4" & annotation_sub2 !=  "CD8")))
L_2_plot <- L_2[cells2, selected_2_Factors]
colnames(L_2_plot) <- paste0("K",selected_2_Factors)
p_2 <- structure_plot(L_2_plot, gap = 10, n = 6000,
                      topics = paste0("K",selected_2_Factors),
                    grouping = annotation_sub2[cells2]) +
  labs(y = "membership",color = "",fill = "") +
  # ylim(0,0.2) +
  theme(legend.position = "bottom")
```

```{r, fig.width=10, fig.height=10}
plot_grid(p_all, p_1, p_2, nrow = 3)
```

