---
title: "Linking GEPs to protein"
author: "Ziang Zhang"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

From the analysis of scRNA data, we obtain the following matrix factorization:

\[ \mathbf{X} = \mathbf{L}\mathbf{F}^T, \]
where \(\mathbf{F}\) denotes the learned GEPs, and \(\mathbf{L}\) denotes the loading of these GEPs in each cell.


In this analysis, we fix the loading matrix \(\mathbf{L}\), just try to obtain the factor matrix \(\mathbf{U}\) from the protein data \(\mathbf{Y}\):
\[\mathbf{Y} = \mathbf{L}\mathbf{U}^T,\]
where \(\mathbf{U}\) denotes the protein levels corresponding to each GEP.

## Analysis


```{r eval = FALSE}
library(flashier)
library(fastTopics)

data_path <- "../data/"
# data_path <- "data/"

protein_mat_normalized <- readRDS(file = paste0(data_path, "protein_mat_normalized.rds"))
scRNA_result <- readRDS(file = paste0(data_path, "flashier_snmf_summary.rds"))

k <- ncol(scRNA_result$L_pm)
factors_considered <- which(scRNA_result$pve > 1e-4)
L_mat <- scRNA_result$L_pm[rownames(protein_mat_normalized), factors_considered, drop = FALSE]
F_mat_int <- matrix(0, nrow = ncol(protein_mat_normalized), ncol = length(factors_considered))

# remove rows that are entirely zero
zero_rows <- which(rowSums(protein_mat_normalized) == 0)
if(length(zero_rows) > 0){
  protein_mat_normalized <- protein_mat_normalized[-zero_rows, , drop = FALSE]
  L_mat <- L_mat[-zero_rows, , drop = FALSE]
}

# remove columns that are entirely zero
zero_cols <- which(colSums(protein_mat_normalized) == 0)
if(length(zero_cols) > 0){
  protein_mat_normalized <- protein_mat_normalized[, -zero_cols, drop = FALSE]
  F_mat_int <- F_mat_int[-zero_cols, , drop = FALSE]
}


# run flashier fixed loading
flash_fixed_loading <- flash_init(protein_mat_normalized, var_type = 2) |>
  flash_set_verbose(1) |>
  flash_factors_init(list(L_mat,
                          F_mat_int)) |>
  flash_factors_fix(kset = 1:length(factors_considered),
                    which_dim = "loadings") |>
  flash_backfit()


protein_flash_summary <- list(L_pm = flash_fixed_loading$L_pm,
                              F_pm = flash_fixed_loading$F_pm,
                              pve = flash_fixed_loading$pve,
                              elbo = flash_fixed_loading$elbo,
                              residuals_sd = flash_fixed_loading$residuals_sd)

saveRDS(protein_flash_summary, file = paste0(data_path, "protein_flash_selected_summary.rds"))
```


```{r}
library(fastTopics)
library(cowplot)
library(ggplot2)
data_path <- "../data/"
# data_path <- "data/"
protein_flash_summary <- readRDS(file = paste0(data_path, "protein_flash_selected_summary.rds"))

protein_flash_summary_lognorm <- readRDS(file = paste0(data_path, "protein_flash_selected_summary_lognorm.rds"))

flashier_snmf_summary <- readRDS(paste0(data_path, "flashier_snmf_summary.rds"))
```


```{r}
D <- diag(1 / apply(protein_flash_summary$F_pm, 2, function(x) (max(x) - min(x))))
protein_flash_summary$F_pm <- protein_flash_summary$F_pm %*% D
protein_flash_summary$L_pm <- protein_flash_summary$L_pm %*% solve(D)

D_lognorm <- diag(1 / apply(protein_flash_summary_lognorm$F_pm, 2, function(x) (max(x) - min(x))))
protein_flash_summary_lognorm$F_pm <- protein_flash_summary_lognorm$F_pm %*% D_lognorm
protein_flash_summary_lognorm$L_pm <- protein_flash_summary_lognorm$L_pm %*% solve(D_lognorm)
```

### Cell-type-Level1 annotation

Take a look at the protein expression corresponding to some selected GEPs that were annotated as cell-type level 1 related.

```{r}
annot_level1 <- readRDS(paste0(data_path, "annot_level1.rds"))
plot(annot_level1$rank, annot_level1$stats, type = "o", xlab = "Rank", ylab = "R2", main = "Celltype Level1", ylim = c(0,1.1))
text(annot_level1$rank, annot_level1$stats, labels = annot_level1$factor_original, pos = 3, cex = 0.5)
abline(h = c(0.1,0.2,0.4), col = "red", lty = 2)
threshold_R2 <- 0.2
threshold_rows <- 10
annot_level1_highlight <- annot_level1[annot_level1$stats >= threshold_R2, ]
annot_level1_highlight <- annot_level1_highlight[1:threshold_rows, ]
annot_level1_highlight <- na.omit(annot_level1_highlight)
```

```{r}
F_pm <- protein_flash_summary$F_pm
colnames(F_pm) <- paste0("F", 1:ncol(flashier_snmf_summary$F_pm))[ flashier_snmf_summary$pve > 1e-4]
kset <- paste0("F", annot_level1_highlight$factor_original[1:5])
F_pm <- F_pm[, kset]
p1 <- annotation_heatmap(F_pm,n = 2,
                         dim = kset,
                         show_dims = kset,
                         select_features = "largest",
                         font_size = 9) +
  labs(title = "select_features = \"largest\"") +
  theme(plot.title = element_text(face = "plain",size = 9))
p2 <- annotation_heatmap(F_pm,n = 2,
                         dim = kset,
                         show_dims = kset,
                         select_features = "distinctive",
                         compare_dims = kset,
                         font_size = 9) +
  labs(title = "select_features = \"distinctive\"") +
  theme(plot.title = element_text(face = "plain",size = 9))
plot_grid(p1,p2,nrow = 1,ncol = 2)
```


Focus on selected GEPs:

```{r}
F_pm <- protein_flash_summary$F_pm
colnames(F_pm) <- paste0("F", 1:ncol(flashier_snmf_summary$F_pm))[ flashier_snmf_summary$pve > 1e-4]

# F_pm <- protein_flash_nn_summary$F_pm
# colnames(F_pm) <- paste0("F", 1:ncol(flashier_snmf_summary$F_pm))[flashier_snmf_summary$pve > 1e-4]
# F_pm <- F_pm[, protein_flash_nn_summary$pve > 0]

kset <- paste0("F", c(68,58,30,3,22,29))
F_pm <- F_pm[, kset]

annotation_heatmap(F_pm,n = 2,
                         dim = kset,
                         show_dims = kset,
                         select_features = "largest",
                         compare_dims = kset,
                         font_size = 10) +
  theme(plot.title = element_text(face = "plain",size = 10))

# ggsave2("figures/protein_factor_selected_GEPs_largest.pdf", width = 6, height = 6)
```

```{r}
# F_pm <- protein_flash_summary$F_pm
# colnames(F_pm) <- paste0("F", 1:ncol(flashier_snmf_summary$F_pm))[ flashier_snmf_summary$pve > 1e-4]

F_pm <- protein_flash_summary_lognorm$F_pm
colnames(F_pm) <- paste0("F", 1:ncol(flashier_snmf_summary$F_pm))[flashier_snmf_summary$pve > 1e-4]
F_pm <- F_pm[, protein_flash_summary_lognorm$pve > 0]

kset <- paste0("F", c(68,58,30,3,22,29))
F_pm <- F_pm[, kset]

annotation_heatmap(F_pm,n = 2,
                         dim = kset,
                         show_dims = kset,
                         select_features = "largest",
                         compare_dims = kset,
                         font_size = 10) +
  theme(plot.title = element_text(face = "plain",size = 10))

# ggsave2("figures/protein_factor_selected_GEPs_largest_lognorm.pdf", width = 6, height = 6)
```




### Cell-type-Level2 annotation

Take a look at the protein expression corresponding to some selected GEPs that were annotated as cell-type level 2 related.


```{r}
annot_level2 <- readRDS(paste0(data_path, "annot_level2.rds"))
plot(annot_level2$rank, annot_level2$stats, type = "o", xlab = "Rank", ylab = "R2", main = "Celltype Level1", ylim = c(0,1.1))
text(annot_level2$rank, annot_level2$stats, labels = annot_level2$factor_original, pos = 3, cex = 0.5)
abline(h = c(0.1,0.2,0.4), col = "red", lty = 2)
threshold_R2 <- 0.2
threshold_rows <- 10
annot_level2_highlight <- annot_level2[annot_level2$stats >= threshold_R2, ]
annot_level2_highlight <- annot_level2_highlight[1:threshold_rows, ]
annot_level2_highlight <- na.omit(annot_level2_highlight)
```


```{r}
F_pm <- protein_flash_summary$F_pm
colnames(F_pm) <- paste0("F", 1:ncol(flashier_snmf_summary$F_pm))[ flashier_snmf_summary$pve > 1e-4]
kset <- paste0("F", annot_level2_highlight$factor_original[1:5])
F_pm <- F_pm[, kset]
p1 <- annotation_heatmap(F_pm,n = 2,
                         dim = kset,
                         show_dims = kset,
                         select_features = "largest",
                         font_size = 9) +
  labs(title = "select_features = \"largest\"") +
  theme(plot.title = element_text(face = "plain",size = 9))
p2 <- annotation_heatmap(F_pm,n = 2,
                         dim = kset,
                         show_dims = kset,
                         select_features = "distinctive",
                         compare_dims = kset,
                         font_size = 9) +
  labs(title = "select_features = \"distinctive\"") +
  theme(plot.title = element_text(face = "plain",size = 9))
plot_grid(p1,p2,nrow = 1,ncol = 2)
```



### Organ annotation

Take a look at the protein expression corresponding to some selected GEPs that were annotated as organ related.

```{r}
organ <- readRDS(paste0(data_path, "organ.rds"))
plot(organ$rank, organ$stats, type = "o", xlab = "Rank", ylab = "R2", main = "Organ", ylim = c(0,1.1))
text(organ$rank, organ$stats, labels = organ$factor_original, pos = 3, cex = 0.5)
abline(h = c(0.1,0.2,0.4), col = "red", lty = 2)
threshold_R2 <- 0.2
threshold_rows <- 10
organ_highlight <- organ[organ$stats >= threshold_R2, ]
organ_highlight <- organ_highlight[1:threshold_rows, ]
organ_highlight <- na.omit(organ_highlight)
```

```{r}
F_pm <- protein_flash_summary$F_pm
colnames(F_pm) <- paste0("F", 1:ncol(flashier_snmf_summary$F_pm))[ flashier_snmf_summary$pve > 1e-4]
kset <- paste0("F", organ_highlight$factor_original[1:5])
F_pm <- F_pm[, kset]
p1 <- annotation_heatmap(F_pm,n = 2,
                         dim = kset,
                         show_dims = kset,
                         select_features = "largest",
                         font_size = 9) +
  labs(title = "select_features = \"largest\"") +
  theme(plot.title = element_text(face = "plain",size = 9))
p2 <- annotation_heatmap(F_pm,n = 2,
                         dim = kset,
                         show_dims = kset,
                         select_features = "distinctive",
                         compare_dims = kset,
                         font_size = 9) +
  labs(title = "select_features = \"distinctive\"") +
  theme(plot.title = element_text(face = "plain",size = 9))
plot_grid(p1,p2,nrow = 1,ncol = 2)
```






