---
title: Semi-NMF analysis of the T cell data
author: Peter Carbonetto
output: workflowr::wflow_html
---

This is my attempt at making sense of the semi-NMF results (with $K =
200$ factors).

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load the packages needed for this analysis:

```{r load-pkgs, message=FALSE}
library(Matrix)
library(fastTopics)
library(flashier)
library(ggplot2)
library(cowplot)
```

Load the meta-data and the semi-NMF results:

```{r load-data}
load("../data/flashier_snmf_ldf.RData")
L <- fl_snmf_ldf$L
F <- fl_snmf_ldf$F
rm(fl_snmf_ldf)
```

Renormalize the memberships in a way that leads to better Structure
plots:

```{r renormalize-memberships}
L <- apply(L,2,function (x) x/quantile(x,0.999))
```

This is my attempt at identifying factors that might capture
the major cell types ("level 1" annotation):

```{r identify-cell-type-factors-1, fig.height=2.5, fig.width=2.5}
cell_type <- sample_info$annotation_level1
n <- nlevels(cell_type)
k <- ncol(L)
B <- matrix(0,k,n)
rownames(B) <- colnames(L)
colnames(B) <- levels(cell_type)
for (i in levels(cell_type)) {
  j <- which(cell_type == i)
  B[,i] <- apply(L[j,],2,median)
}
y <- apply(B,1,function (x) diff(range(x)))
k_set <- order(y,decreasing = TRUE)
qplot(1:k,y[k_set]) +
  labs(x = "rank",y = "value") +
  theme_cowplot(font_size = 12)
```

These are the factors that show the largest differences:

```{r identify-cell-type-factors-2}
y[y > 0.1]
```

Now create a Structure plot with the cells grouped by major cell-type
(the "level 1" annotation):

```{r structure-plot-by-cell-type, fig.height=2.25, fig.width=7.25, results="hide", message=FALSE}
set.seed(1)
cell_type_factors <- c("k3","k7","k22","k27","k29","k30","k58","k68",
                       "k101","k127")
cell_type <- sample_info$annotation_level1
cells <- which(cell_type == "CD4" | cell_type == "CD8")
cells <- sample(cells,1e5)
cells <- sort(c(cells,which(cell_type != "CD4" & cell_type != "CD8")))
structure_plot(L[cells,cell_type_factors],gap = 10,n = 2000,
                    grouping = cell_type[cells]) +
  labs(y = "membership",color = "",fill = "") +
  ylim(0,3) +
  theme(legend.position = "bottom")
```

```{r find-cell-type-factors, eval=FALSE}
i <- which(sample_info$annotation_level1 == "CD8aa")
j <- which(sample_info$annotation_level1 != "CD8aa")
prop_in  <- colMeans(L[i,])
prop_out <- colMeans(L[j,])
plot(prop_in,prop_out,pch = 20)
which(prop_in > 0.02 & prop_out < 0.01)
```

Next let's look for factors that correspond well with the organs.

```{r identify-organ-factors-1, fig.height=2.5, fig.width=2.5}
organ <- sample_info$organ_simplified
n <- nlevels(organ)
k <- ncol(L)
B <- matrix(0,k,n)
rownames(B) <- colnames(L)
colnames(B) <- levels(organ)
for (i in levels(organ)) {
  j <- which(organ == i)
  B[,i] <- apply(L[j,],2,median)
}
y <- apply(B,1,function (x) diff(range(x)))
k_set <- order(y,decreasing = TRUE)
qplot(1:k,y[k_set]) +
  labs(x = "rank",y = "value") +
  theme_cowplot(font_size = 12)
```

These are the factors that show the largest differences:

```{r identify-organ-factors-2}
y[y > 0.1]
```

Here's the Structure plot by organ:

```{r structure-plot-by-organ, fig.height=2.75, fig.width=7.25, results="hide", message=FALSE}
set.seed(1)
organ_factors <- c("k3","k11","k22","k29","k30","k62","k63","k100",
                   "k101","k174")
organ <- sample_info$organ_simplified
cells <- sort(c(sample(which(organ == "spleen"),6e4),
                sample(which(organ == "LN"),6e4),
                which(organ != "spleen" & organ != "LN")))
structure_plot(L[cells,organ_factors],gap = 10,n = 2000,
                 grouping = organ[cells]) +
  labs(y = "membership",color = "",fill = "") +
  ylim(0,3) +
  theme(legend.position = "bottom")
```

TO DO:

+ Create cell-type-specific Structure plots in which the cells
  are grouped by organ.

+ Create cell-type-specific Structure plots in which the cells
  are grouped by the "level 2" annotation (I'll call these "minor
  cell types").

*D.Z.:* There should be a gene program specific for the "resting"
clusters (CD8_cl1/2, Treg_cl1/7, CD4_cl1/2) with typically
Sell/Ccr7/Lef1/Satb1. Lots of cells, clear pattern that has been known
for a long time.

```{r identify-resting-factors, fig.height=2.5, fig.width=2.5}
target_clusters <- c("CD8_cl1","CD8_cl2","Treg_cl1","Treg_cl7",
                     "CD4_cl1","CD4_cl2")
i <- which(is.element(sample_info$annotation_level2,target_clusters))
j <- which(!is.element(sample_info$annotation_level2,target_clusters))
med_in  <- apply(L[i,],2,median)
med_out <- apply(L[j,],2,median)
y <- med_in - med_out
k_set <- order(y,decreasing = TRUE)
y[y > 0.005]
qplot(1:k,y[k_set]) +
  labs(x = "rank",y = "value") +
  theme_cowplot(font_size = 12)
```

Here's the Structure plot for visualizing some of the candidate
"resting" factors:

```{r structure-plot-resting, fig.height=4.5, fig.width=7.25, results="hide", message=FALSE}
set.seed(1)
target_factors <- c("k11","k41","k43")
cells <- which(is.element(sample_info$annotation_level2,target_clusters))
p1 <- structure_plot(L[cells,target_factors],gap = 10,n = 2000,
                 grouping = sample_info[cells,"annotation_level2"],
				 topics = 1:3) +
  labs(y = "membership",color = "",fill = "") +
  theme(legend.position = "bottom") +
  ylim(0,2)
cells <- which(cell_type == "CD4" | cell_type == "CD8")
cells <- sample(cells,1e5)
cells <- sort(c(cells,which(cell_type != "CD4" & cell_type != "CD8")))
p2 <- structure_plot(L[cells,target_factors],gap = 10,n = 2000,
                     grouping = cell_type[cells],topics = 1:3) +
  labs(y = "membership",color = "",fill = "") +
  theme(legend.position = "bottom") +
  ylim(0,2)
plot_grid(p1,p2,nrow = 2,ncol = 1)
```
