---
title: NMF analysis of the T cell data with 20 factors
author: Peter Carbonetto
output: workflowr::wflow_html
---

Here I'm exploring whether a *non-negative matrix factorization* (NMF)
of the shifted log counts is a good model for these data. While I do
not expect that $K = 20$ factors will be sufficient to capture all the
interesting structure, it should reveal the primary structure.  One
other advantage of using fewer factors is that some of the interesting
broader-scale structure could be obscured if it is split up into many
detailed factors.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load the packages needed for this analysis:

```{r load-pkgs, message=FALSE}
library(Matrix)
library(fastTopics)
library(flashier)
library(readr)
library(ggplot2)
library(cowplot)
library(pathways)
library(singlecelljamboreeR)
```

Load the meta-data, the semi-NMF results, and the NMF results:

```{r load-data}
load("../data/flashier_snmf_ldf.RData")
load("../outputs/fl_nmf_k=20.RData")
L <- fl_nmf_ldf$L
F <- fl_nmf_ldf$F
k <- ncol(L)
colnames(L) <- paste0("k",1:20)
colnames(F) <- paste0("k",1:20)
rm(fl_nmf_ldf)
```

Add a couple columns to the meta-data table to make it easier to
perform the analyses below:

```{r prepare-meta-data}
sample_info <-
  transform(sample_info,
    miniverse = grepl("miniverse",sample_info$annotation_level2,fixed = TRUE),
    prolif    = grepl("prolif",sample_info$annotation_level2,fixed = TRUE))
```

```{r gene-sets}
data(gene_sets_mouse)
gene_sets     <- gene_sets_mouse$gene_sets
gene_info     <- gene_sets_mouse$gene_info
gene_set_info <- gene_sets_mouse$gene_set_info
j <- which(with(gene_sets_mouse$gene_set_info,
                (database == "MSigDB-C2" &
                 grepl("CP",sub_category_code,fixed = TRUE)) |
                (database == "MSigDB-C5") &
                 grepl("GO",sub_category_code,fixed = TRUE)))
genes <- sort(intersect(rownames(F),gene_info$Symbol))
i1 <- match(genes,gene_info$Symbol)
i2 <- match(genes,rownames(F))
gene_info     <- gene_info[i1,]
gene_set_info <- gene_set_info[j,]
gene_sets     <- gene_sets[i1,j]
F             <- F[i2,]
rownames(gene_sets)     <- gene_info$Symbol
rownames(gene_set_info) <- gene_set_info$id
```

TO DO:

+ Remove "miniverse" cells from the Structure plots below.

+ Remove "proliferative" cells from the Structure plots below.

+ Start running GSEA on some of the factors.

These are the colors I will use throughout (I try to be consistent
with the use of the colors across the different Structure plots):

```{r colors}
factor_colors <- c("#66a61e","limegreen","#7570b3","#d95f02","#fdbf6f",
                   "#1f78b4","#e31a1c","magenta","#b2df8a","#66c2a5",
				   "#a6cee3","#e6ab02","#a6761d","#1b9e77","#ff7f00",
				   "gold","cyan","#8da0cb","#33a02c","#e7298a")
names(factor_colors) <- paste0("k",1:20)
```

Here I define a function that is used in my exploratory analyses below
to identify factors that might capture the categories of interest
(cell types, organs, conditions, etc). This is done simply by
computing the median factor membership among the cells in each group,
then ranking the factors that exhibit the largest differences in the
median factor membership between two of the groups. (I use the median
instead of the mean because the median is less sensitive to outliers.)

```{r rank-effects-function}
rank_effects_by_group_contrasts <- function (effects_matrix, grouping) {
  n <- nlevels(grouping)
  k <- ncol(effects_matrix)
  B <- matrix(0,k,n)
  rownames(B) <- colnames(effects_matrix)
  colnames(B) <- levels(grouping)
  for (i in levels(grouping)) {
    j     <- which(grouping == i)
    B[,i] <- apply(effects_matrix[j,],2,median)
  }
  return(apply(B,1,function (x) diff(range(x))))
}
```

## Cell-type factors {#factors_cell_type}

This is my attempt at identifying factors that might capture the major
cell types ("level 1" annotation):

```{r identify-cell-type-factors-1, fig.height=2.25, fig.width=2.5, warning=FALSE}
cell_type <- sample_info$annotation_level1
y <- rank_effects_by_group_contrasts(L,cell_type)
k_set <- order(y,decreasing = TRUE)
qplot(1:k,y[k_set]) +
  labs(x = "rank",y = "value") +
  theme_cowplot(font_size = 12)
```

These are the factors that show the largest differences and are
therefore promising candidates for capturing the major cell types:

```{r identify-cell-type-factors-2}
y[y > 0.05]
```

Now create a Structure plot with the cells grouped by major cell-type
(the "level 1" annotation):

```{r structure-plot-by-cell-type, fig.height=2, fig.width=7.25, results="hide", message=FALSE}
set.seed(1)
cell_type_factors <- c("k1","k6","k14","k15","k18","k3","k5")
cell_type <- sample_info$annotation_level1
cells <- which(cell_type == "CD4" | cell_type == "CD8")
cells <- sample(cells,1e5)
cells <- sort(c(cells,which(cell_type != "CD4" & cell_type != "CD8")))
p1 <- structure_plot(L[cells,cell_type_factors],gap = 10,n = 2000,
                     topics = cell_type_factors,
		     	     colors = factor_colors[cell_type_factors],
                     grouping = cell_type[cells]) +
  labs(y = "membership",color = "",fill = "") +
  theme(legend.position = "bottom")
p1
```

It is quite striking to me that the provided cluster labels do not
correspond very well to the NMF factor memberships, which suggests to
me that these "major cell types" not very accurate, and could be
misleading.

## Factors capturing organ effects {#factors_organ}

Next let's look for factors that correspond well with the organs.

```{r identify-organ-factors-1, fig.height=2.25, fig.width=2.5}
organ <- sample_info$organ_simplified
y <- rank_effects_by_group_contrasts(L,organ)
k_set <- order(y,decreasing = TRUE)
qplot(1:k,y[k_set]) +
  labs(x = "rank",y = "value") +
  theme_cowplot(font_size = 12)
```

These are the factors that show the largest differences and therefore
are promising factors for capturing "organ effects":

```{r identify-organ-factors-2}
y[y > 0.05]
```

Here is the Structure plot by organ:

```{r structure-plot-by-organ, fig.height=2.25, fig.width=7.25, results="hide", message=FALSE}
set.seed(1)
organ_factors <- c("k3","k5","k15")
organ <- sample_info$organ_simplified
cells <- sort(c(sample(which(organ == "spleen"),6e4),
                sample(which(organ == "LN"),6e4),
                which(organ != "spleen" & organ != "LN")))
L1 <- L[cells,organ_factors]
L1 <- cbind(k0 = 0.01,L1)
p2 <- structure_plot(L1,gap = 10,n = 2000,grouping = organ[cells],topics = 4:1,
                     colors = c("black",factor_colors[organ_factors])) +
  labs(y = "membership",color = "",fill = "") +
  theme(legend.position = "bottom")
p2
```

TO DO: Improve formatting using... kabble?

```{r gsea-k15}
cell_type_factors <- setdiff(cell_type_factors,"k1")
F1 <- F[,cell_type_factors]
F1 <- compute_le_effects(F1)
gsea_k15 <- singlecelljamboreeR::perform_gsea(F1[,"k15"],gene_sets,
                                              gene_set_info,L = 10,
                                              verbose = TRUE)
out <- gsea_k15$selected_gene_sets
out$top_genes <- sapply(out$top_genes,function (x) paste(x,collapse = " "))
out$lbf  <- round(out$lbf,digits = 6)
out$pip  <- round(out$pip,digits = 6)
out$coef <- round(out$coef,digits = 6)
write_csv(out,"gsea_nmf_k15.csv",quote = "none")
out[c("name","coef","genes")]
```

```{r}
plot(F[,"k15"],F1[,"k15"],pch = 20)
```

## Factors capturing condition effects {#factors_condition}

```{r identify-condition-factors-1, fig.height=2.25, fig.width=2.5}
cond_broad <- sample_info$condition_broad
y <- rank_effects_by_group_contrasts(L,cond_broad)
k_set <- order(y,decreasing = TRUE)
qplot(1:k,y[k_set]) +
  labs(x = "rank",y = "value") +
  theme_cowplot(font_size = 12)
```

These are the factors that show the largest differences and therefore
are promising factors for capturing "condition effects":

```{r identify-condition-factors-2}
y[y > 0.045]
```

Here is the Structure plot by condition:

```{r structure-plot-by-condition, fig.height=2.25, fig.width=7.25, results="hide", message=FALSE}
set.seed(1)
cond_factors <- c("k6","k14")
cond_broad <- sample_info$condition_broad
cells <- sort(c(sample(which(cond_broad == "healthy"),1e5),
                which(cond_broad != "healthy")))
p3 <- structure_plot(L[cells,cond_factors],gap = 10,n = 2000,
	   	  	         topics = c("k6","k14"),
                     grouping = cond_broad[cells],
			         colors = factor_colors[cond_factors]) +
  labs(y = "membership",color = "",fill = "") +
  theme(legend.position = "bottom")
p3
```

There are clearly some factors capturing some sort of effect of
immunization/virus and another capturing infection (virus, bacteria,
parasite, autoimmune), but we need to drill down to the cell types or
perhaps the more detailed conditions to better understand these
factors.

## Structure within the major cell types {#factors_within_cell_types}

Let's now look at whether there are factors that capture minor cell
types within the major cell types. Let's start with the CD4+ T cells.

```{r identify-CD4-factors-1, fig.height=2.25, fig.width=2.5}
cd4_clusters <-
  c("CD4_cl25","CD4_cl1","CD4_cl2","CD4_cl3","CD4_cl4","CD4_cl5","CD4_cl7",
    "CD4_cl9","CD4_cl10","CD4_cl12","CD4_cl13","CD4_cl14","CD4_cl15",
    "CD4_cl16","CD4_cl17","CD4_cl18","CD4_cl20","CD4_cl21","CD4_cl22",
    "CD4_cl23","CD4_cl28")
cells <- which(with(sample_info,
                    annotation_level1 == "CD4" & !prolif & !miniverse))
cluster <- sample_info$annotation_level2
cluster <- factor(cluster[cells],cd4_clusters)
L1      <- L[cells,]
y <- rank_effects_by_group_contrasts(L1,cluster)
k_set <- order(y,decreasing = TRUE)
qplot(1:k,y[k_set]) +
  labs(x = "rank",y = "value") +
  theme_cowplot(font_size = 12)
```

### CD4+ T cells {#cd4_factors}

These are the factors that might capture the different types of CD4+ T
cells:

```{r identify-CD4-factors-2}
y[y > 0.1]
```

Here is the Structure plot for the CD4+ T cells:

```{r structure-plot-CD4, fig.height=2.4, fig.width=7.25, results="hide", message=FALSE}
set.seed(1)
cd4_factors <- c("k6","k8","k14","k17","k18")
big_clusters <- c("CD4_cl1","CD4_cl2","CD4_cl3")
cells <- sort(c(sample(which(is.element(cluster,big_clusters)),3e4),
                which(!is.element(cluster,big_clusters))))
p4 <- structure_plot(L1[cells,cd4_factors],gap = 5,n = 2000,
  		             topics = cd4_factors,
			         grouping = cluster[cells],
			         colors = factor_colors[cd4_factors]) +
  labs(y = "membership",color = "",fill = "") +
  theme(legend.position = "bottom")
p4
```

There is a lot of interesting substructure to make sense of within the
CD4+ cells, but this substructure could also be shared by other cell
types, so we need to look at the other cell types as well to make
sense of this.

### CD8+ T cells {#cd8_factors}

Let's now examine the CD8+ T cells.

```{r identify-CD8-factors-1, fig.height=2.25, fig.width=2.5}
cd8_clusters <- 
  c("CD8_cl1","CD8_cl2","CD8_cl4","CD8_cl5","CD8_cl6","CD8_cl7",
    "CD8_cl10","CD8_cl11","CD8_cl12","CD8_cl13","CD8_cl14","CD8_cl15",
    "CD8_cl18","CD8_cl20","CD8_cl22","CD8_cl25","CD8_cl26","CD8_cl27",
    "CD8_cl28","CD8aa_cl1","CD8aa_cl16","CD8aa_cl20","CD8aa_cl40",
    "CD8aa_cl56")
cells <- which(with(sample_info,
                    (annotation_level1 == "CD8" |
					 annotation_level1 == "CD8aa")
				    & !prolif & !miniverse))
cluster <- sample_info$annotation_level2
cluster <- factor(cluster[cells],cd8_clusters)
L1      <- L[cells,]
y <- rank_effects_by_group_contrasts(L1,cluster)
k_set <- order(y,decreasing = TRUE)
qplot(1:k,y[k_set]) +
  labs(x = "rank",y = "value") +
  theme_cowplot(font_size = 12)
```

These are the factors that might capture the different types of CD8+ T
cells:

```{r identify-CD8-factors-2}
y[y > 0.1]
```

```{r structure-plot-CD8, fig.height=2.4, fig.width=7.25, results="hide", message=FALSE}
set.seed(1)
cd8_factors <- c("k3","k6","k8","k14","k17")
cells <- sort(c(sample(which(cluster == "CD8_cl1"),2e4),
                which(cluster != "CD8_cd1")))
p5 <- structure_plot(L1[cells,cd8_factors],gap = 5,n = 2000,
					 topics = cd8_factors,
					 grouping = cluster[cells],
				 	 colors = factor_colors[cd8_factors]) +
  labs(y = "membership",color = "",fill = "") +
  theme(legend.position = "bottom")
p5
```

It is interesting that there is a lot of shared structure amongst the
CD4+ and CD8+ T cells, but there are also some factors (GEPs) that are
are unique to each.

### Treg cells {#treg_factors}

Now let's take a closer look at the Treg cells:

```{r identify-Treg-factors-1, fig.height=2.25, fig.width=2.5}
treg_clusters <- 
  c("Treg_cl1","Treg_cl2","Treg_cl3","Treg_cl4","Treg_cl7","Treg_cl8",
    "Treg_cl6")
cells   <- which(with(sample_info,
                      annotation_level1 == "Treg" &
					  !prolif & !miniverse))
cluster <- sample_info$annotation_level2
cluster <- factor(cluster[cells],treg_clusters)
L1      <- L[cells,]
y <- rank_effects_by_group_contrasts(L1,cluster)
k_set <- order(y,decreasing = TRUE)
qplot(1:k,y[k_set]) +
  labs(x = "rank",y = "value") +
  theme_cowplot(font_size = 12)
```

These are the factors that might capture the different types of Treg
cells:

```{r identify-Treg-factors-2, eval=FALSE}
y[y > 0.05]
```

Here is the Structure plot for the Treg cells:

```{r structure-plot-Treg, fig.height=2.25, fig.width=7.25, results="hide", message=FALSE}
set.seed(1)
treg_factors <- c("k1","k6","k8","k14","k17","k18")
cells <- sort(c(sample(which(cluster == "Treg_cl1")),
                which(cluster != "Treg_cl1")))
p6 <- structure_plot(L1[cells,treg_factors],gap = 10,n = 2000,
  		             topics = treg_factors,
			         grouping = cluster[cells],
					 colors = factor_colors[treg_factors]) +
  labs(y = "membership",color = "",fill = "") +
  theme(legend.position = "bottom")
p6
```

## Gamma-delta T cells

Here we take a closer look at the $\gamma\delta$ T cells:

```{r identify-gdT-factors-1, fig.height=2.25, fig.width=2.5}
gdt_clusters <-
  c("gdT_cl7","gdT_cl5","gdT_cl9","gdT_cl10","gdT_cl26","gdT_cl43",
    "gdT_cl46","gdT_cl23","gdT_cl4","gdT_cl52","gdT_cl2","gdT_cl11",
	"gdT_cl12","gdT_cl28","gdT_cl33","gdT_cl35","gdT_cl42","gdT_cl50",
	"gdT_cl54")
cells <- which(with(sample_info,
                    annotation_level1 == "gdT" &
	  		        !prolif & !miniverse))
cluster <- sample_info$annotation_level2
cluster <- factor(cluster[cells],gdt_clusters)
L1      <- L[cells,]
y <- rank_effects_by_group_contrasts(L1,cluster)
k_set <- order(y,decreasing = TRUE)
qplot(1:k,y[k_set]) +
  labs(x = "rank",y = "value") +
  theme_cowplot(font_size = 12)
```

These are the factors that might capture the different types of
$\gamma\delta$ T cells cells:

```{r identify-gdT-factors-2, eval=FALSE}
y[y > 0.05]
```

Here is the Structure plot for the $\gamma\delta$ T cells:

```{r structure-plot-gdT, fig.height=2.25, fig.width=7.25, results="hide", message=FALSE}
set.seed(1)
gdt_factors <- c("k3","k5","k6","k8","k18")
p7 <- structure_plot(L1[,gdt_factors],gap = 10,n = 2000,
  		             topics = gdt_factors,
			         grouping = cluster,
					 colors = factor_colors[gdt_factors]) +
  labs(y = "membership",color = "",fill = "") +
  theme(legend.position = "bottom")
p7
```

## Proliferative cells {#proliferative}

A couple factors seem to be capturing "proliferative cells".

```{r prolif-box-plots, fig.height=2, fig.width=5}
pdat <- data.frame(k4     = L[,"k4"],
                   k16    = L[,"k16"],
                   prolif = sample_info$prolif)
q1 <- ggplot(pdat,aes(x = prolif,y = k4)) +
  geom_boxplot(width = 0.35) +
  theme_cowplot(font_size = 12) +
  labs(x = "proliferating cell")
q2 <- ggplot(pdat,aes(x = prolif,y = k16)) +
  geom_boxplot(width = 0.35) +
  theme_cowplot(font_size = 12) +
  labs(x = "proliferating cell")
plot_grid(q1,q2,nrow = 1,ncol = 2)		
```

Here is the Structure plot showing the membership in more detail:

```{r prolif-structure-plot, fig.height=2.25, fig.width=7.25, results="hide", message=FALSE}
set.seed(1)
prolif_factors <- c("k4","k16")
cells1 <- which(is.element(sample_info$annotation_level2,
                           c("CD4_prolif","CD8aa_prolif",
	 					     "DN_prolif","DP_prolif","Treg_prolif",
					 		 "gdT_prolif","nonconv_prolif")))
cells2 <- which(sample_info$annotation_level2 == "CD8_prolif")
cells  <- sort(c(cells1,sample(cells2,5000)))
cluster <- sample_info$annotation_level2
cluster <- factor(cluster[cells])
L1      <- L[cells,]
p8 <- structure_plot(L1[,prolif_factors],gap = 10,n = 2000,
  		             topics = prolif_factors,
			         grouping = cluster,
					 colors = factor_colors[prolif_factors]) +
  labs(y = "membership",color = "",fill = "") +
  theme(legend.position = "bottom")
p8
```

## Review of the results, and open questions {#review}

Let's put everything we have so far into one unified figure:

```{r all-structure-plot, fig.height=13, fig.width=7.25, results="hide", message=FALSE}
plot_grid(p1,p2,p3,p4,p5,p6,p7,
          nrow = 7,ncol = 1,
		  rel_heights = c(5,5,5,6,6,6,6))
```

Factor 7 seems to be a batch effect:

```{r factor-7-box-plot, fig.width=7.25, fig.height=2}
pdat <- data.frame(IGT = sample_info$IGT,
                   k7 = L[,"k7"])
ggplot(pdat,aes(x = IGT,y = k7)) +
  geom_boxplot() +
  theme_cowplot(font_size = 8) +
  theme(axis.ticks  = element_blank(),
        axis.text.x = element_text(angle = 90,hjust = 1))
```

Factors 9, 10, 11 and 19 also seems to be a batch effects:

```{r factor-9-10-11-19-box-plots, fig.width=7.25, fig.height=6}
pdat <- data.frame(IGT  = sample_info$IGT,
                    k9  = L[,"k9"],
					k10 = L[,"k10"],
					k11 = L[,"k11"],
					k19 = L[,"k19"])
q1 <- ggplot(pdat,aes(x = IGT,y = k9)) +
  geom_boxplot() +
  theme_cowplot(font_size = 8) +
  theme(axis.ticks  = element_blank(),
        axis.text.x = element_text(angle = 90,hjust = 1,vjust = 0.5))
q2 <- ggplot(pdat,aes(x = IGT,y = k10)) +
  geom_boxplot() +
  theme_cowplot(font_size = 8) +
  theme(axis.ticks  = element_blank(),
        axis.text.x = element_text(angle = 90,hjust = 1,vjust = 0.5))
q3 <- ggplot(pdat,aes(x = IGT,y = k11)) +
  geom_boxplot() +
  theme_cowplot(font_size = 8) +
  theme(axis.ticks  = element_blank(),
        axis.text.x = element_text(angle = 90,hjust = 1,vjust = 0.5))
q4 <- ggplot(pdat,aes(x = IGT,y = k19)) +
  geom_boxplot() +
  theme_cowplot(font_size = 8) +
  theme(axis.ticks  = element_blank(),
        axis.text.x = element_text(angle = 90,hjust = 1,vjust = 0.5))
plot_grid(q1,q2,q3,q4,nrow = 4,ncol = 1)
```

And possibly factor 13:

```{r factor-13-box-plots, fig.width=2, fig.height=2}
pdat <- data.frame(IGT51 = (sample_info$IGT == "IGT51"),
                   k13 = L[,"k13"])
ggplot(pdat,aes(x = IGT51,y = k13)) +
  geom_boxplot(width = 0.35) +
  theme_cowplot(font_size = 12)
```

Factor 10 may be capturing a "placenta effect" in
baseline/E12.5/E15.5/E18.5 conditions:

```{r factor-10-boxplot-organ, fig.height=2, fig.width=6}
pdat <- data.frame(organ     = sample_info$organ_simplified,
                   condition = sample_info$condition_detailed,
                   k10       = L[,"k10"])
q1 <- ggplot(pdat,aes(x = organ,y = k10)) +
    geom_boxplot() +
    theme_cowplot(font_size = 9) +
    theme(axis.ticks  = element_blank(),
          axis.text.x = element_text(angle = 90,hjust = 1,vjust = 0.5))
q2 <- ggplot(pdat,aes(x = condition,y = k10)) +
    geom_boxplot() +
    theme_cowplot(font_size = 8) +
    theme(axis.ticks  = element_blank(),
          axis.text.x = element_text(angle = 90,hjust = 1,vjust = 0.5))
plot_grid(q1,q2,nrow = 2,ncol = 1)
```

And factor 19 may be capturing an effect of spleen + mammary gland,
in baseline/3dpp conditions: 

```{r factor-19-boxplot-organ, fig.height=4, fig.width=7}
pdat <- data.frame(organ     = sample_info$organ_simplified,
                   condition = sample_info$condition_detailed,
                   k19       = L[,"k19"])
q1 <- ggplot(pdat,aes(x = organ,y = k19)) +
    geom_boxplot() +
    theme_cowplot(font_size = 9) +
    theme(axis.ticks  = element_blank(),
          axis.text.x = element_text(angle = 90,hjust = 1,vjust = 0.5))
q2 <- ggplot(pdat,aes(x = condition,y = k19)) +
    geom_boxplot() +
    theme_cowplot(font_size = 8) +
    theme(axis.ticks  = element_blank(),
          axis.text.x = element_text(angle = 90,hjust = 1,vjust = 0.5))
plot_grid(q1,q2,nrow = 2,ncol = 1)
```

Factor 2 seems to capture quite well the "miniverse" effect (whatever
that is?):

```{r factor-2-box-plot, fig.width=2.5, fig.height=2}
pdat <- data.frame(k2 = L[,"k2"],
                   miniverse = sample_info$miniverse)
ggplot(pdat,aes(x = miniverse,y = k2)) +
  geom_boxplot(width = 0.35) +
  theme_cowplot(font_size = 12)
```

*What is factor 2?*

The hunt for the "resting GEP"...

```{r resting, fig.height=2, fig.width=2.25}
genes <- c("Sell","Ccr7","Lef1","Satb1","Tcf7")
resting_factors <- c("k2","k8","k17")
effects_heatmap(F[genes,resting_factors])
```

## Miscellaneous notes {#misc}

It is interesting that many of the NMF factors correlate quite
strongly with individual factors in the semi-NMF result:

```{r nmf-vs-snmf-1, fig.height=2, fig.width=4.5}
y <- rep(0,20)
names(y) <- paste0("k",1:20)
for (i in 1:20) {
  x <- drop(cor(L[,i],fl_snmf_ldf$L))
  names(x) <- paste0("k",1:200)
  y[i] <- max(abs(x))
}
y <- sort(y,decreasing = TRUE)
pdat <- data.frame(k = names(y),cor = y,stringsAsFactors = FALSE)
pdat <- transform(pdat,k = factor(k,k))
ggplot(pdat,aes(x = k,y = cor)) +
  geom_point() +
  scale_y_continuous(breaks = seq(0,1,0.1)) +
  labs(x = "",y = "correlation") +
  theme_cowplot(font_size = 9)
```

In particular, the majority of the NMF factors have quite good
correlation with one of the semi-NMF factors, and almost all are at
least modestly correlated (>50%) with at least one of the semi-NMF
factors. This suggests that this NMF model with $K = 20$ factors is
doing a reasonably good job capturing the primary structure.

```{r nmf-vs-snmf-2}
sum(y > 0.98)
sum(y > 0.75)
sum(y > 0.5)
```

Complete or emerging interpretation:

1. "Baseline" factor. (56% correlation with factor 15 in the semi-NMF
   k=200 fit.)

2. "Miniverse" effect (76% correlated with factor 2 in the semi-NMF
   k=200 fit).

3. CD8aa, gdT and to a lesser extent DN cells, specific to spleen, LN,
   skin and lung, and to a subset of these conditions: KbxN, Mfurfur,
   MTB, HDM_secondary, saureus. (Also is 80% correlated with factor 3
   in the semi-NMF k=200 fit.)

4. Structure within proliferative cells (93% correlated with factor 20
   in the semi-NMF k=200 fit).
   
5. Skin-specific gdT cells. Very strongly correlated (99%) with factor
   8 in the semi-NMF k=200 fit. Corresponds very closely with cluster
   gdT_cl9.

7. Batch effect (IGT13, IGT14). Also note that it is 94%
   correlated with factor 9 from the semi-NMF k=200 fit.

9. Batch effect (IGT9), in "baseline" condition only. Very strongly
   correlated (99%) with factor 24 in the semi-NMF k=200 fit.

10. Placenta in baseline/E12.5/E15.5/E18.5 (98% correlation with
    factor 19 in the semi-NMF k=200 fit).

11. Batch effect, in "baseline" condition only (94% correlatation with
    factor 39 in the semi-NMF k=200 fit).

13. Batch effect (IGT51).

15. Thymocytes (95% correlation with factor 7 in the semi-NMF
    k=200 fit).

16. Structure within proliferative cells (77% correlated with factor 5
    in the semi-NMF k=200 fit).

18. Strongly correlated (93%) with factor 13 in the semi-NMF k=200
    fit. Active in a subset of conditions (PDAC, MTB, EXX.5, etc),
    mainly in gdT and nonconv cells.

19. Spleen and mammary gland in baseline/3dpp (>99.9% correlation with
    factor 37 in the semi-NMF k=200 fit).

In-progress interpretation:

6. Some CD8+ cells, but maybe active in only some organs and/or conditions?
Modestly correlated (57%) with factor 10 in the semi-NMF k=200 fit.

14. T-reg cells (mostly), but maybe active in only some organs and/or
    conditions? Quite well correlated (76%) with factor 12 in the
    semi-NMF k=200 fit.

Putative "resting" GEPs:

8. *Notes go here.*

17. *Notes go here.*

Other factors:

Elusive and rare factors:

```{r elusive-factors}
sum(L[,12] > 0.5)
sum(L[,20] > 0.5)
```

12. Very strongly correlated (>99.9%) with factor 33 in the semi-NMF
    k=200 fit.

20. Quite well correlalted (84%) with factor 40 in the semi-NMF k=200
    fit.
