---
title: "Extended Data Table 2: Gene signature per GP"
author: "Ziang Zhang"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

Produced by
[`script/ExtendedDataTable2_gene_signature_matrix.R`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/script/ExtendedDataTable2_gene_signature_matrix.R):

```{r edt2-code, code=readLines("../script/ExtendedDataTable2_gene_signature_matrix.R"), eval=FALSE}
```

## Table {#edt2}

One row per signature gene: sort by clicking a column header, or use the
search boxes under each header to filter (e.g. type `GP1_up` into the
**signature_name** box, or a gene symbol into **gene_symbol**). Download the
workbook at
[`figures/final-selected/ExtendedDataTable2_gene_signature_matrix.xlsx`](https://github.com/AgueroZZ/immgenT-GP-analysis/raw/main/figures/final-selected/ExtendedDataTable2_gene_signature_matrix.xlsx).

```{r edt2-preview, echo=FALSE, message=FALSE}
suppressPackageStartupMessages(library(DT))
source("../code/R/table_xlsx.R")
tab <- read_table_xlsx("../figures/final-selected/ExtendedDataTable2_gene_signature_matrix.xlsx")
DT::datatable(
  tab,
  rownames = FALSE,
  filter = "top",
  options = list(
    pageLength = 25,
    lengthMenu = list(c(25, 50, 100, 200, -1), c("25", "50", "100", "200", "All")),
    scrollX = TRUE
  )
) |> DT::formatRound(columns = "score", digits = 3)
```

::: {.figcaption}
**Extended Data Table 2.** Signature genes of all 200 GPs, defined as genes
with score > 0.1 (up) and < -0.1 (down) and capped at the 100 highest-scoring
genes per GP and direction.
:::
