---
title: "Extended Data Table 8 (wide format): comprehensive gene signature matrix"
author: "Ziang Zhang"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

Alternate, per-GP-column layout of
[Extended Data Table 8](ExtendedDataTable8.html); produced by
[`script/ExtendedDataTable8_gene_signature_matrix_wide.R`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/script/ExtendedDataTable8_gene_signature_matrix_wide.R):

```{r edt8wide-code, code=readLines("../script/ExtendedDataTable8_gene_signature_matrix_wide.R"), eval=FALSE}
```

## Table {#edt8wide}

This table is a per-GP grouping (200 GPs x 3 columns each: Gene, Direction,
Score) rather than one row per GP, so it's distributed as an .xlsx with a
merged GP header rather than a flat CSV. Pick a GP below to browse its full
signature gene list (up to 202 rows: up to 100 up- and 100 down-regulated,
plus a truncation note if either was capped -- see below); sort by clicking a
column header, or use the search boxes under **Gene** / **Direction** to
filter further. Download the full spreadsheet at
[`figures/generated/ExtendedDataTable8_gene_signature_matrix_wide.xlsx`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/figures/generated/ExtendedDataTable8_gene_signature_matrix_wide.xlsx).

```{r edt8wide-preview, echo=FALSE, message=FALSE}
suppressPackageStartupMessages({
  library(DT)
  library(crosstalk)
})
source("../code/R/gene_signature_helpers.R")

F_pm_filtered <- readRDS("../data/F_pm_filtered.rds")
F_pm_filtered <- apply(F_pm_filtered, 2, function(x) x / max(abs(x)))
n_gp <- ncol(F_pm_filtered)
gp_labels <- paste0("GP", seq_len(n_gp))

gp_blocks <- build_gp_gene_signature_blocks(F_pm_filtered, cutoff = 0.1, cap = 100)
long_df <- do.call(rbind, Map(function(gp, block) {
  cbind(GP = gp, block, stringsAsFactors = FALSE)
}, gp_labels, gp_blocks))
rownames(long_df) <- NULL

shared <- SharedData$new(long_df, ~GP)
bscols(
  widths = c(3, 9),
  filter_select("edt8wide-gp-filter", "GP", shared, ~GP),
  DT::datatable(
    shared,
    rownames = FALSE,
    filter = "top",
    options = list(
      pageLength = 25,
      lengthMenu = list(c(25, 50, 100, 200, -1), c("25", "50", "100", "200", "All")),
      scrollX = TRUE
    )
  ) |> DT::formatRound(columns = "Score", digits = 3)
)
```

::: {.figcaption}
**Extended Data Table 8 (wide format).** The comprehensive signature gene
list for every GP, one 3-column block per GP (Gene, Direction, Score),
listing every gene with |score| > 0.1 on the same max|.|=1-per-GP-scaled
gene factor matrix Extended Data Table 1 uses, ranked by |score| within
direction (up-regulated genes first, then down-regulated), and capped at the
top 100 up- and top 100 down-regulated genes per GP. When a direction has
more than 100 qualifying genes, one extra row immediately follows its
(capped) gene rows -- e.g. row 101 after 100 up-regulated genes -- noting how
many more were left out (Score blank for that row). Table 1's **Top Genes
+/-** columns are this table's top 5 per direction; this table is the full
list. Different GPs have different signature-gene counts, so shorter blocks
are blank-padded in the downloadable spreadsheet. See
[Extended Data Table 8](ExtendedDataTable8.html) for the same data as a tidy
long table (one row per gene).
:::
