---
title: "Extended Data Table 5: GP during activation"
author: "Ziang Zhang"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

Produced by
[`script/ExtendedDataTable5_GP_during_activation.R`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/script/ExtendedDataTable5_GP_during_activation.R):

```{r edt5-code, code=readLines("../script/ExtendedDataTable5_GP_during_activation.R"), eval=FALSE}
```

## Table {#edt5}

All 200 rows are browsable below: sort by clicking a column header, type a GP
name into the search box under the **GP** column, or use a per-column box to
filter by value range. Download the raw CSV at
[`figures/generated/ExtendedDataTable5_GP_during_activation.csv`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/figures/generated/ExtendedDataTable5_GP_during_activation.csv).

```{r edt5-preview, echo=FALSE, message=FALSE}
suppressPackageStartupMessages(library(DT))
tab <- read.csv("../figures/generated/ExtendedDataTable5_GP_during_activation.csv", check.names = FALSE)
DT::datatable(
  tab,
  rownames = FALSE,
  filter = "top",
  options = list(
    pageLength = 10,
    lengthMenu = list(c(10, 25, 50, 100, -1), c("10", "25", "50", "100", "All")),
    scrollX = TRUE
  )
) |> DT::formatRound(columns = colnames(tab)[-1], digits = 4)
```

::: {.figcaption}
**Extended Data Table 5.** One row per GP describing how its loading changes with
T-cell activation, computed separately in CD4 and CD8: `mean_change_loadings`
(activated minus resting mean loading), `AveExpr` (mean loading over
activated + resting cells), `d` (standardized mean difference = change / pooled
loading SD, the statistic plotted in Figure 3a), and `Ratio_CD8_CD4` (ratio of
the CD8 to CD4 mean loading change). This reproduces the `GP_activation_summary`
of Figure 3a; note that the `d_CD4` / `d_CD8` columns replace the retired
limma-style `z_CD4` / `z_CD8` of the earlier table version.
:::
