---
title: "Extended Data Table 1: summary of GP characteristics and annotations"
author: "Ziang Zhang"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

Produced by
[`script/ExtendedDataTable1_GP_summary.R`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/script/ExtendedDataTable1_GP_summary.R):

```{r edt1-code, code=readLines("../script/ExtendedDataTable1_GP_summary.R"), eval=FALSE}
```

## Table {#edt1}

Showing the first 20 rows (of 200); download the full table at
[`figures/generated/ExtendedDataTable1_GP_summary.csv`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/figures/generated/ExtendedDataTable1_GP_summary.csv).

```{r edt1-preview, echo=FALSE, message=FALSE}
tab <- read.csv("../figures/generated/ExtendedDataTable1_GP_summary.csv",
                check.names = FALSE, stringsAsFactors = FALSE, colClasses = "character")
knitr::kable(head(tab, 20), row.names = FALSE)
```

::: {.figcaption}
**Extended Data Table 1.** One row per GP. **Lineage**, **Cluster** and **Tissue**
list the categories (annotation_level1, annotation_level2, organ_simplified) that
the GP *positively* predicts well -- one-vs-rest AUC > 0.8 and the optimal
decision threshold at or above the GP's median loading, so high (not low) loading
drives the prediction -- computed on healthy non-thymocyte cells (the same
population as Figure 2 / Figure 4), with the median loading taken on that same
population. **Signature genes / proteins** give the top 5 up- and top 5
down-regulated genes and proteins by factor score, among those with |score| > 0.1
on the max|.|=1-per-GP-scaled gene and protein factor matrices.
:::
