---
title: "Extended Data Table 6: protein factor matrix"
author: "Ziang Zhang"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

Produced by
[`script/ExtendedDataTable6_protein_factor_matrix.R`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/script/ExtendedDataTable6_protein_factor_matrix.R):

```{r edt6-code, code=readLines("../script/ExtendedDataTable6_protein_factor_matrix.R"), eval=FALSE}
```

## Table {#edt6}

All 200 rows and every protein column are browsable below: sort by clicking a
column header, type a GP name into the search box under the **GP** column, or
use a per-protein box to filter by score range. Scroll horizontally to see all
protein columns. Download the raw CSV at
[`figures/generated/ExtendedDataTable6_protein_factor_matrix.csv`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/figures/generated/ExtendedDataTable6_protein_factor_matrix.csv).

```{r edt6-preview, echo=FALSE, message=FALSE}
suppressPackageStartupMessages(library(DT))
tab <- read.csv("../figures/generated/ExtendedDataTable6_protein_factor_matrix.csv", check.names = FALSE)
DT::datatable(
  tab,
  rownames = FALSE,
  filter = "top",
  options = list(
    pageLength = 10,
    lengthMenu = list(c(10, 25, 50, 100, -1), c("10", "25", "50", "100", "All")),
    scrollX = TRUE
  )
) |> DT::formatRound(columns = colnames(tab)[-1], digits = 3)
```

::: {.figcaption}
**Extended Data Table 6.** The final processed CITE-seq protein factor matrix used
by Figure 6 (panel 6b), as a plain table: one row per GP, one column per protein,
holding the protein's factor score in that GP. Built from the projected protein
EBMF fit, restricted to the selected "good" proteins (isotype / THY1.1 / excluded
proteins removed) and scaled so each GP column has maximum absolute score 1.
:::
