---
title: "Extended Data Table 1: summary of GP characteristics and annotations"
author: "Ziang Zhang"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

Produced by
[`script/ExtendedDataTable1_GP_summary.R`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/script/ExtendedDataTable1_GP_summary.R):

```{r edt1-code, code=readLines("../script/ExtendedDataTable1_GP_summary.R"), eval=FALSE}
```

## Table {#edt1}

All 200 rows are browsable below: sort by clicking a column header, use the
search boxes under each header to filter (e.g. type a GP name, or a lineage
like `CD8`), or use the box in the top-right corner for a free-text search
across every column. Download the raw CSV at
[`figures/generated/ExtendedDataTable1_GP_summary.csv`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/figures/generated/ExtendedDataTable1_GP_summary.csv).

```{r edt1-preview, echo=FALSE, message=FALSE}
suppressPackageStartupMessages(library(DT))
tab <- read.csv("../figures/generated/ExtendedDataTable1_GP_summary.csv",
                check.names = FALSE, stringsAsFactors = FALSE, colClasses = "character")
DT::datatable(
  tab,
  rownames = FALSE,
  filter = "top",
  options = list(
    pageLength = 10,
    lengthMenu = list(c(10, 25, 50, 100, -1), c("10", "25", "50", "100", "All")),
    scrollX = TRUE
  )
)
```

::: {.figcaption}
**Extended Data Table 1.** One row per GP. **Lineage**, **Cluster** and **Tissue**
list the categories (annotation_level1, annotation_level2, organ_simplified) that
the GP *positively* predicts well -- one-vs-rest AUC > 0.8 and the optimal
decision threshold at or above the GP's median loading, so high (not low) loading
drives the prediction -- computed on healthy non-thymocyte cells (the same
population as Figure 3 / Figure 5), with the median loading taken on that same
population. **Top Genes +/-** and **Top Proteins +/-** give the top 5 up- and
top 5 down-regulated genes and proteins by factor score, among those with
|score| > 0.1 on the max|.|=1-per-GP-scaled gene and protein factor matrices.
:::
