---
title: "Extended Data Table 1: summary of GP characteristics and comments"
author: "Ziang Zhang"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

Produced by
[`script/ExtendedDataTable1_GP_summary.R`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/script/ExtendedDataTable1_GP_summary.R):

```{r edt1-code, code=readLines("../script/ExtendedDataTable1_GP_summary.R"), eval=FALSE}
```

## Table {#edt1}

All 200 rows are browsable below: sort by clicking a column header, use the
search boxes under each header to filter (e.g. type a GP name, or a keyword
like `proliferation` under **Comments**), or use the box in the top-right
corner for a free-text search across every column. Download the raw CSV at
[`figures/final-selected/ExtendedDataTable1_GP_summary.csv`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/figures/final-selected/ExtendedDataTable1_GP_summary.csv).

```{r edt1-preview, echo=FALSE, message=FALSE}
suppressPackageStartupMessages(library(DT))
tab <- read.csv("../figures/final-selected/ExtendedDataTable1_GP_summary.csv",
                check.names = FALSE, stringsAsFactors = FALSE)
DT::datatable(
  tab,
  rownames = FALSE,
  filter = "top",
  options = list(
    pageLength = 10,
    lengthMenu = list(c(10, 25, 50, 100, -1), c("10", "25", "50", "100", "All")),
    scrollX = TRUE
  )
) |> DT::formatRound(columns = "Prop. active cells", digits = 4)
```

::: {.figcaption}
**Extended Data Table 1.** One row per GP. **Top Genes +/-** give the top 15
up- and top 15 down-regulated genes by factor score, among those with
|score| > 0.1 on the max|.|=1-per-GP-scaled gene factor matrix; GPs with fewer
than 15 qualifying genes in a direction list all of them, and the full lists
are [Extended Data Table 2](ExtendedDataTable2.html). **Comments** is a manual,
human-written description of the GP, blank where a GP has no comment.
**Prop. active cells** is the fraction of non-thymocyte cells whose loading
exceeds 0.1 after scaling the GP's loading column by its maximum (the quantity
histogrammed in Figure 2b), and **N active genes** is the number of genes with
|score| > 0.25 on the max|.|=1-per-GP-scaled gene factor matrix (the quantity
histogrammed in Figure 2c) -- a stricter cutoff than the 0.1 used for the
signature-gene columns.
:::
