---
title: "Extended Data Table 7: protein gating"
author: "Ziang Zhang"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

Produced by
[`script/ExtendedDataTable7_protein_gating.R`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/script/ExtendedDataTable7_protein_gating.R):

```{r edt7-code, code=readLines("../script/ExtendedDataTable7_protein_gating.R"), eval=FALSE}
```

## Table {#edt7}

Showing the first 20 rows (of 200), ordered well-aligned GPs first; download the
full table at
[`figures/generated/ExtendedDataTable7_protein_gating.csv`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/figures/generated/ExtendedDataTable7_protein_gating.csv).

```{r edt7-preview, echo=FALSE, message=FALSE}
tab <- read.csv("../figures/generated/ExtendedDataTable7_protein_gating.csv",
                check.names = FALSE, stringsAsFactors = FALSE)
knitr::kable(head(tab, 20), row.names = FALSE)
```

::: {.figcaption}
**Extended Data Table 7.** One row per GP: its curated positive / negative protein
marker signature, and how well a protein gate built from that signature recovers
cells with high GP loading. For each GP, cells are gated by its markers (positive
markers above their threshold, negative at/below); we report the proportion of
gated cells positive for the GP (loading > 0.1; the True Discovery Proportion)
versus that proportion among all cells, and a **Well aligned** flag. The flag is
the formula-positive set (TDP >= 0.25 and greater than the all-cell proportion)
hand-trimmed to those whose protein and loading gates also agree visually on the
MDE embedding -- reproducing the published table's curated 25-GP set.
:::
