---
title: "Extended Data Table 4: GP activity during activation"
author: "Ziang Zhang"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

Produced by
[`script/ExtendedDataTable4_GP_during_activation.R`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/script/ExtendedDataTable4_GP_during_activation.R):

```{r edt4-code, code=readLines("../script/ExtendedDataTable4_GP_during_activation.R"), eval=FALSE}
```

## Table {#edt4}

All 200 rows are browsable below: sort by clicking a column header, type a GP
name into the search box under the **GP** column, or use a per-column box to
filter by value range. Download the workbook at
[`figures/final-selected/ExtendedDataTable4_GP_during_activation.xlsx`](https://github.com/AgueroZZ/immgenT-GP-analysis/raw/main/figures/final-selected/ExtendedDataTable4_GP_during_activation.xlsx).

```{r edt4-preview, echo=FALSE, message=FALSE}
suppressPackageStartupMessages(library(DT))
source("../code/R/table_xlsx.R")
tab <- read_table_xlsx("../figures/final-selected/ExtendedDataTable4_GP_during_activation.xlsx")
DT::datatable(
  tab,
  rownames = FALSE,
  filter = "top",
  options = list(
    pageLength = 10,
    lengthMenu = list(c(10, 25, 50, 100, -1), c("10", "25", "50", "100", "All")),
    scrollX = TRUE
  )
) |> DT::formatRound(columns = colnames(tab)[-1], digits = 4)
```

::: {.figcaption}
**Extended Data Table 4.** Change in GP loading with T cell activation,
computed in CD4 and CD8 cells separately: the mean loading in activated minus
resting cells; d_CD4 and d_CD8, the corresponding standardized differences,
each divided by the s.d. of the GP's loading across all activated and resting
CD4 and CD8 cells; and Ratio_CD8_CD4, the ratio of the CD8 to the CD4 mean
loading change.
:::
