---
title: "Extended Data Table 1: summary of GP characteristics and annotations"
author: "Ziang Zhang"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

Produced by
[`script/ExtendedDataTable1_GP_summary.R`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/script/ExtendedDataTable1_GP_summary.R):

```{r edt1-code, code=readLines("../script/ExtendedDataTable1_GP_summary.R"), eval=FALSE}
```

## Table {#edt1 .tabset}

Two versions of the same table are available below -- click a tab to switch.
They differ only in how many features the signature columns list; everything
else, including the manual **Annotation** column, is identical.

In either version all 200 rows are browsable: sort by clicking a column header,
use the search boxes under each header to filter (e.g. type a GP name, a
lineage like `CD8`, or an annotation keyword like `batch`), or use the box in
the top-right corner for a free-text search across every column.

```{r edt1-setup, echo=FALSE, message=FALSE}
suppressPackageStartupMessages(library(DT))
edt1_table <- function(path) {
  tab <- read.csv(path, check.names = FALSE, stringsAsFactors = FALSE,
                  colClasses = "character")
  DT::datatable(
    tab,
    rownames = FALSE,
    filter = "top",
    options = list(
      pageLength = 10,
      lengthMenu = list(c(10, 25, 50, 100, -1), c("10", "25", "50", "100", "All")),
      scrollX = TRUE
    )
  )
}
```

### Main (top 5)

The published version: **Top Genes +/-** and **Top Proteins +/-** list the top
5 features per direction. Download the raw CSV at
[`figures/final-selected/ExtendedDataTable1_GP_summary.csv`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/figures/final-selected/ExtendedDataTable1_GP_summary.csv).

```{r edt1-preview, echo=FALSE, message=FALSE}
edt1_table("../figures/final-selected/ExtendedDataTable1_GP_summary.csv")
```

### Longer internal (top 15)

The same table with up to 15 features per direction instead of 5 -- same
|score| > 0.1 cutoff and same ranking, so each cell above is exactly the first
5 entries of the cell here. GPs with fewer than 15 qualifying features list all
of them. Download the raw CSV at
[`figures/final-selected/ExtendedDataTable1_GP_summary_top15.csv`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/figures/final-selected/ExtendedDataTable1_GP_summary_top15.csv).

```{r edt1-preview-top15, echo=FALSE, message=FALSE}
edt1_table("../figures/final-selected/ExtendedDataTable1_GP_summary_top15.csv")
```

## {-}

<script>
// A DataTable rendered inside a hidden tab measures its columns at zero width,
// so the header and body misalign the first time that tab is opened. Re-measure
// whichever tables just became visible.
$(document).on('shown.bs.tab', function () {
  if (window.jQuery && $.fn.dataTable) {
    $($.fn.dataTable.tables(true)).DataTable().columns.adjust();
  }
});
</script>

::: {.figcaption}
**Extended Data Table 1.** One row per GP. **Annotation** is the manual,
human-written description of the GP, maintained by hand in
[`curation/GP_manual_annotations.csv`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/curation/GP_manual_annotations.csv)
and blank for GPs not yet annotated; that file replaces the `annotation` column
of the retired hand-maintained `Table S1.xlsx`, and this table is now where the
annotations are published. **Lineage**, **Cluster** and **Tissue**
list the categories (annotation_level1, annotation_level2, organ_simplified) that
the GP *positively* predicts well -- one-vs-rest AUC > 0.8 and the optimal
decision threshold at or above the GP's median loading, so high (not low) loading
drives the prediction -- computed on healthy non-thymocyte cells (the same
population as Figure 3 / Figure 5), with the median loading taken on that same
population. **Top Genes +/-** and **Top Proteins +/-** give the top up- and
top down-regulated genes and proteins by factor score, among those with
|score| > 0.1 on the max|.|=1-per-GP-scaled gene and protein factor matrices --
5 per direction in the main version, 15 in the longer internal version.
:::
