---
title: "Extended Data Table 3: GP AUC by tissue"
author: "Ziang Zhang"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

Produced by
[`script/ExtendedDataTable3_GP_AUC_tissue.R`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/script/ExtendedDataTable3_GP_AUC_tissue.R):

```{r edt3-code, code=readLines("../script/ExtendedDataTable3_GP_AUC_tissue.R"), eval=FALSE}
```

## Table {#edt3}

Showing the first 20 rows (of 200) and first 10 columns; download the full table at
[`figures/generated/ExtendedDataTable3_GP_AUC_tissue.csv`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/figures/generated/ExtendedDataTable3_GP_AUC_tissue.csv).

```{r edt3-preview, echo=FALSE, message=FALSE}
tab <- read.csv("../figures/generated/ExtendedDataTable3_GP_AUC_tissue.csv", check.names = FALSE)
knitr::kable(head(tab[, seq_len(min(10, ncol(tab)))], 20), row.names = FALSE, digits = 3)
```

::: {.figcaption}
**Extended Data Table 3.** One row per GP, one column per tissue
(organ_simplified). Each value is the one-vs-rest AUC for predicting membership
in that tissue from the GP's loading, computed on healthy non-thymocyte cells.
(Preview shows the first 10 of `r ncol(read.csv("../figures/generated/ExtendedDataTable3_GP_AUC_tissue.csv", check.names = FALSE)) - 1` tissue columns.)
:::
