---
title: "Extended Data Table 5: GP AUC by tissue"
author: "Ziang Zhang"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

Produced by
[`script/ExtendedDataTable5_GP_AUC_tissue.R`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/script/ExtendedDataTable5_GP_AUC_tissue.R):

```{r edt5-code, code=readLines("../script/ExtendedDataTable5_GP_AUC_tissue.R"), eval=FALSE}
```

## Table {#edt5}

All 200 rows and every tissue column are browsable below: sort by clicking a
column header, type a GP name into the search box under the **GP** column, or
use a per-tissue box to filter by AUC range. Scroll horizontally to see all
tissue columns. Download the raw CSV at
[`figures/final-selected/ExtendedDataTable5_GP_AUC_tissue.csv`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/figures/final-selected/ExtendedDataTable5_GP_AUC_tissue.csv).

```{r edt5-preview, echo=FALSE, message=FALSE}
suppressPackageStartupMessages(library(DT))
tab <- read.csv("../figures/final-selected/ExtendedDataTable5_GP_AUC_tissue.csv", check.names = FALSE)
DT::datatable(
  tab,
  rownames = FALSE,
  filter = "top",
  options = list(
    pageLength = 10,
    lengthMenu = list(c(10, 25, 50, 100, -1), c("10", "25", "50", "100", "All")),
    scrollX = TRUE
  )
) |> DT::formatRound(columns = colnames(tab)[-1], digits = 3)
```

::: {.figcaption}
**Extended Data Table 5.** One row per GP, one column per tissue
(organ_simplified). Each value is the one-vs-rest AUC for predicting membership
in that tissue from the GP's loading, computed on healthy non-thymocyte cells.
:::
