---
title: "Extended Data Table 7: protein gating (manual protein positivity thresholds)"
author: "Ziang Zhang"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

Produced by
[`script/ExtendedDataTable7_protein_thresholds.R`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/script/ExtendedDataTable7_protein_thresholds.R):

```{r edt7-code, code=readLines("../script/ExtendedDataTable7_protein_thresholds.R"), eval=FALSE}
```

## Table {#edt7}

All 42 rows are browsable below (default order: alphabetical by protein): sort
by clicking a column header, type a protein name into the search box under the
**Protein** column, or filter on a value range under **Manual threshold**.
Download the raw CSV at
[`figures/final-selected/ExtendedDataTable7_protein_thresholds.csv`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/figures/final-selected/ExtendedDataTable7_protein_thresholds.csv).

```{r edt7-preview, echo=FALSE, message=FALSE}
suppressPackageStartupMessages(library(DT))
tab <- read.csv("../figures/final-selected/ExtendedDataTable7_protein_thresholds.csv",
                check.names = FALSE, stringsAsFactors = FALSE)
DT::datatable(
  tab,
  rownames = FALSE,
  filter = "top",
  options = list(
    pageLength = 15,
    lengthMenu = list(c(15, 25, 50, -1), c("15", "25", "50", "All")),
    scrollX = TRUE
  )
) |> DT::formatRound(columns = "Manual threshold", digits = 2)
```

::: {.figcaption}
**Extended Data Table 7.** The surface proteins carrying a manually reviewed
positivity threshold, and the hand-set value used to call a cell positive. Values
are on the log-normalized ADT scale (Seurat `LogNormalize`, `scale.factor` =
3472); a cell is protein-positive when its value is strictly above the threshold
and protein-negative when it is at or below. Thresholds were set by eye from each
protein's ADT distribution, starting from the automatic 2-component
Gaussian-mixture cutoff of pipeline step 3
([`code/pipeline/03_protein_thresholds.R`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/code/pipeline/03_protein_thresholds.R))
and rounding to the reviewed value listed here. These are the thresholds used by
every protein gate in Figure 6 and Figure S6. Proteins outside this list have no
manual threshold and are skipped when they appear in a GP marker signature
(BTLA.CD272, GR1-LY6G-LY6C1-LY6C2, LY108, SLAM.CD150, THY1.2).
:::
