---
title: "Extended Data Table 8: comprehensive gene signature matrix"
author: "Ziang Zhang"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

Produced by
[`script/ExtendedDataTable8_gene_signature_matrix_long.R`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/script/ExtendedDataTable8_gene_signature_matrix_long.R):

```{r edt8-code, code=readLines("../script/ExtendedDataTable8_gene_signature_matrix_long.R"), eval=FALSE}
```

## Table {#edt8}

One row per signature gene: sort by clicking a column header, or use the
search boxes under each header to filter (e.g. type `GP1_up` into the
**signature_name** box, or a gene symbol into **gene_symbol**). Download the
raw CSV at
[`figures/generated/ExtendedDataTable8_gene_signature_matrix_long.csv`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/figures/generated/ExtendedDataTable8_gene_signature_matrix_long.csv).
The same data is also available in a per-GP-column
[wide format](ExtendedDataTable8_wide.html).

```{r edt8-preview, echo=FALSE, message=FALSE}
suppressPackageStartupMessages(library(DT))
tab <- read.csv("../figures/generated/ExtendedDataTable8_gene_signature_matrix_long.csv", stringsAsFactors = FALSE)
DT::datatable(
  tab,
  rownames = FALSE,
  filter = "top",
  options = list(
    pageLength = 25,
    lengthMenu = list(c(25, 50, 100, 200, -1), c("25", "50", "100", "200", "All")),
    scrollX = TRUE
  )
) |> DT::formatRound(columns = "score", digits = 3)
```

::: {.figcaption}
**Extended Data Table 8.** The comprehensive signature gene list for every
GP, as a tidy long table: one row per gene, with **signature_name**
identifying the GP and direction it's a signature gene for (e.g. `GP1_up`)
and **score** its factor loading. Every gene has |score| > 0.1 on the same
max|.|=1-per-GP-scaled gene factor matrix Extended Data Table 1 uses, ranked
by |score| within each GP-direction and capped at the top 100 up- and top 100
down-regulated genes per GP. Table 1's **Top Genes +/-** columns are this
table's top 5 per direction; this table is the full list.
:::
