---
title: "Extended Data Table 7: protein gating"
author: "Ziang Zhang"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

Produced by
[`script/ExtendedDataTable7_protein_gating.R`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/script/ExtendedDataTable7_protein_gating.R):

```{r edt7-code, code=readLines("../script/ExtendedDataTable7_protein_gating.R"), eval=FALSE}
```

## Table {#edt7}

All 200 rows are browsable below (default order: well-aligned GPs first): sort
by clicking a column header, type a GP name into the search box under the
**GP** column, or use the other per-column boxes to filter (e.g. a protein
name in **Positive markers**, or a proportion range). Download the raw CSV at
[`figures/generated/ExtendedDataTable7_protein_gating.csv`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/figures/generated/ExtendedDataTable7_protein_gating.csv).

```{r edt7-preview, echo=FALSE, message=FALSE}
suppressPackageStartupMessages(library(DT))
tab <- read.csv("../figures/generated/ExtendedDataTable7_protein_gating.csv",
                check.names = FALSE, stringsAsFactors = FALSE)
DT::datatable(
  tab,
  rownames = FALSE,
  filter = "top",
  options = list(
    pageLength = 10,
    lengthMenu = list(c(10, 25, 50, 100, -1), c("10", "25", "50", "100", "All")),
    scrollX = TRUE
  )
) |> DT::formatRound(
  columns = c(
    "Prop. positive among gated (True Discovery Proportion)",
    "Prop. positive among all cells"
  ),
  digits = 3
)
```

::: {.figcaption}
**Extended Data Table 7.** One row per GP: its curated positive / negative protein
marker signature, and how well a protein gate built from that signature recovers
cells with high GP loading. For each GP, cells are gated by its markers (positive
markers above their threshold, negative at/below); we report the proportion of
gated cells positive for the GP (loading > 0.1; the True Discovery Proportion)
versus that proportion among all cells, and a **Well aligned** flag. The flag is
the formula-positive set (TDP >= 0.25 and greater than the all-cell proportion)
hand-trimmed to those whose protein and loading gates also agree visually on the
MDE embedding -- reproducing the published table's curated 25-GP set.
:::
