---
title: "Extended Data Figure 8: Cluster-level GP membership within each T cell lineage"
author: "Ziang Zhang"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r doc-code, include=FALSE}
source("../code/R/doc_code.R")
```

All seven panels are produced by
[`script/FigureS8.R`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/script/FigureS8.R).
They are the per-cluster, all-GP counterpart of
[Fig. 3B](Figure3.html#fig3b), which shows six hand-picked lineage-defining GPs
grouped by lineage: here every GP that marks a sub-lineage cluster is shown, and
cells are grouped by cluster within each lineage. The GPs are selected from the
cluster AUCs published as
[Extended Data Table 6](ExtendedDataTable6.html), and
[`script/verify_structure_plot_gps.R`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/script/verify_structure_plot_gps.R)
re-derives every panel's GP set from that published table, failing if it
disagrees with the panels or with the caption below. The code is shown for
reference (not re-executed on this page, since it loads the full cell-by-GP
loading matrix); the images are its pre-rendered output.

## Setup

```{r figs8-setup, code=code_for("../script/FigureS8.R", "doc:setup"), eval=FALSE}
```

## Panel definitions and GP selection {#figs8-panels}

The panel map, the AUC threshold and the display filters live in
`code/R/structure_plot_panels.R`, which the verification script reads as well,
so the figure and the check cannot disagree about what is shown:

```{r figs8-panels-code, code=code_for("../code/R/structure_plot_panels.R", "doc:panels"), eval=FALSE}
```

```{r figs8-selection-code, code=code_for("../script/FigureS8.R", "GP selection"), eval=FALSE}
```

## (a-g) Cluster-level GP membership by lineage {#figs8}

```{r figs8-code, code=code_for("../script/FigureS8.R", "s8a-s8g"), eval=FALSE}
```

```{r figs8-img, echo=FALSE, out.width="100%"}
knitr::include_graphics(c(
  "assets/FigureS8/s8a.png", "assets/FigureS8/s8b.png",
  "assets/FigureS8/s8c.png", "assets/FigureS8/s8d.png",
  "assets/FigureS8/s8e.png", "assets/FigureS8/s8f.png",
  "assets/FigureS8/s8g.png"
))
```

::: {.figcaption}
**Extended Data Fig. 8a-g.** Structure plots of single-cell GP activity
(membership / GP loading) in healthy non-thymocyte cells, one panel per T cell
lineage, with cells grouped by sub-lineage cluster (`annotation_level2`):
**(a) CD8, 16 GPs; (b) CD4, 22 GPs; (c) Treg, 11 GPs; (d) gdT, 44 GPs;
(e) CD8aa, 9 GPs; (f) Tz, 10 GPs; (g) DN, 17 GPs** -- 69 distinct GPs across the
seven panels. A GP is shown in a panel when its one-vs-rest AUC for predicting
membership in at least one cluster of that lineage, from the GP's loading,
exceeds 0.9 (**AUC > 0.9**); those are the values of
[Extended Data Table 6](ExtendedDataTable6.html), computed on the same healthy
non-thymocyte cells. Selection uses all of a lineage's clusters, including any
too small to draw. Each vertical bar is one cell, partitioned among the GPs
shown; bar height is that cell's total membership over those GPs and is not
normalized, so it exceeds 1 where a cell loads on several of them. One color is
assigned to each of the 69 GPs and held fixed across panels, so a GP marking
clusters in two lineages reads the same in both. Within a cluster, cells are
ordered by a one-dimensional t-SNE of their memberships. Clusters with fewer
than 100 cells are not drawn, larger clusters are subsampled to at most 2,000
cells, and at most 10,000 cells are drawn per panel -- bar widths therefore
reflect the cells drawn rather than cluster size. Thymocytes and DP cells are
excluded, as in [Fig. 3](Figure3.html).
:::

## Record of what the panels drew {#figs8-record}

The GP sets, the palette, and the clusters each panel drew or omitted are written
out alongside the panels, which is what
[`script/verify_structure_plot_gps.R`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/script/verify_structure_plot_gps.R)
compares against [Extended Data Table 6](ExtendedDataTable6.html) and against
the caption above:

```{r figs8-record-code, code=code_for("../script/FigureS8.R", "What each panel drew"), eval=FALSE}
```
