---
title: "Fitting CITE-seq protein programs with fixed scRNA loadings"
output: workflowr::wflow_html
---

This page documents the CITE-seq protein factorization used to connect surface
proteins to the scRNA-derived gene programs (GPs). The workflow reads the
provided Seurat object, extracts and log-normalizes its ADT assay, and uses the
cell loadings learned from the scRNA-seq EBMF fit as fixed covariates to
estimate a protein score for each GP.

## Overview

Let $Y$ be the cells-by-proteins matrix of log-normalized ADT measurements and
let $L$ be the cells-by-GPs loading matrix from the scRNA-seq fit. We estimate
the protein-by-GPs score matrix $U$ in

$$
Y \approx L U^T,
$$

while keeping $L$ fixed. An ordinary least-squares projection initializes $U$;
`flashier` then refines the protein scores under point-Laplace priors and
protein-specific residual variances.

The complete cluster-scale source scripts are:

- [`prepare_citeseq_protein_matrices_20260206.R`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/code/other/prepare_citeseq_protein_matrices_20260206.R)
- [`fit_citeseq_fixed_loading_ebmf_20260206.R`](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/code/other/fit_citeseq_fixed_loading_ebmf_20260206.R)

## Key steps

### 0. Setup

The script takes the path to a provided Seurat object and an analysis directory.
After loading the required packages, it reads the object and its cell metadata.
The object must contain an `ADT` assay and the `cellID`, `cite_seq`, and
`nCount_ADT` metadata columns.

```{r, eval=FALSE}
library(flashier)
library(Seurat)

seurat_obj <- readRDS(path_to_seurat_object)
seurat_meta <- seurat_obj@meta.data
```

### 1. ADT normalization

The ADT assay is stored with proteins in rows and cells in columns. We use
Seurat `LogNormalize` with the rounded mean ADT library size as the scale
factor, then transpose the result so that cells are rows and proteins are
columns.

```{r, eval=FALSE}
adt_scale_factor <- round(mean(seurat_meta$nCount_ADT, na.rm = TRUE))

seurat_obj <- NormalizeData(
  seurat_obj,
  assay = "ADT",
  normalization.method = "LogNormalize",
  scale.factor = adt_scale_factor
)
protein_mat_normalized_lognorm <- t(seurat_obj[["ADT"]]$data)
saveRDS(
  protein_mat_normalized_lognorm,
  file.path(data_path, "protein_mat_normalized_lognorm.rds")
)
```

For cell $i$ and protein $j$, the normalized value is

$$
Y_{ij} = \log\left(1 + \frac{C_{ij}}{\sum_q C_{iq}} \times s\right),
$$

where $C_{ij}$ is the raw ADT count and $s$ is `adt_scale_factor`.

### 2. Select measured cells and align matrices

Only cells marked `cite_seq == TRUE` are eligible. Their cell IDs are
intersected explicitly with both the normalized protein matrix and the scRNA
loading matrix, then both matrices are placed in the same order before fitting.

```{r, eval=FALSE}
cells_measured <- seurat_meta$cellID[
  !is.na(seurat_meta$cite_seq) & seurat_meta$cite_seq
]
cells_used <- cells_measured[
  cells_measured %in% rownames(protein_mat_normalized_lognorm) &
    cells_measured %in% rownames(scRNA_result$L_pm)
]

L_mat <- scRNA_result$L_pm[cells_used, , drop = FALSE]
protein_mat_normalized <- as.matrix(
  protein_mat_normalized_lognorm[cells_used, , drop = FALSE]
)
```

### 3. OLS initialization

With $L$ fixed, the least-squares initialization is

$$
\hat U^T = (L^T L)^{-1} L^T Y.
$$

The implementation uses a QR decomposition instead of forming the inverse
explicitly.

```{r, eval=FALSE}
qrL <- qr(L_mat)
U_t <- qr.coef(qrL, protein_mat_normalized)
U <- t(U_t)
saveRDS(U, file.path(data_path, "protein_projection_OLS_lognorm.rds"))
```

### 4. Fixed-loading flashier fit

The OLS protein scores initialize `flashier`. The cell loading for every GP is
then fixed, so backfitting updates the protein scores and residual variances but
does not alter the scRNA-derived cellular programs.

- `ebnm_point_laplace` allows each protein score to be exactly zero or positive
  or negative.
- `var_type = 2` estimates a separate residual variance for every protein.
- `flash_factors_fix(..., which_dim = "loadings")` fixes all columns of $L$.

```{r, eval=FALSE}
flash_fixed_loading <- flash_init(protein_mat_normalized, var_type = 2) |>
  flash_set_verbose(1) |>
  flash_factors_init(
    list(L_mat, U),
    ebnm_fn = ebnm_point_laplace
  ) |>
  flash_factors_fix(
    kset = seq_len(ncol(L_mat)),
    which_dim = "loadings"
  )
```

### 5. Backfitting schedule

The fit is checkpointed after 20, 40, 80, 120, 160, and 200 cumulative
iterations. Extrapolation is disabled for the first two 20-iteration stages
and enabled for each subsequent 40-iteration stage.

| Cumulative iteration | Additional iterations | Extrapolate |
|---:|---:|:---|
| 20 | 20 | No |
| 40 | 20 | No |
| 80 | 40 | Yes |
| 120 | 40 | Yes |
| 160 | 40 | Yes |
| 200 | 40 | Yes |

```{r, eval=FALSE}
checkpoint_schedule <- data.frame(
  cumulative_iterations = c(20L, 40L, 80L, 120L, 160L, 200L),
  additional_iterations = c(20L, 20L, 40L, 40L, 40L, 40L),
  extrapolate = c(FALSE, FALSE, TRUE, TRUE, TRUE, TRUE)
)

for (i in seq_len(nrow(checkpoint_schedule))) {
  flash_fixed_loading <- flash_backfit(
    flash_fixed_loading,
    extrapolate = checkpoint_schedule$extrapolate[[i]],
    maxiter = checkpoint_schedule$additional_iterations[[i]],
    verbose = 2
  )
  # The source script saves both the full fit and a compact summary here.
}
```

The final compact output,
`protein_flash_selected_summary_lognorm_backfit200.rds`, contains `L_pm`,
`F_pm`, `pve`, `elbo`, and `residuals_sd`. This is the protein-program matrix
used by Figure 6, Figure S6, and the protein-related Extended Data tables.
